---
title: "NFL BDB 2021 — Exploratory Data Analysis"
subtitle: "STAT 1961 Capstone"
author: "David Neumann"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: [brite, ../theme/capstone.scss]
    code-summary: "Show code"
    fig-width: 8
    fig-height: 5
    embed-resources: true
execute:
  warning: false
  message: false
---

# 0. Setup

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(arrow)
library(here)
library(nflverse)
library(gt)
library(gtExtras)
library(patchwork)
library(scales)

source(here("R", "constants.R"))
source(here("R", "utils.R"))
source(here("R", "standardize.R"))
source(here("R", "coverage.R"))
source(here("R", "sample_rules.R"))
source(here("R", "viz.R"))
```

## 0.1 What this notebook is, and is not

This notebook reports. It does not build. Every table below reads a
persisted artifact; nothing here defines the sample, and nothing here writes
to `data/`.

| Layer | Owns | Files |
|---|---|---|
| Mirror | Faithful copy of the Kaggle download | `scripts/01_raw_to_parquet.R` |
| Canonical | Naming, deduplication, `side` derivation, coordinate standardization, key types | `scripts/02_build_canonical.R` |
| Index | Measured facts about every play and player-play | `scripts/03_build_play_index.R` |
| Sample | Exclusion decisions | `scripts/04_build_sample.R`, `R/sample_rules.R` |
| Report | Diagnostics and prose | this file |

The test separating the index layer from the sample layer: a column that
answers *"what is true of this play"* belongs in `play_index`; a column that
answers *"should this play be kept"* is a decision and belongs in
`sample_flags`. `has_kinematic_defect` is the first kind. `keep_clean_window`
is the second.

Six chunks below still query the tracking dataset directly, because they need
row-level distributions rather than play-level summaries: coordinate bounds,
kinematic histograms, column missingness, the zero-offense geometry test, the
direction-convention test, and the standardization check. Everything else
reads the index.

```{r}
#| label: load_data

trk <- open_dataset(here("data", "processed", "tracking"))

plays <- read_parquet(here("data", "processed", "plays.parquet"))
players <- read_parquet(here("data", "processed", "players.parquet"))
games <- read_parquet(here("data", "processed", "games.parquet"))
targets <- read_parquet(here("data", "processed", "targeted_receiver.parquet"))
coverages <- read_parquet(here("data", "processed", "coverages_week1.parquet"))

play_index <- read_parquet(here("data", "processed", "play_index.parquet"))
player_play <- read_parquet(here(
  "data",
  "processed",
  "player_play_index.parquet"
))
play_events <- read_parquet(here("data", "processed", "play_events.parquet"))
defects <- read_parquet(here("data", "processed", "kinematic_defects.parquet"))

sample_flags <- read_parquet(here("data", "processed", "sample_flags.parquet"))
sample_funnel <- read_parquet(here(
  "data",
  "processed",
  "sample_funnel.parquet"
))
analytic_sample <- read_parquet(here(
  "data",
  "processed",
  "analytic_sample.parquet"
))
```

A convenience table used repeatedly below: play-level context joined onto the
index. Small enough to build once in memory.

```{r}
#| label: play_context

play_context <- play_index |>
  left_join(
    plays |>
      select(
        game_id,
        play_id,
        week,
        pass_result,
        offense_formation,
        type_dropback,
        personnel_d,
        down,
        quarter,
        yards_to_go,
        los_x,
        play_direction
      ),
    by = c("game_id", "play_id")
  )
```

# 1. Dataset Inventory

## 1.1 Basic Counts

Row counts come off the index rather than an aggregate over 18.3M rows. The
identity $n_{\text{rows}} = n_{\text{player rows}} + n_{\text{ball rows}}$
holds because ball rows are one per frame after deduplication, which 2.3
verifies.

```{r}
#| label: counts

tibble(
  quantity = c(
    "games",
    "plays (plays.parquet)",
    "plays (tracking)",
    "distinct tracked players",
    "player-plays",
    "player rows",
    "ball rows",
    "tracking rows"
  ),
  n = c(
    nrow(games),
    nrow(plays),
    nrow(play_index),
    n_distinct(player_play$nfl_id),
    nrow(player_play),
    sum(play_index$n_rows),
    sum(play_index$n_ball_frames),
    sum(play_index$n_rows) + sum(play_index$n_ball_frames)
  )
) |>
  mutate(n = comma(n)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

Distinct tracked players equals `nrow(players)`, so `players.parquet` contains
exactly the players who appear in tracking rather than the full league. This is
asserted in `03_build_play_index.R`; 2.7 displays the check.

```{r}
#| label: dedup-log

# 02_build_canonical.R deduplicates the raw week files and logs what it
# removed. Surfaced here so the cleaning step appears in the record.
read_parquet(here("data", "processed", "dedup_log.parquet")) |>
  mutate(rows_removed = rows_in - rows_out) |>
  filter(rows_removed > 0) |>
  gt() |>
  fmt_number(columns = -week, decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

## 1.2 Players tracked per play

Linemen are excluded from tracking on all but a handful of plays — see 2.5 for
the exceptions — but there may be varying numbers of non-linemen on the field
depending on personnel packages.

```{r}
#| label: players-per-play

play_index |>
  count(n_players, sort = TRUE) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.01)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: extreme-player-counts

extreme_plays <- play_context |>
  filter(n_players <= 11 | n_players >= 18) |>
  select(
    game_id,
    play_id,
    n_players,
    n_offense,
    n_defense,
    offense_formation,
    pass_result
  )

extreme_plays |>
  count(n_players, has_formation = !is.na(offense_formation), sort = TRUE) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

::: {.callout-note appearance="simple"}

These plays fall into three buckets:

1. **Special teams fakes or aborted plays.** No `offense_formation`. Removed by
   `keep_live_play` in 4.2.
2. **Goal line plays.** High numbers of linemen. Retained; a goal-line
   exclusion is a question-level decision, not a data-quality one.
3. **Defensive packages with few linemen.** Standard situations, retained.

The filter is tied to the named defect (`offense_formation` missing) rather
than to `n_players`, which would select on defensive scheme rather than on
data quality.

:::

## 1.3 Players tracked per play, by offense and defense

Offense should be consistent, with 6 on most plays and 5 on some (6OL
packages). Defense is more varied.

```{r}
#| label: players-per-side

play_index |>
  count(n_offense, n_defense, sort = TRUE) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.01)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

## 1.4 Play frames

```{r}
#| label: frames-per-play

ggplot(play_index, aes(x = n_frames)) +
  geom_histogram(binwidth = 5, fill = "lightsteelblue3", color = "black") +
  labs(
    title = "Frames per play",
    subtitle = "10 Hz, so 50 frames = 5 seconds",
    x = "frames",
    y = "plays"
  ) +
  theme_minimal()
```

`gapless` in the player-level index compares a player's row count against their
frame span, so it fails for a duplicated row as readily as for a missing one.
On the pre-deduplication build every non-gapless player-play turned out to be
duplication rather than a gap; 2.3 documents that. A non-gapless row here now
means a genuine mid-play dropout.

```{r}
#| label: cadence

player_play |>
  count(gapless, starts_at_one) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.01)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

## 1.5 Reconciling raggedness

The row-count identity
$n_{\text{rows}} = n_{\text{players}} \times n_{\text{frames}}$ can fail two
ways, and the per-player gap check only sees one of them: it looks *within* a
player and never checks agreement *across* players in the same play. A player
who drops out at frame 40 while everyone else runs to frame 60 is gapless,
starts at frame one, and still breaks the identity.

`ragged_cause` is computed in `03_build_play_index.R` from `n_gapped` and
`n_distinct_last`.

```{r}
#| label: raggedness-causes

play_index |>
  count(ragged_cause, sort = TRUE) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.01)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

Uneven endpoints should be the only mechanism on a deduplicated build. An
`unexplained` count above zero means the identity is failing for a reason
neither diagnostic covers; `internal gaps` above zero means either a genuine
mid-play dropout or that duplication has been reintroduced upstream, which 2.3
distinguishes.

```{r}
#| label: raggedness-severity

# min matters as much as max: a negative frames_lost means surplus rows, which
# is duplication rather than attrition. Reporting only the median, p90, and
# max hid exactly that on the pre-deduplication build.
play_index |>
  filter(ragged) |>
  summarize(
    plays = n(),
    min_frames_lost = min(frames_lost),
    median_frames_lost = median(frames_lost),
    p90_frames_lost = quantile(frames_lost, 0.9),
    max_frames_lost = max(frames_lost)
  ) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

# 2. Dataset Integrity

## 2.1 Coordinate bounds

A football field is $120 \times 53.3$ yards including end zones. Values slightly
outside these are expected as players run out of bounds, but large differences
would indicate corruption.

```{r}
#| label: coord-bounds

trk |>
  summarize(
    across(
      c(x, y),
      list(min = min, max = max, mean = mean),
      .names = "{.col}_{.fn}"
    )
  ) |>
  collect() |>
  gt() |>
  fmt_number(decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: coord-oob

trk |>
  summarize(
    beyond_x = sum(x < -10 | x > 130, na.rm = TRUE),
    beyond_y = sum(y < -10 | y > (190 / 3), na.rm = TRUE)
  ) |>
  collect() |>
  pivot_longer(everything(), names_to = "check", values_to = "rows") |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

## 2.2 Kinematics plausibility

Speed `s` is yd/s. Elite human top speed is about $10.5$ yd/s, so anything
above $13$ is not a football player. Per-frame displacement `dis` is bounded by
the same speed over a $0.1$ s frame, i.e. $1.3$ yd. `dir` and `o` should lie in
$[0, 360)$.

`MAX_SPEED`, `MAX_DIS`, and `MAX_ACCEL` now live in `R/constants.R` rather than
being defined inline here, because `03_build_play_index.R` flags defective rows
with them and a threshold that differed between the script and the notebook
would make the two disagree about which plays are clean.

```{r}
#| label: kinematics-thresholds-shown

tibble(
  threshold = c("MAX_SPEED (yd/s)", "MAX_DIS (yd/frame)", "MAX_ACCEL (yd/s^2)"),
  value = c(MAX_SPEED, MAX_DIS, MAX_ACCEL)
) |>
  gt() |>
  fmt_number(columns = value, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: kinematics-bins

bin_w <- c(s = 0.1, a = 0.1, dis = 0.01, dir = 2, o = 2)

kin_hist <- imap(bin_w, \(w, col) {
  trk |>
    filter(!is_ball) |>
    mutate(bin = floor(!!sym(col) / w) * w) |>
    count(bin) |>
    collect() |>
    mutate(name = col, width = w)
}) |>
  list_rbind() |>
  mutate(name = factor(name, levels = names(bin_w)))

kin_na <- kin_hist |>
  filter(is.na(bin)) |>
  select(name, n_missing = n)

kin_hist <- filter(kin_hist, !is.na(bin))
```

```{r}
#| label: kinematics-missing

kin_na |> gt() |> gt_theme_538(quiet = TRUE)
```

```{r}
#| label: kinematics-dist

ggplot(kin_hist, aes(x = bin, y = n, width = width)) +
  geom_col(fill = "lightsteelblue3") +
  facet_wrap(~name, scales = "free") +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(title = "Kinematic column distributions", x = NULL, y = NULL) +
  theme_minimal()
```

The tail quantiles come from the same bin counts, exact to the bin width.
Arrow's `quantile()` is a t-digest approximation, which is the wrong tool when
the target is a handful of rows in the extreme tail. Exact extrema are a
separate scalar aggregate, cheap because nothing is materialized.

```{r}
#| label: kinematics-tails

kin_quantiles <- kin_hist |>
  filter(name %in% c("s", "a", "dis")) |>
  arrange(name, bin) |>
  group_by(name) |>
  summarize(
    p50 = bin[which.max(cumsum(n) / sum(n) >= 0.50)],
    p99 = bin[which.max(cumsum(n) / sum(n) >= 0.99)],
    .groups = "drop"
  )

kin_extrema <- trk |>
  filter(!is_ball) |>
  summarize(
    across(c(s, a, dis), list(min = min, max = max), .names = "{.col}_{.fn}")
  ) |>
  collect() |>
  pivot_longer(
    everything(),
    names_to = c("name", ".value"),
    names_pattern = "(.*)_(min|max)$"
  )

kin_quantiles |>
  left_join(kin_extrema, by = "name") |>
  relocate(min, .after = name) |>
  gt() |>
  fmt_number(columns = -name, decimals = 4) |>
  gt_theme_538(quiet = TRUE)
```

### Implausible rows

`kinematic_defects.parquet` is one row per defective tracking row, with the
violated bound (`which`) and the row's position relative to the measurement
window (`position_in_play`) already attached by
`03_build_play_index.R`.

```{r}
#| label: kinematics-defects

tibble(
  quantity = c(
    "defective rows",
    "plays affected",
    "players affected",
    "games affected"
  ),
  n = c(
    nrow(defects),
    n_distinct(defects$game_id, defects$play_id),
    n_distinct(defects$nfl_id),
    n_distinct(defects$game_id)
  )
) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: kinematics-defects-detail

defects |>
  count(which, week) |>
  pivot_wider(names_from = which, values_from = n, values_fill = 0) |>
  arrange(week) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: defects-by-week

# Speed violations concentrate in a small number of weeks. That clustering is a
# collection artifact rather than noise.
defects |>
  group_by(week) |>
  summarize(
    rows = n(),
    plays = n_distinct(game_id, play_id),
    games = n_distinct(game_id),
    max_s = max(s),
    max_dis = max(dis),
    .groups = "drop"
  ) |>
  slice_max(rows, n = 5) |>
  gt() |>
  fmt_number(columns = c(max_s, max_dis), decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

::: {.callout-note appearance="simple"}

A whole-play exclusion rule discards a 60-frame play for a single bad frame,
which is heavy-handed if the defect sits outside the interval actually being
measured. Section 4.1 scopes the rule against the snap-to-throw window.

:::

## 2.3 Key uniqueness and ball coverage

Every downstream distance calculation assumes one row per player per frame. A
silent duplicate would not error; it would quietly double-weight a player in
any per-frame aggregate.

The raw week files contain exactly-duplicated player rows on a small number of
plays. `02_build_canonical.R` removes them with a full-row `distinct()` and
asserts on the key afterwards, and `03_build_play_index.R` re-checks the key
across the whole processed dataset. This section is a tripwire rather than a
discovery: a nonzero count means the deduplication regressed.

```{r}
#| label: key-uniqueness

tibble(
  check = c(
    "plays with duplicate (play, player, frame) keys",
    "plays with internal frame gaps",
    "player-plays not gapless"
  ),
  n = c(
    sum(play_index$has_duplicate_rows),
    sum(play_index$n_gapped > 0),
    sum(!player_play$gapless)
  )
) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

Ball rows carry `nfl_id = NA`, so the uniqueness check also covers "exactly one
ball row per frame". What it does not cover is plays where the ball is missing
entirely, or present for only part of the play.

```{r}
#| label: ball-coverage

play_index |>
  count(ball_status, sort = TRUE) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.01)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

`n_ball_frames` counts distinct frames, so duplicate ball rows would not
inflate it. "More ball frames than play frames" is therefore a genuine span
difference: the ball is tracked on frames where no player is.

## 2.4 Missingness

Split by `is_ball`, because ball rows legitimately have no `position`, `route`,
or `jersey_number`. Pooling the two makes a structural blank look like a data
defect and hides real defects behind it.

```{r}
#| label: missingness

na_cols <- c(
  "nfl_id",
  "display_name",
  "position",
  "jersey_number",
  "team_abbr",
  "side",
  "x",
  "y",
  "s",
  "a",
  "dis",
  "o",
  "dir",
  "event",
  "route",
  "time",
  "play_direction"
)

trk |>
  group_by(is_ball) |>
  summarize(
    across(all_of(na_cols), ~ sum(is.na(.x))),
    n_rows = n(),
    .groups = "drop"
  ) |>
  collect() |>
  pivot_longer(all_of(na_cols), names_to = "column", values_to = "n_missing") |>
  mutate(
    pct_missing = n_missing / n_rows,
    row_type = if_else(is_ball, "ball", "player")
  ) |>
  select(column, row_type, n_missing, pct_missing) |>
  pivot_wider(names_from = row_type, values_from = c(n_missing, pct_missing)) |>
  arrange(desc(pct_missing_player)) |>
  gt() |>
  fmt_percent(columns = starts_with("pct_"), decimals = 2) |>
  fmt_number(columns = starts_with("n_missing"), decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

`team_abbr` has no missing values among player rows, and `side` has none either
now that `02_build_canonical.R` asserts on it. Both matter for 2.5. The nonzero
`dir` and `o` counts matter for 2.6, where an unguarded `median()` over them
returns `NA`.

The same audit on the play table, which is small enough to do in one pass.

```{r}
#| label: missingness-plays

plays |>
  summarize(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  mutate(pct_missing = n_missing / nrow(plays)) |>
  arrange(desc(n_missing)) |>
  gt() |>
  fmt_percent(columns = pct_missing, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

`absolute_yardline_number`, `los_x`, `game_clock`, `type_dropback`, and both
pre-snap score columns are missing on an identical count of plays. That is one
coherent block of play-level metadata rather than several independent gaps; 2.6
tests whether the block is outcome-neutral.

## 2.5 Side assignment

Section 1.3 found plays with zero offensive players tracked. The first
hypothesis was a fallthrough in how `side` was derived:

```r
side = case_when(
  is_ball ~ "ball",
  team_abbr == possession_team ~ "offense",
  TRUE ~ "defense"
)
```

If either `team_abbr` or `possession_team` were `NA`, the middle condition would
evaluate to `NA`, fall through, and label the row `"defense"` — everything
unresolvable silently becoming a defender. Three checks reject that hypothesis:
`team_abbr` has no missing values (2.4), the abbreviation vocabularies agree
exactly, and every suspect play's `possession_team` matches one of its own two
teams. The fallthrough cannot fire, so the cause is elsewhere. The derivation in
`02_build_canonical.R` now carries an explicit `NA` branch and an assertion
anyway, which converts a silent failure mode into a loud one for future
rebuilds.

```{r}
#| label: suspect-sides

suspect_sides <- play_index |>
  filter(n_offense == 0 | n_offense > 6 | n_defense == 0) |>
  left_join(
    plays |>
      select(
        game_id,
        play_id,
        possession_team,
        home_team,
        away_team,
        offense_formation,
        play_description
      ),
    by = c("game_id", "play_id")
  )

suspect_sides |>
  count(n_offense, n_defense, sort = TRUE) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: team-vocabulary

# The abbreviation vocabularies must agree exactly or the == comparison fails
# for every play involving the mismatched franchise.
poss_teams <- sort(unique(plays$possession_team))
game_teams_vocab <- sort(unique(c(plays$home_team, plays$away_team)))

tibble(
  direction = c(
    "in possession_team, not in games",
    "in games, not in possession_team"
  ),
  values = c(
    paste0(setdiff(poss_teams, game_teams_vocab), collapse = ", "),
    paste0(setdiff(game_teams_vocab, poss_teams), collapse = ", ")
  )
) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: side-mismatch-detail

# A play whose possession_team matches neither of its own two teams would be
# unassignable regardless of vocabulary overlap.
suspect_sides |>
  mutate(
    poss_matches_a_team = possession_team == home_team |
      possession_team == away_team
  ) |>
  count(n_offense == 0, poss_matches_a_team) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

### What the zero-offense plays actually are

With the fallthrough ruled out, the discriminating test is geometric. Offense
lines up behind the line of scrimmage — 2.6 puts the median offensive
$x_{\text{rel}}$ at about $-1.9$ yd — while defenders line up in front of it. A
tracked group sitting behind the LOS and containing a quarterback would be the
offense mislabeled, meaning `possessionTeam` is wrong for that play. A group
sitting downfield with no quarterback is genuinely the defense, with the offense
simply unrecorded.

```{r}
#| label: zero-offense-geometry

zero_offense_keys <- play_index |>
  filter(n_offense == 0) |>
  select(game_id, play_id)

zero_offense_geom <- trk |>
  filter(!is_ball, event == "ball_snap") |>
  select(game_id, play_id, nfl_id, position, team_abbr, x) |>
  collect() |>
  cast_keys() |>
  inner_join(zero_offense_keys, by = c("game_id", "play_id")) |>
  left_join(
    plays |>
      select(game_id, play_id, los_x, possession_team, home_team, away_team),
    by = c("game_id", "play_id")
  ) |>
  mutate(x_rel = x - los_x) |>
  group_by(game_id, play_id, possession_team, home_team, away_team) |>
  summarize(
    n_tracked = n(),
    tracked_team = paste(sort(unique(team_abbr)), collapse = "/"),
    median_x_rel = median(x_rel, na.rm = TRUE),
    has_qb = any(position == "QB"),
    positions = paste(sort(unique(position)), collapse = " "),
    .groups = "drop"
  )

zero_offense_geom |>
  select(
    game_id,
    play_id,
    n_tracked,
    tracked_team,
    possession_team,
    median_x_rel,
    has_qb,
    positions
  ) |>
  gt() |>
  fmt_number(columns = median_x_rel, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: zero-offense-verdict

zero_offense_geom |>
  mutate(
    reading = case_when(
      has_qb & median_x_rel < 0 ~ "offense mislabeled (possessionTeam wrong)",
      !has_qb & median_x_rel > 0 ~ "offense genuinely untracked",
      .default = "ambiguous - inspect individually"
    )
  ) |>
  count(reading, sort = TRUE) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

Every one of these plays reads as a genuine tracking gap: no quarterback, and
players positioned downfield of the line of scrimmage. `possessionTeam` is not
mislabeled anywhere in the dataset and the `side` derivation is working. These
plays exit the sample at `keep_sides` in 4.2.

Plays with a full complement of offensive players are the candidate mirror image
of the same defect.

```{r}
#| label: full-offense-plays

play_index |>
  filter(n_offense >= 10) |>
  left_join(
    player_play |>
      filter(side == "offense") |>
      group_by(game_id, play_id) |>
      summarize(
        positions = paste(sort(unique(position)), collapse = " "),
        .groups = "drop"
      ),
    by = c("game_id", "play_id")
  ) |>
  select(game_id, play_id, n_offense, n_defense, positions) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

Section 1.2 qualifies the no-linemen claim. Quantify the exception rather than
leaving it as prose.

```{r}
#| label: lineman-exceptions

player_play |>
  filter(side == "offense") |>
  count(position, sort = TRUE) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.01)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

Defensive-line and secondary positions appearing on the offensive side are
partly legitimate — teams do line defensive tackles up as goal-line blockers —
and partly the full-offense plays above. The two tables together say which.

## 2.6 Direction convention and standardization

`R/viz.R` asserts that `dir` is measured in degrees clockwise from $+y$, so that

$$v_x = s \sin(\text{dir} \cdot \pi/180), \qquad v_y = s \cos(\text{dir} \cdot \pi/180)$$

That assertion lives in a code comment. Testing it against observed displacement
makes it a result, and the same test catches corrupt frames.

Frame-to-frame differencing needs `lag()`, which Arrow will not push down, so
this runs on a random sample of plays rather than all 18.3M rows.

```{r}
#| label: convention-sample

set.seed(1961)

conv_keys <- play_index |>
  slice_sample(n = 300) |>
  select(game_id, play_id)

conv <- trk |>
  filter(!is_ball) |>
  select(game_id, play_id, nfl_id, frame_id, x, y, s, dis, dir) |>
  collect() |>
  cast_keys() |>
  inner_join(conv_keys, by = c("game_id", "play_id")) |>
  arrange(game_id, play_id, nfl_id, frame_id) |>
  group_by(game_id, play_id, nfl_id) |>
  mutate(
    contiguous = frame_id == lag(frame_id) + 1,
    dx = x - lag(x),
    dy = y - lag(y)
  ) |>
  ungroup() |>
  filter(contiguous) |>
  mutate(
    displacement = sqrt(dx^2 + dy^2),
    rad = dir * pi / 180,
    step = s / TRACKING_HZ,
    # A: dir clockwise from +y  (the convention viz.R assumes)
    err_a = sqrt((step * sin(rad) - dx)^2 + (step * cos(rad) - dy)^2),
    # B: dir counterclockwise from +x  (the plausible alternative)
    err_b = sqrt((step * cos(rad) - dx)^2 + (step * sin(rad) - dy)^2)
  )
```

```{r}
#| label: convention-verdict

# na.rm is required: dir carries missing values (2.4), and without it both
# medians return NA and the test produces no verdict.
conv |>
  summarize(
    n_frame_pairs = sum(!is.na(err_a)),
    n_dropped_na_dir = sum(is.na(err_a)),
    `median |displacement - dis|` = median(
      abs(displacement - dis),
      na.rm = TRUE
    ),
    `cor(displacement, s/10)` = cor(displacement, step, use = "complete.obs"),
    `median err, dir from +y` = median(err_a, na.rm = TRUE),
    `median err, dir from +x` = median(err_b, na.rm = TRUE)
  ) |>
  pivot_longer(everything(), names_to = "statistic") |>
  gt() |>
  fmt_number(columns = value, decimals = 4) |>
  gt_theme_538(quiet = TRUE)
```

The separation is decisive rather than marginal. Measuring `dir` from $+y$ lands
at the rounding scale of the source data, while measuring from $+x$ lands at the
step length itself ($\approx 0.28$ yd at the median), because the wrong
convention reflects the predicted step about the $45°$ line. `viz.R` is right,
and the $|\text{displacement} - \text{dis}|$ figure sitting at $\approx 0.01$ yd
rules out a coordinate transform bug at the same time.

### Missing line of scrimmage

Standardization is the other assertion worth proving, but `los_x` is needed to
express position relative to the line of scrimmage and 2.4 showed it missing on
a block of plays. Characterize that block first.

```{r}
#| label: los-missingness

plays |>
  mutate(
    na_los = is.na(los_x),
    na_yardline = is.na(absolute_yardline_number)
  ) |>
  count(na_los, na_yardline, play_direction) |>
  arrange(desc(n)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: los-gap-context

# If los_x is missing, is the play otherwise well-formed?
plays |>
  filter(is.na(los_x)) |>
  count(has_formation = !is.na(offense_formation), pass_result, quarter) |>
  arrange(desc(n)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: los-missing-bias

plays |>
  mutate(los_missing = if_else(is.na(los_x), "los missing", "los present")) |>
  count(los_missing, pass_result) |>
  group_by(los_missing) |>
  mutate(share = n / sum(n)) |>
  ungroup() |>
  select(pass_result, los_missing, share) |>
  pivot_wider(names_from = los_missing, values_from = share, values_fill = 0) |>
  mutate(difference = `los missing` - `los present`) |>
  arrange(desc(abs(difference))) |>
  gt() |>
  fmt_percent(columns = -pass_result, decimals = 1) |>
  gt_theme_538(quiet = TRUE)
```

::: {.callout-warning appearance="simple"}

Recorded in `notes/decisions.md`: the missing-LOS block runs roughly 72%
incomplete against a 31% baseline, so excluding it is not missing-at-random with
respect to outcome. Any outcome model fit on the retained sample carries that
selection, and the Methods section states its direction and magnitude.

:::

### Rotation, not reflection

`standardize_direction()` rotates left-moving plays by $180°$ by reflecting
*both* $x$ and $y$. Flipping only $x$ would mirror the field, swapping offensive
left and right while producing output that looks entirely plausible. The
signature of that bug is a $y$ distribution that splits by `play_direction`
while $x$ looks fine.

```{r}
#| label: snap-positions

snap_offense <- trk |>
  filter(!is_ball, side == "offense", event == "ball_snap") |>
  select(game_id, play_id, nfl_id, x, y) |>
  collect() |>
  cast_keys() |>
  left_join(
    plays |> select(game_id, play_id, los_x, play_direction),
    by = c("game_id", "play_id")
  ) |>
  mutate(x_rel = x - los_x)
```

```{r}
#| label: standardization-check

snap_offense |>
  summarize(
    player_plays = n(),
    missing_x = sum(is.na(x)),
    missing_y = sum(is.na(y)),
    missing_los = sum(is.na(los_x)),
    usable = sum(!is.na(x_rel) & !is.na(y)),
    pct_usable = usable / player_plays
  ) |>
  gt() |>
  fmt_percent(columns = pct_usable, decimals = 2) |>
  fmt_number(columns = -pct_usable, decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: standardization-quantiles

std_long <- snap_offense |>
  select(play_direction, x_rel, y) |>
  pivot_longer(c(x_rel, y), names_to = "axis") |>
  filter(!is.na(value), !is.na(play_direction))

std_long |>
  group_by(axis, play_direction) |>
  summarize(
    n = n(),
    p10 = quantile(value, 0.10),
    p50 = quantile(value, 0.50),
    p90 = quantile(value, 0.90),
    .groups = "drop"
  ) |>
  gt() |>
  fmt_number(columns = starts_with("p"), decimals = 2) |>
  fmt_number(columns = n, decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: standardization-plot
#| fig-height: 3.5

std_long |>
  ggplot(aes(value, colour = play_direction)) +
  stat_ecdf(linewidth = 0.7) +
  facet_wrap(~axis, scales = "free_x") +
  labs(
    title = "Offensive positions at snap, by original play direction",
    subtitle = "Correct rotation: the two curves coincide on both axes",
    x = "yards",
    y = "ECDF",
    colour = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The curves coincide on both axes, so the transform is a rotation. Compare the
quantiles, not a test statistic: at $n \approx 10^5$ per group a two-sample KS
test rejects on differences far too small to matter, so the $p$-value carries no
information here.

## 2.7 Join integrity

Arrow stores integer keys as `int64`, while a locally constructed frame may hold
`integer` or `double`. A type mismatch on the join key produces zero matches
rather than an error. `cast_keys()` in `R/utils.R` now casts every identifier to
`int32` at the point of collection, so the whole processed layer shares one key
type and the mismatch cannot arise.

These five checks are asserted in `03_build_play_index.R`, which warns on
failure. Re-displayed here because a build-time assertion nobody sees is not a
finding.

```{r}
#| label: join-integrity

tibble(
  check = c(
    "tracking nfl_id absent from players",
    "players never appearing in tracking",
    "tracking plays absent from plays table",
    "plays table entries absent from tracking",
    "plays with zero tracked players",
    "duplicated play keys in play_index"
  ),
  n = c(
    nrow(anti_join(
      distinct(player_play, nfl_id),
      distinct(players, nfl_id),
      by = "nfl_id"
    )),
    nrow(anti_join(
      distinct(players, nfl_id),
      distinct(player_play, nfl_id),
      by = "nfl_id"
    )),
    nrow(anti_join(
      distinct(play_index, game_id, play_id),
      distinct(plays, game_id, play_id),
      by = c("game_id", "play_id")
    )),
    nrow(anti_join(
      distinct(plays, game_id, play_id),
      distinct(play_index, game_id, play_id),
      by = c("game_id", "play_id")
    )),
    sum(play_index$n_players == 0),
    sum(duplicated(play_index[c("game_id", "play_id")]))
  )
) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

All six checks return zero. `players.parquet` contains exactly the players who
appear in tracking, so the player join is complete in both directions, and the
play and game keys agree across all three tables.

# 3. Trajectory

Everything in the coverage, route, separation, and decision-timing space is
anchored to events. Nothing in sections 1 and 2 has touched them.

## 3.1 Event vocabulary

The `event` column is populated on every row of the frame in which the event
occurs, so `play_events.parquet` is deduplicated to one row per
(play, frame, event).

```{r}
#| label: event-vocabulary

play_events |>
  distinct(game_id, play_id, event) |>
  count(event, name = "plays") |>
  left_join(count(play_events, event, name = "frames"), by = "event") |>
  mutate(pct_of_plays = percent(plays / nrow(play_index), accuracy = 0.1)) |>
  arrange(desc(plays)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

## 3.2 Event presence per play

`play_index` carries `f_<event>` for the first frame of each anchor event, and
`n_ev_<event>` for the multiplicity of the three anchors where multiplicity
matters.

```{r}
#| label: event-frames

play_index |>
  summarize(across(starts_with("f_"), ~ sum(!is.na(.x)))) |>
  pivot_longer(everything(), names_to = "event", values_to = "plays") |>
  mutate(
    event = str_remove(event, "^f_"),
    pct = percent(plays / nrow(play_index), accuracy = 0.1)
  ) |>
  arrange(desc(plays)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

A play with no `ball_snap` has no time origin. A play with no `pass_forward` is
a sack, a scramble, or a penalty-aborted snap: a population to define
explicitly, not to leave implicit.

```{r}
#| label: missing-throw-reason

play_context |>
  mutate(
    has_snap = !is.na(f_ball_snap),
    has_throw = !is.na(f_pass_forward)
  ) |>
  count(has_snap, has_throw, pass_result) |>
  arrange(desc(n)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

### What the no-throw plays are

`keep_any_throw` is the largest single exclusion in the funnel, so its
composition determines whether it is attrition or a population definition.

```{r}
#| label: no-throw-composition

sample_flags |>
  filter(apply_flags(sample_flags, c("keep_live_play", "keep_snap"))) |>
  filter(!keep_any_throw) |>
  count(pass_result, sort = TRUE) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

::: {.callout-important appearance="simple"}

The no-throw block is overwhelmingly sacks (`pass_result == "S"`), consistent
with league-wide 2018 sack totals. `keep_any_throw` is therefore a *population
definition* — plays where a forward pass was released — rather than a
data-quality exclusion, and it is grouped as such in 4.2. The handful of
non-sack residuals are recorded pass attempts with no throw event of any kind:
either penalty-nullified or event-labeling gaps. Recorded in
`notes/decisions.md`.

:::

The mirror-image check: a sack that *does* carry a throw event would survive
into the sample as a pass attempt.

```{r}
#| label: sacks-surviving

analytic_sample |>
  count(pass_result, throw_anchor) |>
  pivot_wider(
    names_from = throw_anchor,
    values_from = n,
    values_fill = 0
  ) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

### Non-standard throw anchors

Some completions and incompletions have no `pass_forward` event at all. The
forward pass on those plays is recorded under a different label, so a
`pass_forward`-only anchor discards real passes as though they were sacks.

```{r}
#| label: alternate-throw-anchor

play_context |>
  filter(is.na(f_pass_forward), pass_result %in% c("C", "I", "IN")) |>
  mutate(recoverable = !is.na(f_pass_shovel) | !is.na(f_pass_lateral)) |>
  count(pass_result, recoverable) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

`keep_any_throw` in 4.2 accepts `pass_shovel` and recovers those plays.
`f_throw` in `analytic_sample.parquet` resolves the precedence once, in the same
place the rule accepted it, so no downstream file re-derives it.

## 3.3 Decision timing

$$t_{\text{throw}} = \frac{f_{\text{throw}} - f_{\text{ball\_snap}}}{10}$$

```{r}
#| label: duplicate-anchors

# min() silently takes the first of several. Where a play has more than one
# forward pass, "first" may be the wrong one.
play_context |>
  select(game_id, play_id, pass_result, starts_with("n_ev_")) |>
  pivot_longer(
    starts_with("n_ev_"),
    names_to = "event",
    values_to = "occurrences"
  ) |>
  filter(occurrences > 1) |>
  mutate(event = str_remove(event, "^n_ev_")) |>
  count(event, pass_result) |>
  pivot_wider(names_from = pass_result, values_from = n, values_fill = 0) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: timing

timing <- play_context |>
  mutate(f_throw = coalesce(f_pass_forward, f_pass_shovel)) |>
  filter(!is.na(f_ball_snap), !is.na(f_throw)) |>
  mutate(
    t_throw = (f_throw - f_ball_snap) / TRACKING_HZ,
    t_arrive = (f_pass_arrived - f_ball_snap) / TRACKING_HZ,
    t_flight = (f_pass_arrived - f_throw) / TRACKING_HZ,
    frames_pre_snap = f_ball_snap - first_frame,
    frames_post_throw = last_frame - f_throw
  )
```

```{r}
#| label: time-to-throw

timing |>
  summarize(
    across(
      c(t_throw, t_arrive, t_flight),
      list(
        n = ~ sum(!is.na(.x)),
        p05 = ~ quantile(.x, 0.05, na.rm = TRUE),
        p50 = ~ quantile(.x, 0.50, na.rm = TRUE),
        p95 = ~ quantile(.x, 0.95, na.rm = TRUE),
        min = ~ min(.x, na.rm = TRUE),
        max = ~ max(.x, na.rm = TRUE)
      )
    )
  ) |>
  pivot_longer(
    everything(),
    names_to = c("quantity", ".value"),
    names_pattern = "^(t_[a-z]+)_(.*)$"
  ) |>
  gt() |>
  fmt_number(columns = -c(quantity, n), decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: time-to-throw-dist
#| fig-height: 3.5

timing |>
  filter(between(t_throw, 0, 10)) |>
  ggplot(aes(t_throw)) +
  geom_histogram(binwidth = 0.1, fill = "lightsteelblue3") +
  labs(
    title = "Time from snap to forward pass",
    subtitle = paste0(
      "n = ",
      comma(sum(between(timing$t_throw, 0, 10), na.rm = TRUE)),
      " plays; truncated at 10 s"
    ),
    x = "seconds",
    y = "plays"
  ) +
  theme_minimal()
```

Negative or zero `t_throw` values are impossible and indicate event-ordering
problems rather than fast decisions.

```{r}
#| label: timing-anomalies

timing |>
  summarize(
    `t_throw <= 0` = sum(t_throw <= 0, na.rm = TRUE),
    `t_throw > 10 s` = sum(t_throw > 10, na.rm = TRUE),
    `t_flight < 0` = sum(t_flight < 0, na.rm = TRUE),
    `arrival before throw` = sum(f_pass_arrived < f_throw, na.rm = TRUE),
    `no pre-snap frames` = sum(frames_pre_snap <= 0, na.rm = TRUE)
  ) |>
  pivot_longer(everything(), names_to = "anomaly", values_to = "plays") |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

## 3.4 Observation window

How much usable tracking sits on either side of the throw. This sets the
feasible lag and lead for any time-indexed feature.

```{r}
#| label: observation-window

timing |>
  select(frames_pre_snap, frames_post_throw) |>
  pivot_longer(everything(), names_to = "window") |>
  group_by(window) |>
  summarize(
    p05 = quantile(value, 0.05, na.rm = TRUE),
    p50 = quantile(value, 0.50, na.rm = TRUE),
    p95 = quantile(value, 0.95, na.rm = TRUE),
    `seconds at p05` = p05 / TRACKING_HZ,
    .groups = "drop"
  ) |>
  gt() |>
  fmt_number(columns = -window, decimals = 1) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: pre-snap-constant

# p05, p50, and p95 coincide, so the pre-snap window is a structural constant
# rather than a distribution. Confirm directly.
timing |>
  count(frames_pre_snap, sort = TRUE) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) |>
  slice_head(n = 6) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

::: {.callout-important appearance="simple"}

The pre-snap window is fixed at 10 frames on over 99% of plays — exactly $1.0$
s. Every pre-snap feature, whether alignment, leverage, or motion detection, has
that and no more, and no amount of filtering buys extra. Recorded in
`notes/decisions.md`.

:::

## 3.5 Targeted receiver

```{r}
#| label: targets

play_context |>
  mutate(has_target = !is.na(target_nfl_id)) |>
  count(pass_result, has_target) |>
  pivot_wider(
    names_from = has_target,
    values_from = n,
    values_fill = 0,
    names_prefix = "target_"
  ) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

A target is only usable if that player is actually tracked on the play, which
`03_build_play_index.R` resolves as `target_tracked`.

```{r}
#| label: target-tracked

play_index |>
  filter(!is.na(target_nfl_id)) |>
  count(target_tracked) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

A non-missing `target_nfl_id` is therefore not sufficient. `keep_target` in 4.3
requires the target to be both named and tracked.

## 3.6 Route coverage

`route` is populated only for eligible receivers, so the NA rate is structural
for some positions and a defect for others. It is `NA` for every defender, so
`n_routes` in the index is an offense-only count without needing a side filter.

```{r}
#| label: route-coverage

player_play |>
  filter(side == "offense") |>
  group_by(position) |>
  summarize(
    player_plays = n(),
    with_route = sum(!is.na(route)),
    pct_with_route = with_route / player_plays,
    .groups = "drop"
  ) |>
  arrange(desc(player_plays)) |>
  gt() |>
  fmt_percent(columns = pct_with_route, decimals = 1) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: route-vocabulary

player_play |>
  filter(!is.na(route)) |>
  count(route, sort = TRUE) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: routes-per-play

play_index |>
  count(n_routes) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: zero-route-plays

# Sacks explain some zero-route plays. Anything left over is a gap in the route
# labels on plays where a pass was actually thrown.
zero_route <- play_context |>
  filter(n_routes == 0) |>
  mutate(f_throw = coalesce(f_pass_forward, f_pass_shovel))

zero_route |>
  count(pass_result, has_throw = !is.na(f_throw)) |>
  arrange(desc(n)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

The thrown-pass subset of these is enriched in incompletions at roughly the same
rate as the missing-LOS block in 2.6. Two degraded blocks of similar size and
similar outcome skew are more likely one block than two, which is testable.

```{r}
#| label: route-los-overlap

los_missing_keys <- plays |>
  filter(is.na(los_x)) |>
  select(game_id, play_id)

zero_route_thrown <- zero_route |>
  filter(!is.na(f_throw)) |>
  select(game_id, play_id)

tibble(
  quantity = c(
    "plays missing los_x",
    "plays with a throw and no route labels",
    "in both sets",
    "zero-route plays surviving keep_los"
  ),
  n = c(
    nrow(los_missing_keys),
    nrow(zero_route_thrown),
    nrow(inner_join(
      los_missing_keys,
      zero_route_thrown,
      by = c("game_id", "play_id")
    )),
    nrow(anti_join(
      zero_route_thrown,
      los_missing_keys,
      by = c("game_id", "play_id")
    ))
  )
) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

The last row is what matters. Near zero means `keep_los` already removes the
route gap and no separate exclusion is needed. A large value means a second
non-ignorable gap that any route-conditioned framing has to handle on its own.

## 3.7 Coverage label availability

This is the binding constraint on any coverage-framed question.

`coverages_week1.parquet` carries the original eight labels verbatim. Collapsing
them to a man/zone binary is a modeling decision, so it lives in
`classify_coverage()` in `R/coverage.R` and is applied at the sample layer
rather than baked into the data.

```{r}
#| label: coverage-availability

coverage_join <- plays |>
  select(game_id, play_id, week, pass_result, offense_formation, personnel_d) |>
  left_join(
    coverages |> select(game_id, play_id, coverage),
    by = c("game_id", "play_id")
  ) |>
  mutate(coverage_class = classify_coverage(coverage))

tibble(
  quantity = c(
    "plays in dataset",
    "plays with a coverage label",
    "share of dataset labeled",
    "weeks represented among labeled plays"
  ),
  value = c(
    comma(nrow(coverage_join)),
    comma(sum(!is.na(coverage_join$coverage))),
    percent(mean(!is.na(coverage_join$coverage)), accuracy = 0.1),
    as.character(n_distinct(coverage_join$week[!is.na(coverage_join$coverage)]))
  )
) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: coverage-class-mapping

coverage_join |>
  filter(!is.na(coverage)) |>
  count(coverage_class, coverage) |>
  arrange(coverage_class, desc(n)) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

Eight labels with a minimum cell of one play cannot support an eight-class
model, so `classify_coverage()` collapses them to the man/zone binary.
`Prevent Zone` is the one label mapped to `NA` rather than to a class: a single
garbage-time play, and a description of game state rather than of scheme. It
drops out of any coverage-conditional sample. The precedence is explicit —
`Man` is tested before `Zone` — so a future label containing both words
classifies as man rather than depending on `case_when()` ordering.

```{r}
#| label: coverage-by-outcome

coverage_join |>
  filter(!is.na(coverage_class)) |>
  count(coverage_class, pass_result) |>
  pivot_wider(names_from = pass_result, values_from = n, values_fill = 0) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

# 4. Sample Definition

The rule lives in `R/sample_rules.R` and is applied by
`scripts/04_build_sample.R`. This section reports it.

## 4.1 Scoping the kinematic exclusion

A whole-play rule discards a play for a defect anywhere in it. Only the
snap-to-throw interval feeds a separation or timing measurement, so
`03_build_play_index.R` locates each defective row relative to that window.

```{r}
#| label: defect-window

defects |>
  count(position_in_play, sort = TRUE) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: defect-window-plays

play_index |>
  filter(has_kinematic_defect) |>
  count(defect_in_window) |>
  mutate(
    interpretation = if_else(
      defect_in_window,
      "must exclude: defect inside the measurement window",
      "recoverable: defect outside the measurement window"
    )
  ) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

::: {.callout-important appearance="simple"}

**Decision: the scoped rule.** A play is excluded only when a defective row
falls between the snap and the throw. The majority of defective rows land after
the throw, so a whole-play rule would discard hundreds of plays whose
measurement window is clean.

The cost is that retained plays may contain uncleaned frames outside the window,
which the Methods section must state. `has_kinematic_defect` and
`defect_in_window` travel with the persisted sample so any model can test
sensitivity to this choice without rebuilding the funnel. The conservative rule
is reported alongside in 4.2 as the comparison.

:::

## 4.2 Attrition funnel

Each filter is tied to a *named defect* rather than to `n_players`, which would
select on defensive scheme rather than on data quality.

The flags are split into two blocks, and the ordering matters for
interpretation. **Population** filters define which plays are in scope at all:
live plays where a quarterback released a forward pass. **Quality** filters
identify plays that are in scope but cannot be measured. Conjunction is
order-independent, so the surviving count is the same under any ordering; what
running population first buys is that each quality step's `dropped` reads as
"plays I wanted but cannot use" rather than being inflated by plays that were
never in scope.

```{r}
#| label: funnel-scoped

sample_funnel |>
  filter(rule == "scoped") |>
  select(group, step, dropped, remaining, pct_of_start) |>
  gt(groupname_col = "group") |>
  fmt_percent(columns = pct_of_start, decimals = 1) |>
  fmt_number(columns = c(dropped, remaining), decimals = 0) |>
  sub_missing(missing_text = "—") |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: funnel-marginals

# The funnel's `dropped` is conditional on the preceding steps. These are the
# unconditional counts. The missing-LOS block is larger than the number of
# plays the funnel attributes to keep_los, because some of those plays were
# already out on population grounds.
flag_marginals(sample_flags) |>
  arrange(group, desc(n_failing)) |>
  select(group, criterion, n_failing, pct_failing) |>
  gt(groupname_col = "group") |>
  fmt_percent(columns = pct_failing, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: funnel-rule-comparison

# The two rules differ at exactly two steps: the throw anchor and the kinematic
# scope. Everything else is shared, so the comparison is the final count plus
# those two rows.
sample_funnel |>
  filter(flag %in% c(
    "keep_any_throw",
    "keep_throw",
    "keep_clean_window",
    "keep_clean_kin"
  )) |>
  select(rule, step, dropped, remaining) |>
  bind_rows(
    sample_funnel |>
      group_by(rule) |>
      slice_max(step_index, n = 1) |>
      ungroup() |>
      transmute(rule, step = "FINAL", dropped = NA_integer_, remaining)
  ) |>
  gt(groupname_col = "rule") |>
  fmt_number(columns = c(dropped, remaining), decimals = 0) |>
  sub_missing(missing_text = "—") |>
  gt_theme_538(quiet = TRUE)
```

`keep_no_dupes` drops zero plays, since `02_build_canonical.R` deduplicates
upstream. The row stays in the funnel as the tripwire described in 2.3: a
nonzero count means the cleaning step regressed.

## 4.3 Sample size under each candidate question

Same base sample, three different question framings. The point of this table is
to make the cost of each framing explicit before committing to one.

```{r}
#| label: funnel-candidates

candidates <- list(
  `A. Decision timing (all weeks)` = SAMPLE_RULE,
  `B. Separation at throw, targeted receiver` = c(SAMPLE_RULE, "keep_target"),
  `C. Coverage-conditional separation` = c(
    SAMPLE_RULE,
    "keep_target",
    "keep_coverage"
  )
)

imap(candidates, \(flags, nm) {
  keep <- apply_flags(sample_flags, flags)
  tibble(
    question = nm,
    plays = sum(keep),
    pct_of_start = mean(keep),
    weeks = n_distinct(sample_flags$week[keep])
  )
}) |>
  list_rbind() |>
  gt() |>
  fmt_percent(columns = pct_of_start, decimals = 1) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: candidate-c-cells

# For question C, the per-class cell counts are what actually limit it.
keep_c <- apply_flags(sample_flags, candidates[[3]])

sample_flags |>
  filter(keep_c) |>
  count(coverage_class, coverage) |>
  arrange(coverage_class, desc(n)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: candidate-c-binary

sample_flags |>
  filter(keep_c) |>
  count(coverage_class, pass_result) |>
  pivot_wider(names_from = pass_result, values_from = n, values_fill = 0) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

::: {.callout-important appearance="simple"}

The open decision: coarse man/zone coverage on the single labeled week, or drop
coverage as a covariate and use the full-season sample. The three rows of 4.3
are the whole argument.

The completion-rate difference between man and zone above is descriptive only.
It is confounded by down, distance, field position, and personnel, and a model
would have to condition on all of them.

:::

## 4.4 The persisted sample

`scripts/04_build_sample.R` writes `analytic_sample.parquet`. This notebook no
longer writes it — the sample is a build artifact, not a byproduct of rendering
a document.

```{r}
#| label: sample-summary

tibble(
  quantity = c(
    "plays retained",
    "share of all plays",
    "weeks represented",
    "plays with a tracked target",
    "plays with a coverage label",
    "plays retained only under the scoped rule"
  ),
  value = c(
    comma(nrow(analytic_sample)),
    percent(nrow(analytic_sample) / nrow(plays), accuracy = 0.1),
    as.character(n_distinct(analytic_sample$week)),
    comma(sum(analytic_sample$target_tracked)),
    comma(sum(!is.na(analytic_sample$coverage_class))),
    comma(sum(analytic_sample$has_kinematic_defect))
  )
) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: sample-by-week

analytic_sample |>
  count(week) |>
  left_join(count(plays, week, name = "all_plays"), by = "week") |>
  mutate(retention = n / all_plays) |>
  gt() |>
  fmt_percent(columns = retention, decimals = 1) |>
  gt_theme_538(quiet = TRUE)
```

Retention should be roughly flat across weeks. A week that stands out points at
a collection artifact concentrated in that week rather than at a property of the
game.

```{r}
#| label: sample-columns

tibble(column = names(analytic_sample), type = map_chr(analytic_sample, \(x)
  class(x)[1])) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

`has_kinematic_defect` and `defect_in_window` travel with the sample rather than
being filtered away, so a downstream model can test sensitivity to the
scoped-versus-conservative choice without rebuilding the funnel. Every surviving
play satisfies `defect_in_window == FALSE`; `has_kinematic_defect == TRUE` marks
the plays retained under the scoped rule that the conservative rule would have
dropped.

## 4.5 Not yet filtered

`SAMPLE_RULE` is deliberately question-agnostic: it removes plays that are not
live, not pass attempts, or not measurable. It does **not** remove play types
that are measurable but may be inappropriate for a particular question. All of
the following are currently retained:

- QB spikes — clock management, no intended receiver
- Throwaways — a real throw with no receiver targeted
- Screens — legitimate passes, but separation dynamics differ
- Penalty-nullified plays — real tracking, no official play
- Batted or tipped passes at the line — outcome set before separation matters
- Hail Marys — coverage geometry unrelated to the usual model
- Goal-line plays with extra linemen (1.2)

These are population decisions belonging to a research question, not data-defect
decisions belonging to the pipeline. The intended approach is to flag them in
`sample_flags` without filtering, so each candidate question selects its own
population on top of a shared base. Deferred until the question is settled.

# 5. Where this leaves the project

Decisions taken so far, with counts, are recorded in `notes/decisions.md`, which
is intended to become the Methods section of the December report. The open items
are the research question fork and the coverage-label tradeoff in 4.3.
