---
title: "Defining the arrival point"
subtitle: "STAT 1961 Capstone — measurement decision for the completion model"
author: "David Neumann"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: [brite, ../theme/capstone.scss]
    code-summary: "Show code"
    fig-width: 8
    fig-height: 5
    embed-resources: true
execute:
  warning: false
  message: false
---

# 0. Setup

```{r}
#| label: setup
#| include: false
if (Sys.getenv("RENV_PROJECT") == "") {
  root <- normalizePath("..", mustWork = TRUE)
  Sys.setenv(RENV_PROJECT = root)
  setwd(root)
  source("renv/activate.R")
}

library(tidyverse)
library(arrow)
library(here)
library(gt)
library(gtExtras)
library(scales)

source(here("R", "constants.R"))
source(here("R", "utils.R"))
source(here("R", "arrival.R"))
```

## 0.1 What this notebook is

Every coverage feature in the completion model is anchored to two points:
the ball at the throw, and the ball at arrival. The throw is unambiguous —
`analytic_sample.parquet` carries `f_throw` with the `pass_forward` /
`pass_shovel` precedence already resolved. Arrival is not.

This notebook establishes why the arrival point has to be *derived from the
tracking data* rather than read off an event label, rejects two label-based
alternatives with counts, and validates the definition that replaces them.
The definition itself lives in `R/arrival.R`; `scripts/06_build_throw_frame.R`
applies it at full-season scale.

Like `analysis/01_eda.qmd`, this notebook reads and reports. It writes nothing
to `data/`.

Everything below except §1 and §2 runs on **week 1 only**. That is a
prototyping decision: 890 plays is enough to characterize a measurement and
fast enough to iterate on. §8 lists the full-season checks to re-run once `06`
exists.

```{r}
#| label: load_data

proto_week <- 1L

trk <- open_dataset(here("data", "processed", "tracking"))

plays <- read_parquet(here("data", "processed", "plays.parquet"))
play_index <- read_parquet(here("data", "processed", "play_index.parquet"))
analytic_sample <- read_parquet(here(
  "data",
  "processed",
  "analytic_sample.parquet"
))
```

Ball rows are roughly one-fourteenth of the tracking data, so the whole season
of them collects comfortably into memory. `dis` is carried because it turns out
to be the discriminating column in §7 — the kinematic defect scan in `03`
filters `!is_ball`, so a ball moving at 2 yd/frame was never flagged and the
column has gone unused until now.

```{r}
#| label: ball_rows

ball <- trk |>
  filter(is_ball) |>
  select(game_id, play_id, frame_id, x, y, s, dis) |>
  collect() |>
  cast_keys()

tibble(
  quantity = c(
    "ball rows",
    "missing s",
    "missing dis",
    "median dis (yd/frame)"
  ),
  value = c(
    comma(nrow(ball)),
    as.character(sum(is.na(ball$s))),
    as.character(sum(is.na(ball$dis))),
    sprintf("%.2f", median(ball$dis, na.rm = TRUE))
  )
) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

One missing row in 1.25 million. Ball kinematics are effectively complete,
which is what makes §7 possible.

# 1. The problem

## 1.1 `pass_arrived` is missing, and the missingness tracks the outcome

```{r}
#| label: arrival-availability

analytic_sample |>
  mutate(has_arrival = !is.na(f_pass_arrived)) |>
  count(pass_result, has_arrival) |>
  group_by(pass_result) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  arrange(pass_result, desc(has_arrival)) |>
  gt() |>
  fmt_percent(columns = pct, decimals = 1) |>
  gt_theme_538(quiet = TRUE)
```

Roughly half of all incompletions carry no `pass_arrived` event, against
under one percent of completions. The consequence is not a smaller sample, it
is a different response variable.

```{r}
#| label: base-rate-shift

throws <- analytic_sample |> filter(pass_result %in% c("C", "I", "IN"))

tibble(
  population = c("all thrown passes", "requires pass_arrived"),
  plays = c(nrow(throws), sum(!is.na(throws$f_pass_arrived))),
  completions = c(
    sum(throws$pass_result == "C"),
    sum(throws$pass_result == "C" & !is.na(throws$f_pass_arrived))
  )
) |>
  mutate(completion_rate = completions / plays) |>
  gt() |>
  fmt_percent(columns = completion_rate, decimals = 1) |>
  fmt_number(columns = c(plays, completions), decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

::: {.callout-important appearance="simple"}

Requiring an arrival event moves the completion rate of the modeling
population by roughly 13 percentage points, by deleting about half of all
incompletions. This is selection on the response variable, and unlike the
missing-LOS block in `01_eda.qmd` §2.6 it cannot be handled by stating its
direction in Methods. It has to be avoided.

:::

## 1.2 It is not a collection artifact

Kinematic defects cluster by week (`01_eda.qmd` §2.2), which is what justified
treating them as a data-quality exclusion. Arrival missingness does not
cluster, so it is a labeling convention rather than a week-specific collection
failure — which means no subset of weeks escapes it.

```{r}
#| label: arrival-by-week

analytic_sample |>
  group_by(week) |>
  summarize(
    plays = n(),
    pct_missing_arrival = mean(is.na(f_pass_arrived))
  ) |>
  arrange(desc(pct_missing_arrival)) |>
  gt() |>
  fmt_percent(columns = pct_missing_arrival, decimals = 1) |>
  gt_theme_538(quiet = TRUE)
```

# 2. Rejected: the `pass_outcome_*` family as a substitute

The obvious repair is to fall back on a label that *is* present on
incompletions. `pass_outcome_caught`, `pass_outcome_incomplete`,
`pass_outcome_interception`, and `pass_outcome_touchdown` are mutually
exhaustive in principle, so take whichever fires first.

`pmin(..., na.rm = TRUE)` returns the earliest label present and `NA` only when
all four are absent. Earliest matters on tipped passes, where more than one can
fire.

```{r}
#| label: outcome-anchor

arr <- analytic_sample |>
  select(game_id, play_id, week, pass_result, f_throw, f_pass_arrived) |>
  left_join(
    play_index |>
      select(
        game_id,
        play_id,
        f_pass_outcome_caught,
        f_pass_outcome_incomplete,
        f_pass_outcome_interception,
        f_pass_outcome_touchdown,
        f_out_of_bounds
      ),
    by = c("game_id", "play_id")
  ) |>
  mutate(
    f_outcome = pmin(
      f_pass_outcome_caught,
      f_pass_outcome_incomplete,
      f_pass_outcome_interception,
      f_pass_outcome_touchdown,
      na.rm = TRUE
    )
  )

arr |>
  count(
    pass_result,
    has_arrival = !is.na(f_pass_arrived),
    has_outcome = !is.na(f_outcome)
  ) |>
  arrange(pass_result, desc(n)) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

Coverage is not the problem: an outcome label exists on 99.7% of plays,
including nearly every incompletion that `pass_arrived` misses. The problem is
*where* it fires.

```{r}
#| label: outcome-anchor-displacement

arr |>
  filter(!is.na(f_pass_arrived), !is.na(f_outcome)) |>
  left_join(
    ball |>
      select(game_id, play_id, frame_id, x, y) |>
      rename(x_a = x, y_a = y),
    by = c("game_id", "play_id", "f_pass_arrived" = "frame_id")
  ) |>
  left_join(
    ball |>
      select(game_id, play_id, frame_id, x, y) |>
      rename(x_o = x, y_o = y),
    by = c("game_id", "play_id", "f_outcome" = "frame_id")
  ) |>
  mutate(
    gap_frames = f_outcome - f_pass_arrived,
    d_ball = sqrt((x_o - x_a)^2 + (y_o - y_a)^2),
    dx_ball = x_o - x_a
  ) |>
  group_by(pass_result) |>
  summarize(
    plays = n(),
    p50_gap_frames = median(gap_frames),
    p50_d_ball = median(d_ball, na.rm = TRUE),
    p90_d_ball = quantile(d_ball, 0.90, na.rm = TRUE),
    p50_dx_ball = median(dx_ball, na.rm = TRUE),
    .groups = "drop"
  ) |>
  gt() |>
  fmt_number(columns = starts_with("p50_d"), decimals = 2) |>
  fmt_number(columns = p90_d_ball, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

::: {.callout-warning appearance="simple"}

**Rejected.** On plays carrying both labels, the ball sits a median of
$\approx 1.9$ yd (completions) and $\approx 3.5$ yd (incompletions) away from
where `pass_arrived` put it, with p90 around 4.4 and 8.1 yd. The displacement
is systematically downfield (`p50_dx_ball` positive) and its magnitude depends
on the outcome.

The coverage features work at a scale of a couple of yards (separation, a
defender's reach). An anchor whose p90 error is 8 yd cannot support them. The physical story: `pass_outcome_incomplete` fires when the
ball hits the turf, which is well past the receiver.

:::

# 3. The kinematic approach series

With both label families rejected, arrival has to be *measured*. The raw
material is the distance from the ball to the targeted receiver on every frame
after the throw.

Three choices are embedded in that sentence and each is a decision.

**Whose position.** The targeted receiver, so this requires `target_tracked`.
Plays without a tracked target are throwaways and spikes, and they are
disproportionately incomplete — so `keep_target` is outcome-correlated. That is
acceptable because it is a *population definition* (the question is about
targeted passes), not a measurement failure, and it belongs in the `08` funnel
framed that way. It is not acceptable to smuggle the same selection in through
a measurement rule, which is what §1 would have done.

**Search window.** Capped at `MAX_FLIGHT_FRAMES` = 50 frames, i.e. 5.0 s.
Observed flight-time p99 is under 3.5 s (§7.2), so the cap only bounds
pathological cases.

**Which frame in the window.** This is the whole question, and §4 through §7
work through three answers.

```{r}
#| label: build-approach

targ <- analytic_sample |>
  filter(target_tracked, week == proto_week) |>
  select(
    game_id,
    play_id,
    week,
    pass_result,
    f_throw,
    f_pass_arrived,
    target_nfl_id
  )

rec <- trk |>
  filter(!is_ball, week == proto_week) |>
  select(game_id, play_id, frame_id, nfl_id, x, y) |>
  collect() |>
  cast_keys() |>
  inner_join(
    targ |> select(game_id, play_id, target_nfl_id),
    by = c("game_id", "play_id", "nfl_id" = "target_nfl_id")
  ) |>
  select(game_id, play_id, frame_id, x, y)

appr <- build_approach(
  ball = ball |> select(game_id, play_id, frame_id, x, y, dis),
  receiver = rec,
  throws = targ
)

tibble(
  quantity = c(
    "plays in week 1 sample",
    "plays with a tracked target",
    "plays with an approach series",
    "frames in the search window"
  ),
  n = c(
    sum(analytic_sample$week == proto_week),
    nrow(targ),
    n_distinct(appr$game_id, appr$play_id),
    nrow(appr)
  )
) |>
  gt() |>
  fmt_number(columns = n, decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

# 4. Rejected: first local minimum

The intuitive rule is "the frame where the gap stops shrinking":

$$f_{\text{arr}} = \min\{f \in \mathcal{F} : d_{f+1} > d_f\}$$

There is no $z$ coordinate in this dataset, so ground contact is invisible and
a *global* minimum can land after a bounce. A local minimum is the end of the
flight by construction. That was the reasoning; the data rejects it.

```{r}
#| label: v1-local-minimum

arrival_v1 <- appr |>
  group_by(game_id, play_id) |>
  summarize(
    n_searched = n(),
    i_local = {
      rise <- which(diff(d) > 0)[1]
      if (is.na(rise)) length(d) else rise
    },
    f_arr = frame_id[i_local],
    d_arr = d[i_local],
    hit_edge = i_local == n_searched,
    .groups = "drop"
  ) |>
  inner_join(
    targ |> select(game_id, play_id, pass_result, f_throw),
    by = c("game_id", "play_id")
  ) |>
  mutate(t_flight = (f_arr - f_throw) / TRACKING_HZ)

arrival_v1 |>
  group_by(pass_result) |>
  summarize(
    plays = n(),
    p50_d_arr = median(d_arr),
    p90_d_arr = quantile(d_arr, 0.90),
    max_d_arr = max(d_arr),
    edge = sum(hit_edge),
    under_0.2s = sum(t_flight < 0.2),
    .groups = "drop"
  ) |>
  gt() |>
  fmt_number(columns = starts_with(c("p50", "p90", "max")), decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

::: {.callout-warning appearance="simple"}

**Rejected.** Three failure modes, each visible in a different column.

1. **Early trip.** Flight times of a single frame. `diff(d) > 0` fires on one
   frame of measurement jitter just after release, and the search stops.
2. **Plateau ambiguity.** After a catch the ball travels *with* the receiver,
   so $d$ sits near zero for many frames and which one shows the first tiny
   uptick is decided by noise.
3. **Never closes.** On badly thrown passes $d$ is still decreasing at the
   window edge, so the rule returns the boundary as a default.

Face validity fails on the class where it should be trivial: on a completion
the ball ends up in the receiver's hands, so `max_d_arr` for `C` ought to be
small, and it is not.

:::

# 5. Rejected: closest approach within a tolerance

Anchoring on the *global* minimum fixes failure modes 1 and 3, because a single
noisy frame can no longer end the search. The plateau then needs resolving
toward the catch rather than the carry, which suggests taking the earliest
frame that is essentially as close as the minimum:

$$f_{\text{arr}} = \min\{f \in \mathcal{F} : d_f \le \min_{g} d_g + \epsilon\},
\qquad \epsilon = 0.25 \text{ yd}$$

```{r}
#| label: v2-tolerance

EPS_V2 <- 0.25

arrival_v2 <- appr |>
  group_by(game_id, play_id) |>
  summarize(
    n_searched = n(),
    i_arr = which(d <= min(d) + EPS_V2)[1],
    i_min = which.min(d),
    f_arr = frame_id[i_arr],
    d_arr = d[i_arr],
    x_arr = x_ball[i_arr],
    y_arr = y_ball[i_arr],
    d_min = d[i_min],
    frames_earlier = i_min - i_arr,
    .groups = "drop"
  )

arrival_v2 |>
  inner_join(
    targ |> select(game_id, play_id, pass_result),
    by = c("game_id", "play_id")
  ) |>
  group_by(pass_result) |>
  summarize(
    plays = n(),
    p50_d_arr = median(d_arr),
    p90_d_arr = quantile(d_arr, 0.90),
    max_d_arr = max(d_arr),
    p50_earlier = median(frames_earlier),
    p90_earlier = quantile(frames_earlier, 0.90),
    corrected = sum(frames_earlier > 0),
    .groups = "drop"
  ) |>
  gt() |>
  fmt_number(columns = starts_with(c("p50_d", "p90_d", "max")), decimals = 2) |>
  fmt_number(columns = starts_with(c("p50_e", "p90_e")), decimals = 1) |>
  gt_theme_538(quiet = TRUE)
```

Face validity is now excellent: every week-1 completion has the ball within
about 1.1 yd of the targeted receiver at the selected frame, against 28 yd
under §4. And the tolerance is doing real work — it moves the frame on the
large majority of completions, by a median of several frames, while doing
almost nothing on incompletions where there is no carry phase.

That asymmetry is the tell, and it is what §6 investigates.

## 5.1 The tolerance fails its own sensitivity check

$\epsilon = 0.25$ yd was introduced as a numerical tolerance — a tie-break
across a flat plateau, on the order of the coordinate rounding scale. If that
were true, the selected frame would be insensitive to it.

```{r}
#| label: eps-sensitivity

eps_grid <- c(0.10, 0.25, 0.50, 1.00)

eps_sens <- map(eps_grid, \(e) {
  appr |>
    group_by(game_id, play_id) |>
    summarize(
      i = which(d <= min(d) + e)[1],
      f = frame_id[i],
      dd = d[i],
      .groups = "drop"
    ) |>
    mutate(eps = e)
}) |>
  list_rbind()

eps_sens |>
  select(game_id, play_id, eps, f) |>
  pivot_wider(names_from = eps, values_from = f, names_prefix = "f_") |>
  summarize(
    `frame agreement, 0.10 vs 0.25` = mean(f_0.1 == f_0.25),
    `frame agreement, 0.50 vs 0.25` = mean(f_0.5 == f_0.25),
    `median shift, 0.10` = median(f_0.1 - f_0.25),
    `median shift, 0.50` = median(f_0.5 - f_0.25)
  ) |>
  pivot_longer(everything(), names_to = "statistic") |>
  gt() |>
  fmt_number(columns = value, decimals = 3) |>
  gt_theme_538(quiet = TRUE)
```

::: {.callout-warning appearance="simple"}

**Rejected, on two counts.**

Halving $\epsilon$ from 0.25 to 0.10 changes the selected frame on the
majority of plays. So there is no plateau to break ties on: $d$ shrinks
gradually through the catch, and $\epsilon$ is not absorbing rounding, it is
*choosing how close counts as arrived*. That is a substantive modeling
parameter wearing a numerical-tolerance costume.

Separately, the criterion is relative. Since $\min d \approx 0$ on completions
and $\approx 1$ yd on incompletions, $d \le \min d + \epsilon$ applies a
different absolute standard to each outcome class — the exact property this
whole exercise exists to eliminate.

:::

# 6. Why the tolerance was needed: flight versus carry

§5 showed the correction is large on completions and absent on incompletions.
The ball's own per-frame displacement says why. Indexing every frame relative
to the §5 anchor, and reading it as a profile rather than a summary:

```{r}
#| label: flight-carry-profile

appr |>
  inner_join(
    arrival_v2 |> select(game_id, play_id, f_arr),
    by = c("game_id", "play_id")
  ) |>
  inner_join(
    targ |> select(game_id, play_id, pass_result),
    by = c("game_id", "play_id")
  ) |>
  mutate(rel = frame_id - f_arr) |>
  filter(pass_result %in% c("C", "I"), between(rel, -8, 3)) |>
  group_by(pass_result, rel) |>
  summarize(p50_dis = median(dis, na.rm = TRUE), n = n(), .groups = "drop") |>
  pivot_wider(names_from = pass_result, values_from = c(p50_dis, n)) |>
  gt() |>
  fmt_number(columns = starts_with("p50"), decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

Both classes fly at $\approx 2.1$ yd/frame, i.e. $\approx 21$ yd/s. Then
completions decay to $\approx 0.6$ by the anchor frame while incompletions hold
flight speed straight through it.

This is a football being caught. On a completion the §5 anchor sits several
frames *after* the ball stopped flying, and those frames are the receiver
gathering the ball in. On an incompletion there is nothing to gather.

## 6.1 The two regimes separate cleanly

```{r}
#| label: ball-dis-bimodality

dis_window <- appr |> filter(!is.na(dis))

quantile(dis_window$dis, c(0.05, 0.10, 0.25, 0.50, 0.75, 0.90, 0.95)) |>
  enframe(name = "quantile", value = "dis") |>
  gt() |>
  fmt_number(columns = dis, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: ball-dis-histogram
#| fig-height: 4

ggplot(dis_window, aes(dis)) +
  geom_histogram(binwidth = 0.05, fill = "lightsteelblue3") +
  geom_vline(xintercept = DIS_FLIGHT, colour = "firebrick", linewidth = 0.7) +
  coord_cartesian(xlim = c(0, 4)) +
  labs(
    title = "Ball displacement per frame, throw to +5 s",
    subtitle = paste0(
      "Bimodal: carry and dead ball below ~1.0, flight at ~2.1. ",
      "Red line is DIS_FLIGHT = ",
      sprintf("%.2f", DIS_FLIGHT)
    ),
    x = "yd / frame",
    y = "frames"
  ) +
  theme_minimal()
```

Two well-separated modes with a thin valley between them, floor around
1.0–1.2 yd/frame. This figure is the justification for the threshold in §7 and
belongs in the written report.

# 7. Adopted: the last in-flight frame

$$f_{\text{arr}} = \max\{f \le f_{\min} : \texttt{dis}_f \ge \delta\},
\qquad f_{\min} = \arg\min_{f \in \mathcal{F}} d_f,
\qquad \delta = \texttt{DIS\_FLIGHT}$$

The last frame on which the ball is still unambiguously in flight.

$\delta$ is defensible where $\epsilon$ was not, because it separates two
physically distinct regimes rather than picking a point on a continuum.
`DIS_FLIGHT` equals `MAX_DIS` **by argument, not by coincidence**: `MAX_DIS` is
the per-frame displacement no *player* can exceed, so a ball above it cannot be
in anyone's hands. §6.1 confirms the threshold lands in the valley.

One frame now serves as both the lane endpoint and the flight-time endpoint,
and no term in the definition refers to the outcome.

```{r}
#| label: v3-detect

arrival_v3 <- detect_arrival(appr) |>
  inner_join(
    targ |> select(game_id, play_id, pass_result, f_throw, f_pass_arrived),
    by = c("game_id", "play_id")
  ) |>
  mutate(t_flight = (f_arr - f_throw) / TRACKING_HZ)
```

## 7.1 Class similarity, not proximity to zero

Under §5 the target was $d_{\text{arr}} \to 0$. That is the wrong target: a
caught ball genuinely gets closer to the receiver than a dropped one, so an
anchor that forces both to zero is measuring the outcome. The right target is
$d_{\text{arr}}$ being *similar across classes*, with closest approach retained
separately as the face-validity check.

```{r}
#| label: v3-class-similarity

arrival_v3 |>
  group_by(pass_result) |>
  summarize(
    plays = n(),
    p50_d_arr = median(d_arr),
    p90_d_arr = quantile(d_arr, 0.90),
    p50_d_min = median(d_min),
    p50_to_min = median(frames_to_min),
    p90_to_min = quantile(frames_to_min, 0.90),
    fallbacks = sum(used_fallback),
    edge = sum(at_window_edge),
    .groups = "drop"
  ) |>
  gt() |>
  fmt_number(columns = starts_with("p50_d"), decimals = 2) |>
  fmt_number(columns = p90_d_arr, decimals = 2) |>
  fmt_number(columns = ends_with("to_min"), decimals = 1) |>
  gt_theme_538(quiet = TRUE)
```

`p50_d_min` on completions confirms that closest approach is the catch, so the
underlying series is right and only the frame selection changed.
`frames_to_min` is the catch-plus-carry duration, not a gather time — $f_{\min}$
is the deepest point of the carry, since $d$ stays near zero while the receiver
holds the ball.

## 7.2 Flight time

```{r}
#| label: v3-flight-time

arrival_v3 |>
  summarize(
    p01 = quantile(t_flight, 0.01),
    p50 = quantile(t_flight, 0.50),
    p99 = quantile(t_flight, 0.99),
    max = max(t_flight),
    `under 0.4 s` = sum(t_flight < 0.4)
  ) |>
  pivot_longer(everything(), names_to = "statistic") |>
  gt() |>
  fmt_number(columns = value, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

p99 under 3.5 s is what justifies the 50-frame window cap. The handful of
sub-0.4 s flights are not rule failures — a shovel pass or a ball batted at the
line genuinely has a one-frame flight, and the beyond-the-LOS population filter
in `08` removes them.

```{r}
#| label: v3-short-flights

arrival_v3 |>
  filter(t_flight < 0.4) |>
  left_join(
    analytic_sample |> select(game_id, play_id, throw_anchor),
    by = c("game_id", "play_id")
  ) |>
  count(pass_result, throw_anchor, t_flight) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

## 7.3 The decisive check: implied ball speed by depth

A measurement artifact in the arrival *timing* would show up as an
outcome-dependent implied ball speed. Flight time legitimately differs by
outcome — deep passes are both longer in the air and completed less often — so
the test has to condition on lane length. 2-yard bins, which is tight enough
that median lane length matches across classes within a bin.

```{r}
#| label: v3-speed-equalization

lanes_v3 <- arrival_v3 |>
  left_join(
    ball |>
      select(game_id, play_id, frame_id, x, y) |>
      rename(x0 = x, y0 = y),
    by = c("game_id", "play_id", "f_throw" = "frame_id")
  ) |>
  mutate(
    lane_len = sqrt((x_arr - x0)^2 + (y_arr - y0)^2),
    v_ball = lane_len / t_flight
  )

lanes_v3 |>
  filter(pass_result %in% c("C", "I"), between(lane_len, 8, 30)) |>
  mutate(bin = cut(lane_len, seq(8, 30, by = 2))) |>
  group_by(bin, pass_result) |>
  summarize(
    n = n(),
    p50_lane = median(lane_len),
    p50_v = median(v_ball),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = pass_result,
    values_from = c(n, p50_lane, p50_v)
  ) |>
  gt() |>
  fmt_number(columns = starts_with("p50"), decimals = 1) |>
  gt_theme_538(quiet = TRUE)
```

::: {.callout-important appearance="simple"}

**This is the result the definition rests on.** Implied ball speed agrees
between completions and incompletions to within about 1 yd/s in every depth
bin, and sits flat at $\approx 21$ yd/s across depth — a physically sensible
constant for a thrown football.

Under the §5 anchor the same table split 12.0 against 17.0 yd/s at short depth,
converging only past 26 yd, which is the signature of a fixed number of
contaminated frames rather than a football fact.

Flat-across-depth is the stronger half of the result: the definition does not
merely equalize two classes, it recovers a quantity that behaves the way ball
flight should.

:::

## 7.4 Sensitivity: the flight threshold and the gap tolerance

$\delta$ first. Two questions: does the selected frame move when the threshold
moves, and does the speed equalization in §7.3 survive.

```{r}
#| label: delta-sensitivity

delta_grid <- c(1.0, 1.3, 1.6, 1.9)

delta_sens <- map(delta_grid, \(dd) {
  detect_arrival(appr, dis_flight = dd) |>
    select(game_id, play_id, f_arr, d_arr) |>
    mutate(delta = dd)
}) |>
  list_rbind()

delta_sens |>
  group_by(delta) |>
  summarize(p50_d_arr = median(d_arr), p90_d_arr = quantile(d_arr, 0.90)) |>
  gt() |>
  fmt_number(columns = starts_with("p"), decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: delta-agreement

delta_sens |>
  select(game_id, play_id, delta, f_arr) |>
  pivot_wider(names_from = delta, values_from = f_arr, names_prefix = "f_") |>
  summarize(
    `agreement, 1.0 vs 1.3` = mean(f_1 == f_1.3),
    `agreement, 1.6 vs 1.3` = mean(f_1.6 == f_1.3),
    `median shift, 1.0` = median(f_1 - f_1.3),
    `median shift, 1.6` = median(f_1.6 - f_1.3)
  ) |>
  pivot_longer(everything(), names_to = "statistic") |>
  gt() |>
  fmt_number(columns = value, decimals = 3) |>
  gt_theme_538(quiet = TRUE)
```

Agreement well above 0.8 across a nearly two-fold range of $\delta$, with
median shift zero — against 0.38 for the $\epsilon$ check in §5.1. That is the
difference between separating two regimes and picking a point on a continuum.

Now the gap tolerance. `flight_end_index()` merges flight-speed runs separated
by at most `ARRIVAL_GAP_TOL` frames. The two extremes are both wrong:
tolerance 0 lets one frame of jitter on a wobbling ball end the flight early,
while an unbounded tolerance lets a jittery frame during a catch-and-run be
selected as the flight end.

```{r}
#| label: gap-tolerance-extremes

# Quantifying the disagreement the tolerance exists to split. The
# unbounded rule is recovered at a tolerance wider than any window.
gap_extremes <- inner_join(
  detect_arrival(appr, gap_tol = 0L) |>
    select(game_id, play_id, f_run = f_arr),
  detect_arrival(appr, gap_tol = MAX_FLIGHT_FRAMES) |>
    select(game_id, play_id, f_last = f_arr),
  by = c("game_id", "play_id")
)

gap_extremes |>
  summarize(
    plays = n(),
    differ = sum(f_run != f_last),
    pct_differ = mean(f_run != f_last),
    p50_gap = median((f_last - f_run)[f_run != f_last]),
    max_gap = max(f_last - f_run)
  ) |>
  gt() |>
  fmt_percent(columns = pct_differ, decimals = 1) |>
  fmt_number(columns = c(plays, differ, p50_gap, max_gap), decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: gap-tolerance-sensitivity

gap_grid <- c(0L, 1L, 2L, 3L, 5L)

gap_sens <- map(gap_grid, \(g) {
  detect_arrival(appr, gap_tol = g) |>
    inner_join(
      targ |> select(game_id, play_id, pass_result),
      by = c("game_id", "play_id")
    ) |>
    group_by(pass_result) |>
    summarize(
      p50_d_arr = median(d_arr),
      p90_d_arr = quantile(d_arr, 0.90),
      .groups = "drop"
    ) |>
    mutate(gap_tol = g)
}) |>
  list_rbind()

gap_sens |>
  filter(pass_result %in% c("C", "I")) |>
  pivot_wider(
    names_from = pass_result,
    values_from = c(p50_d_arr, p90_d_arr)
  ) |>
  relocate(gap_tol) |>
  gt() |>
  fmt_number(columns = -gap_tol, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

::: {.callout-note appearance="simple"}

`ARRIVAL_GAP_TOL` is set from this table. The value to pick is the one where
$d_{\text{arr}}$ is stable in `gap_tol` and similar between classes; if the
table is flat across the grid, the parameter does not matter and the choice
costs nothing to defend. Record the chosen value and this table in
`notes/decisions.md` §6.

:::

## 7.5 Bounce check

Anchoring the search at $f_{\min}$ opens one door: a ball that skips off the
turf and comes back *closer* to the receiver than it was in flight would put
$f_{\min}$ on the bounce. There is no $z$ coordinate, so the test is the
signature — $d$ rising substantially between the arrival frame and the minimum.

```{r}
#| label: bounce-check

appr |>
  inner_join(
    arrival_v3 |> select(game_id, play_id, f_arr, f_min),
    by = c("game_id", "play_id")
  ) |>
  filter(frame_id >= f_arr, frame_id <= f_min) |>
  group_by(game_id, play_id) |>
  summarize(max_rise = max(d) - first(d), .groups = "drop") |>
  summarize(
    plays = n(),
    `rise > 1 yd` = sum(max_rise > 1),
    `rise > 3 yd` = sum(max_rise > 3),
    p99_rise = quantile(max_rise, 0.99)
  ) |>
  pivot_longer(everything(), names_to = "statistic") |>
  gt() |>
  fmt_number(columns = value, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

A handful of plays and a p99 under a yard. The ball moves monotonically toward
its closest approach, so no flight/carry discriminator beyond $\delta$ is
needed and the $f_{\min}$ bound is a belt-and-braces guard rather than a load
-bearing one.

## 7.6 Relationship to `pass_arrived`, where it exists

Not a target to hit — §1 established the label is the inferior anchor — but the
disagreement should be *explainable*, and the direction of the explanation
matters.

```{r}
#| label: v3-vs-label

val <- arrival_v3 |>
  filter(!is.na(f_pass_arrived)) |>
  left_join(
    ball |>
      select(game_id, play_id, frame_id, x, y) |>
      rename(x_lab = x, y_lab = y),
    by = c("game_id", "play_id", "f_pass_arrived" = "frame_id")
  ) |>
  mutate(
    gap_frames = f_arr - f_pass_arrived,
    d_ball = sqrt((x_arr - x_lab)^2 + (y_arr - y_lab)^2)
  )

val |>
  group_by(pass_result) |>
  summarize(
    plays = n(),
    p50_gap = median(gap_frames),
    p50_d_ball = median(d_ball, na.rm = TRUE),
    p90_d_ball = quantile(d_ball, 0.90, na.rm = TRUE),
    .groups = "drop"
  ) |>
  gt() |>
  fmt_number(columns = starts_with("p50_d"), decimals = 2) |>
  fmt_number(columns = p90_d_ball, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

The label fires at a *consistent* ball-to-receiver distance across outcome
classes — about a yard out — so the labelers are doing something coherent:
marking the ball arriving in the receiver's vicinity.

```{r}
#| label: label-distance

appr |>
  select(game_id, play_id, frame_id, d) |>
  inner_join(
    arrival_v3 |> select(game_id, play_id, pass_result, f_pass_arrived, d_arr),
    by = c("game_id", "play_id", "frame_id" = "f_pass_arrived")
  ) |>
  group_by(pass_result) |>
  summarize(
    plays = n(),
    p50_d_at_label = median(d),
    p90_d_at_label = quantile(d, 0.90),
    p50_d_at_arr = median(d_arr),
    .groups = "drop"
  ) |>
  gt() |>
  fmt_number(columns = starts_with("p"), decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

What differs by outcome is what happens *after* that yard. On a completion the
ball keeps coming until it is in the hands, which is a few frames later; on an
incompletion nobody gathers it, so the label and the flight end coincide. Same
uniform definition, different amount of remaining flight, because the outcome
determines how close the ball ultimately gets.

The check that this is lateness rather than a different event: on plays where
the label fires first, is the ball still closing?

```{r}
#| label: label-monotonicity

appr |>
  inner_join(
    val |>
      filter(gap_frames > 0) |>
      select(game_id, play_id, pass_result, f_pass_arrived, f_arr),
    by = c("game_id", "play_id")
  ) |>
  filter(frame_id >= f_pass_arrived, frame_id <= f_arr) |>
  group_by(game_id, play_id, pass_result) |>
  summarize(
    monotone = all(diff(d) <= 0),
    net_change = last(d) - first(d),
    .groups = "drop"
  ) |>
  group_by(pass_result) |>
  summarize(
    plays = n(),
    pct_monotone = mean(monotone),
    p50_net = median(net_change),
    p90_net = quantile(net_change, 0.90),
    .groups = "drop"
  ) |>
  gt() |>
  fmt_percent(columns = pct_monotone, decimals = 1) |>
  fmt_number(columns = starts_with("p50_net"), decimals = 2) |>
  fmt_number(columns = p90_net, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

The gap closes monotonically on the large majority of plays and `p50_net` is
negative in every class: the ball was still approaching when the label fired
and kept approaching until the adopted frame. The label is early, and this
anchor is the later and physically correct one.

Interceptions are a small class here and genuinely ambiguous — tipped balls,
defenders carrying the ball away. Noted, not tuned to.

# 8. What carries forward

**Adopted.** `detect_arrival()` in `R/arrival.R`, with `DIS_FLIGHT`,
`ARRIVAL_GAP_TOL`, and `MAX_FLIGHT_FRAMES` in `R/constants.R`. Applied at full
-season scale by `scripts/06_build_throw_frame.R`.

**Diagnostics travel with the output, and are not filtered on:** `d_arr`,
`d_min`, `f_min`, `frames_to_min`, `used_fallback`, `at_window_edge`. Same
pattern as `has_kinematic_defect` / `defect_in_window` in the sample layer, so
a model can test sensitivity without rebuilding.

::: {.callout-important appearance="simple"}

**`d_arr` and `d_min` are never model features.** A ball ending up a tenth of a
yard from the receiver *is* the catch. Either one would drive log loss down
while saying nothing about whether coverage geometry carries information.

Equally, badly thrown passes are **not filtered on `d_arr`**. A pass that never
got near its target is a real play with real coverage geometry, and it is
almost by definition an incompletion — excluding it would reintroduce exactly
the selection on outcome that §1 rejected.

:::

**Full-season checks**, run in `analysis/04_arrival_season.qmd` once `06`
existed; all pass (`notes/decisions.md` §6.9):

1. §7.1 class similarity, all 17 weeks. The week-1 numbers are the benchmark.
2. §7.3 speed equalization, all 17 weeks. The result the definition rests on.
3. §7.5 bounce check, since a rarer artifact may only appear at 17× the sample.
4. Three plays rendered with `render_play()` — one completion, one incompletion,
   one interception — stepped to `f_arr` to confirm visually that the ball is
   where the number says it is. A sign error or an off-by-one is invisible in a
   summary table and obvious in an animation.

**Since resolved** (`notes/decisions.md` §4.8, §7): `keep_arrival` is a
diagnostic only, the beyond-the-LOS threshold is $\text{depth}_{\text{arr}} > 0$,
and `qb_hit` is a covariate. The passing window is time-based rather than a
fixed-width corridor.
