---
title: "Completion probability — baseline models"
subtitle: "STAT 1961 Capstone — preliminary results"
author: "David Neumann"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: [brite, ../theme/capstone.scss]
    code-summary: "Show code"
    fig-width: 8
    fig-height: 5
    embed-resources: true
execute:
  warning: false
  message: false
---

# 0. Setup

```{r}
#| label: setup
#| include: false
if (Sys.getenv("RENV_PROJECT") == "") {
  root <- normalizePath("..", mustWork = TRUE)
  Sys.setenv(RENV_PROJECT = root)
  setwd(root)
  source("renv/activate.R")
}

library(tidyverse)
library(arrow)
library(here)
library(gt)
library(gtExtras)
library(scales)

# Attached after tidyverse deliberately: nlme (pulled in by mgcv) masks
# dplyr::collapse(), which is unused here. mgcv must be ATTACHED rather
# than merely installed, because the MODEL_SPECS formulas are written
# with bare s() and the formula environment has to resolve it.
library(mgcv)

source(here("R", "constants.R"))
source(here("R", "utils.R"))
source(here("R", "pbp.R"))
source(here("R", "evaluate.R"))

fig_dir <- here("figs", "week06")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
```

## 0.1 What this notebook is, and is not

This notebook reports. It reads persisted artifacts and writes nothing to
`data/`, the same contract as `analysis/01_eda.qmd` and
`analysis/02_arrival_anchor.qmd`. Figures are written to `figs/week06/` and
referenced statically from the slide deck.

Two things about it are **provisional**, and both belong on the slide rather
than in a footnote.

::: {.callout-important appearance="simple"}

**The model frame is assembled here, in §1.1.** That assembly is
`scripts/08_build_model_frame.R`'s job and that script does not exist yet.
Nothing is written to disk, so the architectural rule holds, but these results
are not yet reproducible from a build artifact.

**The question's population filters are applied inline, in §1.2**: a targeted
receiver (`notes/decisions.md` §4.9) and a ball that arrives beyond the line of
scrimmage ($\text{depth}_{\text{arr}} > 0$, §7). Spikes and scrambles are
already absent from the analytic sample. §1–8 report the floor, the
play-by-play baseline, and the external benchmark; §9 adds stage 3, separation
at the throw. Stage 4 features do not exist yet.

:::

```{r}
#| label: load_data

analytic_sample <- read_parquet(here(
  "data",
  "processed",
  "analytic_sample.parquet"
))
plays <- read_parquet(here("data", "processed", "plays.parquet"))
pbp <- read_parquet(here("data", "processed", "pbp.parquet"))
arrival <- read_parquet(here("data", "processed", "arrival.parquet"))
features <- read_parquet(here("data", "processed", "features.parquet"))

# pbp.parquet does not carry receiver names yet, so they are read from the
# nflverse mirror. 08 takes this over.
receivers <- read_parquet(
  here("data", "interim", "pbp_2018.parquet"),
  col_select = c("old_game_id", "play_id", "receiver_player_name")
) |>
  standardize_pbp_keys()
```

# 1. The model frame

## 1.1 Assembly

Six sources, joined on `(game_id, play_id)`. The join takes from each table
only the columns the other two lack, because `analytic_sample.parquet` already
carries `down`, `yards_to_go`, `los_x`, and `week` — pulling them again would
produce suffixed `.x` / `.y` pairs rather than an error.

- `analytic_sample.parquet` — the spine, and the response via `pass_result`
- `plays.parquet` — `number_of_pass_rushers`, which `04_build_sample.R` did not
  carry forward because the sample rule did not need it
- `pbp.parquet` — `pass_location`, `qb_hit`, `shotgun`, charted `air_yards`
  (kept for comparison only),
  `posteam_type`, and the benchmark columns `cp` / `has_cp`
- the nflverse mirror — `receiver_player_name`, for the population filter in
  §1.2
- `arrival.parquet` — `depth_arr`, the tracking-derived depth that replaces
  charted `air_yards` (`notes/decisions.md` §6.9), and the arrival diagnostics
- `features.parquet` — the stage 3 features (§9)

`plays.parquet` is not widened by re-running `04`: the analytical sample is a
locked artifact of the sample layer, and adding model covariates to it would
conflate the sample layer with the feature layer.

```{r}
#| label: assemble

throws <- analytic_sample |>
  left_join(
    plays |> select(game_id, play_id, number_of_pass_rushers),
    by = c("game_id", "play_id")
  ) |>
  left_join(pbp, by = c("game_id", "play_id")) |>
  left_join(receivers, by = c("game_id", "play_id")) |>
  left_join(
    select(arrival, game_id, play_id, depth_arr, used_fallback, at_window_edge),
    by = c("game_id", "play_id")
  ) |>
  left_join(
    select(features, game_id, play_id, sep_throw, closing_throw),
    by = c("game_id", "play_id")
  ) |>
  filter(pass_result %in% c("C", "I", "IN"))

tibble(
  quantity = c(
    "plays in analytic_sample",
    "thrown passes (C, I, IN)",
    "removed: sacks carrying a throw event",
    "matched to play-by-play",
    "carrying the cp benchmark"
  ),
  n = c(
    nrow(analytic_sample),
    nrow(throws),
    nrow(analytic_sample) - nrow(throws),
    sum(throws$pbp_matched),
    sum(throws$has_cp)
  )
) |>
  mutate(pct_of_throws = n / nrow(throws)) |>
  gt() |>
  fmt_percent(columns = pct_of_throws, decimals = 1) |>
  fmt_number(columns = n, decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

Sacks are removed because they are not throws. `keep_any_throw` in the sample
rule defines the population as plays where a forward pass was *released*
(`notes/decisions.md` §4.6), and a sack carries a throw event on a small number
of plays; this filter is on the outcome code rather than the event, so it
catches those.

## 1.2 Population: a targeted receiver, beyond the line

nflverse computes `cp` only when a receiver is named, and the question is the
completion probability of a pass *to a receiver*. Throwaways, intentional
grounding, and passes with no identifiable target are out of scope
(`notes/decisions.md` §4.9). The question is also about passes thrown beyond
the line of scrimmage; screens and other throws at or behind it have different
coverage dynamics (§7). Four steps, applied here until
`scripts/08_build_model_frame.R` exists:

- **Population:** nflverse names a receiver.
- **Quality:** BDB's targeted receiver is tracked. Without that, no coverage
  geometry can be measured.
- **Quality:** an arrival is measured (`arrival.parquet`).
- **Population:** the ball arrives beyond the line,
  $\text{depth}_{\text{arr}} > 0$.

Like every exclusion, each step is checked for a differential completion rate.

```{r}
#| label: population

pop <- throws |>
  mutate(
    receiver_named = !is.na(receiver_player_name),
    target_tracked = coalesce(target_tracked, FALSE),
    is_complete = pass_result == "C"
  )

everything <- rep(TRUE, nrow(pop))
named <- pop$receiver_named
tracked <- named & pop$target_tracked
arrived <- tracked & !is.na(pop$depth_arr)
beyond <- arrived & pop$depth_arr > 0

pop_funnel <- tibble(
  step = c(
    "thrown passes",
    "nflverse receiver named (population)",
    "targeted receiver tracked (quality)",
    "arrival measured (quality)",
    "ball arrives beyond the line (population)"
  ),
  keep = list(everything, named, tracked, arrived, beyond),
  before = list(everything, everything, named, tracked, arrived)
) |>
  mutate(
    dropped = map2_int(before, keep, \(b, k) sum(b & !k)),
    remaining = map_int(keep, sum),
    complete_kept = map_dbl(keep, \(k) mean(pop$is_complete[k])),
    complete_dropped = map2_dbl(
      before,
      keep,
      \(b, k) if (any(b & !k)) mean(pop$is_complete[b & !k]) else NA_real_
    )
  ) |>
  select(-keep, -before)

pop_funnel |>
  gt() |>
  fmt_number(columns = c(dropped, remaining), decimals = 0) |>
  fmt_percent(columns = c(complete_kept, complete_dropped), decimals = 1) |>
  sub_missing(missing_text = "") |>
  gt_theme_538(quiet = TRUE)

model_input <- pop[beyond, ]

# The benchmark is defined on every play in the population.
stopifnot(all(model_input$has_cp))
```

The `r comma(pop_funnel$dropped[2])` throws with no named receiver complete at
`r percent(pop_funnel$complete_dropped[2], accuracy = 0.1)`: throwaways,
grounding, and untargeted passes. Excluding them is a definition of the
population, not attrition, and it is the same exclusion nflverse makes for
`cp`. The `r pop_funnel$dropped[3]` targeted receivers who are not tracked
complete at `r percent(pop_funnel$complete_dropped[3], accuracy = 0.1)`, the
opposite skew. That is a measurement loss, small enough to state rather than
correct. `r pop_funnel$dropped[4]` more have no measured arrival.

The beyond-the-line filter removes `r comma(pop_funnel$dropped[5])` throws,
which complete at `r percent(pop_funnel$complete_dropped[5], accuracy = 0.1)`
against `r percent(pop_funnel$complete_kept[5], accuracy = 0.1)` for those kept:
screens and other throws at the line are easy completions. It is a population
definition, stated with its composition.

Every play that remains carries `cp`, so the stages and the benchmark are scored
on the same rows.

## 1.3 Derived predictors

`prepare_model_frame()` in `R/evaluate.R` owns the response and two derived
columns, so that every stage and the benchmark see identical inputs. The
reasoning for each is in that file's header; the observable effects:

```{r}
#| label: prepare

mf <- prepare_model_frame(model_input)

tibble(
  transformation = c(
    "response: complete = 1 for C, 0 for I and IN",
    "number_of_pass_rushers clamped to [2, 7]",
    "dist_to_sticks = depth_arr - yards_to_go",
    "home = posteam_type == 'home'"
  ),
  affected = c(
    sum(mf$complete),
    sum(
      model_input$number_of_pass_rushers != mf$number_of_pass_rushers,
      na.rm = TRUE
    ),
    sum(!is.na(mf$dist_to_sticks)),
    sum(mf$home, na.rm = TRUE)
  )
) |>
  gt() |>
  fmt_number(columns = affected, decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

An interception counts as an incompletion. This is how nflverse `cp` is defined,
and matching it is what makes the benchmark comparison meaningful rather than a
comparison of two different response variables.

## 1.4 The scoring population

Every stage's log loss has to be computed on the same rows, or the *differences*
between stages are not differences. `glm()` and `mgcv::gam()` default to
`na.action = na.omit`, so a stage referencing a column with missingness would
quietly fit and score fewer plays than the stage below it — no warning, no
error, and a delta computed across two different datasets.

`assert_complete()` therefore **asserts rather than filters**. Dropping rows is
an exclusion, exclusions are checked for differential completion rates and
logged with counts (`notes/decisions.md` §4.7), and the modeling layer is not
permitted to change the population silently.

```{r}
#| label: required-vars

tibble(variable = model_vars()) |>
  gt() |>
  tab_header(title = "Columns required by MODEL_SPECS") |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: incomplete-columns

rep <- incomplete_report(mf)
n_incomplete <- sum(!complete.cases(mf[model_vars()]))

rep$by_column |>
  gt() |>
  fmt_percent(columns = pct_missing, decimals = 4) |>
  gt_theme_538(quiet = TRUE)
```

`r n_incomplete` plays in this population are missing a required variable:
`closing_throw` is undefined where the direction of the receiver or the nearest
defender is missing at the throw (`scripts/07_build_features.R`). The
requirement is the union across all specs, so these plays leave every stage,
not just stage 3. Their outcome composition is below.

```{r}
#| label: incomplete-composition

rep$composition |>
  gt() |>
  fmt_percent(columns = share, decimals = 1) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: scoring-population

vars <- model_vars()
mf_complete <- filter(mf, complete.cases(mf[vars]))

assert_complete(mf_complete)

tibble(
  quantity = c(
    "thrown passes",
    "complete cases across all specs",
    "dropped",
    "completions",
    "base rate"
  ),
  value = c(
    comma(nrow(mf)),
    comma(nrow(mf_complete)),
    comma(nrow(mf) - nrow(mf_complete)),
    comma(sum(mf_complete$complete)),
    sprintf("%.4f", mean(mf_complete$complete))
  )
) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

The base rate of `r sprintf("%.4f", mean(mf_complete$complete))` is the number
stage 1 exists to reproduce, and the number every later stage has to beat.

# 2. Resampling

## 2.1 Leave-one-week-out, not random $k$-fold

Plays are not independent. Two plays from the same game share a quarterback, a
defense, a game plan, the weather, and the officiating crew. Under a random
split, one of them can sit in the training set while the other is scored, and
the model predicts the second partly by having memorized conditions from the
first. The resulting estimate is optimistic and **carries no signature** — there
is nothing in the number that reveals the problem.

Grouping the folds by week makes every fold a clean block: fit on 16 weeks,
predict the 17th, rotate. It also mirrors how such a model would actually be
used, which is a better justification than a technical one. The general argument
is in Roberts et al. (2017), *Ecography*.

No seed is needed. With one group per fold there is no random assignment of
groups to folds, so the only thing that could vary is the returned order, and
`make_folds()` sorts.

```{r}
#| label: folds

folds <- make_folds(mf_complete)

tibble(
  held_out_week = folds$held_out,
  train_n = map_int(folds$splits, \(s) nrow(rsample::analysis(s))),
  test_n = map_int(folds$splits, \(s) nrow(rsample::assessment(s)))
) |>
  gt() |>
  fmt_number(columns = c(train_n, test_n), decimals = 0) |>
  grand_summary_rows(
    columns = test_n,
    fns = list(total = ~ sum(.)),
    fmt = ~ fmt_number(., decimals = 0)
  ) |>
  gt_theme_538(quiet = TRUE)
```

The `test_n` total equals the scoring population exactly, which is the property
that makes the pooled out-of-fold predictions in §4 a complete prediction for
every play: each one is predicted exactly once, by a fit that never saw its
week.

# 3. Stage 2, interpreted

Two different jobs, kept separate. The 17 cross-validation fits answer *how well
does this predict* and are discarded — their only product is the prediction
table in §4. A single fit on all the data answers *what did it learn*, and that
is the object below.

::: {.callout-warning appearance="simple"}

**Nothing in this section is out-of-sample.** No metric reported here is a
performance estimate. Section 5 onward is where performance is measured.

:::

```{r}
#| label: full-fit

f2 <- fit_full(mf_complete, MODEL_SPECS[["2. play-by-play"]])

summary(f2)
```

## 3.1 Reading the parametric terms

```{r}
#| label: parametric-terms

pt <- summary(f2)$p.table
b <- function(term, col = "Estimate", digits = 2) {
  sprintf(paste0("%.", digits, "f"), pt[term, col])
}

pt |>
  as.data.frame() |>
  rownames_to_column("term") |>
  as_tibble() |>
  rename(estimate = Estimate, se = `Std. Error`, p = `Pr(>|z|)`) |>
  select(term, estimate, se, p) |>
  arrange(p) |>
  gt() |>
  fmt_number(columns = c(estimate, se), decimals = 3) |>
  fmt_scientific(columns = p, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

One of these is worth reporting as a finding rather than a control. `qb_hit` is
no longer a single coefficient; see §3.2.

Depth is the tracking-derived $\text{depth}_{\text{arr}}$, not charted
`air_yards`. The earlier `air_yards == 0` indicator is gone with it: it captured
the charting convention of recording throws at the line as exactly zero
(`notes/decisions.md` §6.9), and the tracking depth is continuous.

**`shotgun` at `r b("shotgun", digits = 3)`** (SE
`r b("shotgun", "Std. Error", 3)`), which is the opposite sign from the
conventional wisdom that passing statistics are inflated out of shotgun. Both
can be true: shotgun throws are systematically shorter, so the raw advantage is
confounded by depth, and conditioning on depth inverts it. This coefficient was
*not* estimable in the first specification, which also included
`offense_formation`: an empty-backfield snap is essentially always from shotgun,
so the two were aliased and the standard error on `shotgun` was $0.349$.

`pass_location` is kept at three levels. Against left, middle is
`r b("pass_locationmiddle")` and right `r b("pass_locationright")`
($p$ = `r b("pass_locationright", "Pr(>|z|)")`): middle differs from both
sides, while left and right are not distinguishable in this population. The
coding is kept by argument rather than revised on a $p$-value
(`notes/decisions.md` §9.3).

## 3.2 Reading the smooths

`edf` is the effective degrees of freedom the penalty actually chose. An `edf`
near $1$ means the penalty shrank the smooth to a straight line — the term is in
the model but the relationship is linear.

```{r}
#| label: smooth-terms

st <- summary(f2)$s.table

st |>
  as.data.frame() |>
  rownames_to_column("smooth") |>
  as_tibble() |>
  rename(p = `p-value`) |>
  gt() |>
  fmt_number(columns = c(edf, Ref.df, `Chi.sq`), decimals = 2) |>
  fmt_scientific(columns = p, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

`s(dist_to_sticks)` comes back weak: `edf`
`r sprintf("%.2f", st["s(dist_to_sticks)", "edf"])`,
$p$ = `r sprintf("%.2f", st["s(dist_to_sticks)", "p-value"])`. The motivation was
sound — additive smooths in depth and `yards_to_go` cannot represent their
difference, which is an interaction, and a 3-yard throw on 3rd-and-2 is a
different play from a 3-yard throw on 3rd-and-15 — but the data give it little
support. Conditional on depth, whether the throw reaches the line to
gain barely affects whether it is caught. In hindsight this is sensible: the
sticks govern *play calling*, which is upstream of the throw rather than a
property of it.

**It stays in the model anyway.** Stage 2 exists to be a strong baseline, and
pruning a baseline on in-sample significance is what makes a nested comparison
look tuned. It costs one effective degree of freedom out of
`r comma(nrow(mf_complete))` observations.
It also spans the `yards_to_go` direction, so dropping it would remove that
variable from the model entirely.

**The QB-hit effect varies with depth.** `qb_hit` enters as
`s(depth_arr, by = qb_hit)`: a smooth in depth that applies only to plays
with a hit, so the hit's effect on the logit is a function of depth. A
diagnosis of an earlier specification, which held the hit effect constant,
found its effect fading from about $-1.6$ behind the line to about $-0.3$ on
throws of 20+ yards, while the model applied about $-0.8$ everywhere; that
over-spread the predictions for deep throws (`notes/decisions.md` §9.3). A
plausible mechanism: `qb_hit` records a hit on the play, not its timing, and
on deep throws the hit more often comes after the release.

```{r}
#| label: hit-by-depth

hit_depths <- c(2, 10, 20, 30)
hit_nd <- mf_complete[rep(1, length(hit_depths)), ] |>
  mutate(depth_arr = hit_depths, qb_hit = 1)
hit_tm <- predict(f2, newdata = hit_nd, type = "terms", se.fit = TRUE)
hit_eff <- tibble(
  depth_arr = hit_depths,
  effect = hit_tm$fit[, "s(depth_arr):qb_hit"],
  se = hit_tm$se.fit[, "s(depth_arr):qb_hit"]
)

hit_eff |>
  gt() |>
  tab_header(title = "Effect of a QB hit on the logit, by depth") |>
  fmt_number(columns = c(effect, se), decimals = 2) |>
  gt_theme_538(quiet = TRUE)

he <- function(a) sprintf("%.2f", hit_eff$effect[hit_eff$depth_arr == a])
```

A hit changes the logit by `r he(2)` at 2 yards, `r he(10)` at 10 yards,
`r he(20)` at 20 yards, and `r he(30)` at 30. The band widens at the deep end,
where few hit plays reach (see the smooth plots below).

Including `qb_hit` at all, rather than excluding hit plays, follows
`notes/decisions.md` §7: the flag does not distinguish "hit during the throw"
from "hit afterwards," so filtering on it would remove real passes and may
itself select on outcome.

```{r}
#| label: gam-check

gam.check(f2)
kc <- k.check(f2)
```

The basis dimension check is the part to read. `k-index` below $1$ with a small
$p$-value suggests the basis is too small — *especially* when `edf` sits close to
`k'`. Every `edf` sits well below its `k'`: the largest is
`r sprintf("%.2f", max(kc[, "edf"]))`, for
`r rownames(kc)[which.max(kc[, "edf"])]` with `k'` of
`r kc[which.max(kc[, "edf"]), "k'"]`. The default bases are adequate. (For
`number_of_pass_rushers`, which takes six values, the `k-index` test is
unreliable in any case: it compares residuals of neighbouring observations,
and most neighbours share a value.)

Do not over-read the residual plots. Deviance residuals from a binomial fit on a
binary response always separate into two arcs; that is a property of binary data
rather than a diagnosis.

```{r}
#| label: smooth-plots
#| fig-height: 6

par(mfrow = c(2, 3))
plot(f2, shade = TRUE, seWithMean = TRUE, scale = 0)
par(mfrow = c(1, 1))
```

`seWithMean = TRUE` includes the uncertainty in the intercept, which is the
interval you want when reading a smooth as a partial effect. `scale = 0` lets
each panel set its own $y$ range, so a flat term looks flat rather than being
stretched to fill the plot.

## 3.3 Model fit, in context

```{r}
#| label: fit-summary

sm <- summary(f2)

tibble(
  quantity = c(
    "deviance explained",
    "adjusted R-squared",
    "-REML",
    "n",
    "total effective degrees of freedom"
  ),
  value = c(
    percent(sm$dev.expl, accuracy = 0.01),
    sprintf("%.3f", sm$r.sq),
    sprintf("%.1f", f2$gcv.ubre),
    comma(sm$n),
    sprintf("%.1f", sum(f2$edf))
  )
) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

::: {.callout-note appearance="simple"}

**Deviance explained of `r percent(sm$dev.expl, accuracy = 0.01)` is the
expected magnitude, not a problem.**
Completion is a genuinely noisy outcome and play-by-play information explains
little of it. Two consequences worth stating up front.

First, nflverse `cp` sees roughly this same information, so it should not be
dramatically better than stage 2 — and if it is, that difference is informative
about model class rather than about features.

Second, the unexplained variance is the room stages 3 and 4 have to work in. A
baseline that explained most of the outcome would leave nothing for coverage
geometry to claim. This number is the reason the project is worth doing.

:::

# 4. Out-of-fold predictions

Each stage is fit 17 times and every play is predicted once, by a fit that never
saw its week. The result is one row per (play, model).

Predictions are **pooled** rather than averaged across folds. With folds of
unequal size, the mean of per-fold log losses and the log loss of the pooled
predictions are different numbers, and only the pooled version can be subset
afterwards — to the `has_cp` rows, or to separation buckets — without refitting
anything.

```{r}
#| label: cv-predict

# Stages 1 and 2 here; stage 3 is cross-validated in §9.
preds <- cv_predict_all(mf_complete, folds, specs = MODEL_SPECS[1:2], verbose = FALSE)
pred_mean <- tapply(preds$.pred, preds$model, mean)
s1_range <- range(preds$.pred[preds$model == "1. intercept"])
s2_range <- range(preds$.pred[preds$model == "2. play-by-play"])

preds |>
  group_by(model) |>
  summarize(
    n = n(),
    min = min(.pred),
    p25 = quantile(.pred, 0.25),
    median = median(.pred),
    mean = mean(.pred),
    p75 = quantile(.pred, 0.75),
    max = max(.pred),
    .groups = "drop"
  ) |>
  gt() |>
  fmt_number(columns = -c(model, n), decimals = 3) |>
  fmt_number(columns = n, decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

Two things in that table are checks rather than description.

**The mean prediction should sit on the base rate.** It does for both stages:
`r sprintf("%.4f", pred_mean[["1. intercept"]])` and
`r sprintf("%.4f", pred_mean[["2. play-by-play"]])` against
`r sprintf("%.4f", mean(mf_complete$complete))`. A visible gap here would be the
first sign of a calibration problem, and it shows up before any metric is
computed.

**Stage 1's spread is the fold-to-fold variation in the base rate itself**, not a
model learning anything: each of its 17 distinct predictions is the completion
rate of the *other* 16 weeks. The range,
`r sprintf("%.3f", s1_range[1])` to `r sprintf("%.3f", s1_range[2])`, is therefore
a useful incidental measurement: the leave-one-week-out completion rate in 2018
varied by `r sprintf("%.1f", 100 * diff(s1_range))` percentage points.

```{r}
#| label: prediction-distributions
#| fig-height: 4

p_dist <- preds |>
  ggplot(aes(.pred, fill = model)) +
  geom_histogram(binwidth = 0.02, alpha = 0.75, position = "identity") +
  geom_vline(
    xintercept = mean(mf_complete$complete),
    linetype = "dashed",
    linewidth = 0.5
  ) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(
    title = "Out-of-fold predicted completion probability, by stage",
    subtitle = paste0(
      "Dashed line is the sample base rate (",
      percent(mean(mf_complete$complete), accuracy = 0.1),
      "). Stage 1 predicts it by construction."
    ),
    x = "predicted P(complete)",
    y = "plays",
    fill = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

ggsave(
  here("figs", "week06", "prediction_distributions.png"),
  p_dist,
  width = 8,
  height = 4,
  dpi = 200
)

p_dist
```

A model that discriminates has to *spread out*: pushing predictions away from the
base rate is how it earns a lower log loss, and being wrong while spread out is
how it earns a worse one. Stage 2 spans `r sprintf("%.2f", s2_range[1])` to
`r sprintf("%.2f", s2_range[2])`. Whether that spread is justified is the
question §5–§7 answer.

# 5. Scoring

## 5.1 Log loss

The metric is the mean negative log-likelihood of the observed outcomes under the
out-of-fold predictions, in nats:

$$
\text{LL} = -\frac{1}{n}\sum_{i=1}^{n}\Big[y_i \log \hat p_i + (1 - y_i)\log(1 - \hat p_i)\Big]
$$

It is a proper scoring rule: it is minimized in expectation only by reporting the
true probability, so it rewards calibration and discrimination together rather
than one at the expense of the other. `pointwise_log_loss()` and `log_loss()` in
`R/evaluate.R` are written by hand rather than taken from yardstick, and they
assert $0 < \hat p < 1$ rather than clipping. The reasons are in that file.

**The reference point is the binary entropy of the base rate.** A constant
prediction at $\bar y$ scores

$$
H(\bar y) = -\big[\bar y \log \bar y + (1 - \bar y)\log(1 - \bar y)\big],
$$

and stage 1 cannot beat it out of fold. Each held-out week $w$ is predicted by
the completion rate of the other 16 weeks, which lies on the *opposite* side of
$\bar y$ from that week's own rate $r_w$. Log loss within a week is convex in the
prediction and minimized at $r_w$, so every week scores at least as badly as it
would at $\bar y$. A stage 1 value below $H(\bar y)$ would mean information is
leaking across folds.

Both expectations are asserted below, so a render fails rather than reporting a
broken harness.

```{r}
#| label: log-loss

ybar <- mean(mf_complete$complete)
h_ybar <- -(ybar * log(ybar) + (1 - ybar) * log1p(-ybar))

scores <- preds |>
  group_by(model) |>
  summarize(
    n = n(),
    log_loss = log_loss(complete, .pred),
    .groups = "drop"
  ) |>
  mutate(
    vs_entropy = log_loss - h_ybar,
    vs_intercept = log_loss - log_loss[model == "1. intercept"]
  )
ll <- set_names(scores$log_loss, scores$model)

stopifnot(
  scores$log_loss[scores$model == "1. intercept"] >= h_ybar,
  scores$log_loss[scores$model == "2. play-by-play"] <
    scores$log_loss[scores$model == "1. intercept"]
)

scores |>
  gt() |>
  tab_header(
    subtitle = sprintf(
      "Out-of-fold, n = %s. H(base rate) = %.4f nats.",
      comma(nrow(mf_complete)),
      h_ybar
    ),
    title = "Log loss by stage"
  ) |>
  fmt_number(columns = c(log_loss, vs_entropy, vs_intercept), decimals = 4) |>
  fmt_number(columns = n, decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

**Stage 1 sits `r sprintf("%.5f", ll[["1. intercept"]] - h_ybar)` nats above the
floor**, which is the size the argument above predicts: the 17 fold means span
only `r sprintf("%.3f", s1_range[1])` to `r sprintf("%.3f", s1_range[2])` (§4),
so the penalty for predicting each week with the other weeks' rate is tiny but
cannot be negative.

**Stage 2 improves on it by
`r sprintf("%.4f", ll[["1. intercept"]] - ll[["2. play-by-play"]])` nats.** As a
share of the stage 1 loss that is
`r percent(1 - ll[["2. play-by-play"]] / ll[["1. intercept"]], accuracy = 0.1)`,
against `r percent(sm$dev.expl, accuracy = 0.01)` deviance explained in-sample
(§3.3). The out-of-fold figure being only slightly smaller is the expected
direction, and its small size says the stage 2 smooths are not overfitting at
this sample size.

This is the number stages 3 and 4 have to improve on. Whether a difference of a
given size is distinguishable from fold-to-fold noise is a question for the
paired interval, which follows.

# 6. Paired $\Delta$ log loss

## 6.1 The estimator

Two stages are compared on the same plays, so the comparison is paired. For each
play,

$$
d_i = \ell_i^{(b)} - \ell_i^{(a)}, \qquad
\Delta = \frac{1}{n}\sum_{i=1}^{n} d_i = \text{LL}^{(b)} - \text{LL}^{(a)},
$$

where $\ell_i$ is the per-play log loss of §5.1. Negative $\Delta$ favours stage
$b$. Pairing removes the variation in difficulty that both stages share, so the
interval on $\Delta$ is far narrower than intervals on the two log losses
separately would suggest.

Plays are not independent, so the interval resamples **clusters** rather than
plays. With clusters $g = 1, \dots, G$, cluster sums $S_g = \sum_{i \in g} d_i$
and sizes $n_g$, each bootstrap draw samples $G$ clusters with replacement and
computes

$$
\Delta^{*} = \frac{\sum_{g \in \text{draw}} S_g}{\sum_{g \in \text{draw}} n_g},
$$

and the 95% interval is the 2.5th and 97.5th percentiles of $10{,}000$ draws.
The interval holds the out-of-fold predictions fixed: it measures how much
$\Delta$ would vary across samples of plays, not across refits.

Two clusterings are reported. **Game** (253 clusters) is the primary choice:
plays within a game share a quarterback, a defense, and conditions. **Week** (17
clusters) matches the cross-validation folds, but 17 is too few for a percentile
bootstrap to be reliable, so it is a sensitivity check. As cross-checks,
`se_cluster` is the analytic cluster-robust standard error,

$$
\widehat{\text{SE}} = \frac{1}{n}\sqrt{\frac{G}{G-1}\sum_{g}\big(S_g - n_g\,\Delta\big)^2},
$$

which should agree with the bootstrap standard error, and `se_iid` ignores
clustering altogether.

## 6.2 Stage 2 against stage 1

```{r}
#| label: paired-delta

deltas <- bind_rows(
  paired_delta(preds, "1. intercept", "2. play-by-play", cluster = "game_id"),
  paired_delta(preds, "1. intercept", "2. play-by-play", cluster = "week")
)

# The paired mean is exactly the difference of the §5.1 log losses.
stopifnot(isTRUE(all.equal(
  deltas$estimate,
  rep(diff(scores$log_loss), 2)
)))

deltas |>
  mutate(design_effect = (se_cluster / se_iid)^2) |>
  select(
    cluster,
    n_clusters,
    estimate,
    conf_low,
    conf_high,
    se_boot,
    se_cluster,
    se_iid,
    design_effect
  ) |>
  gt() |>
  tab_header(
    title = "Stage 2 minus stage 1, out-of-fold log loss",
    subtitle = "Cluster bootstrap, 10,000 draws, 95% percentile interval"
  ) |>
  fmt_number(columns = c(estimate, conf_low, conf_high), decimals = 4) |>
  fmt_number(columns = c(se_boot, se_cluster, se_iid), decimals = 5) |>
  fmt_number(columns = design_effect, decimals = 2) |>
  gt_theme_538(quiet = TRUE)

dg <- deltas[deltas$cluster == "game_id", ]
dw <- deltas[deltas$cluster == "week", ]
deff <- function(d) sprintf("%.2f", (d$se_cluster / d$se_iid)^2)
```

**Stage 2 improves on stage 1 by `r sprintf("%.4f", -dg$estimate)` nats, 95%
interval [`r sprintf("%.4f", dg$conf_low)`, `r sprintf("%.4f", dg$conf_high)`]
clustering by game.** The interval is about
`r round(-dg$estimate / dg$se_boot)` standard errors from zero, so the
baseline's improvement over the base rate is not in question.

**The bootstrap agrees with the analytic standard error** under game clustering
(`r sprintf("%.5f", dg$se_boot)` against `r sprintf("%.5f", dg$se_cluster)`),
which is the check that the resampling code is right.

**Clustering matters little for this comparison.** The design effect is
`r deff(dg)` for games: within-game correlation in the paired differences is
small, because game-level factors that make plays harder or easier affect
both stages' losses alike and cancel in $d_i$. The week figure (`r deff(dw)`) is
within the noise of a standard error estimated from 17 clusters, whose own
relative error is roughly $1/\sqrt{2(G-1)} \approx 18\%$; it does not indicate
that weeks behave differently from games.

This need not hold for stages 3 and 4. Coverage features are partly team-level,
so their improvements over stage 2 could be concentrated in particular games,
and game clustering is retained so the interval would widen if they are.

# 7. Calibration

## 7.1 Measures

Log loss rewards calibration and discrimination together. Calibration asks the
first question on its own: when the model says 70%, do 70% of those passes
complete? Three summary measures (Van Calster et al., 2019):

$$
\text{CITL} = \bar y - \bar{\hat p},
$$

$$
\operatorname{logit} P(y_i = 1) = a + \operatorname{logit}\hat p_i,
\qquad
\operatorname{logit} P(y_i = 1) = a' + b \operatorname{logit}\hat p_i .
$$

- **Calibration-in-the-large** (CITL) is the gap between the observed and
  predicted rates; positive means the model underpredicts.
- **The calibration intercept** $a$ is the same gap on the logit scale, with the
  slope fixed at 1.
- **The calibration slope** $b$ is 1 for a calibrated model. Below 1, the
  predictions are too extreme (too close to 0 and 1); above 1, too timid.

Intervals are game-cluster bootstrap percentiles, as in §6, holding the
predictions fixed and refitting the two recalibration models on each draw.

The curves show the same thing visually: a smoothed calibration curve (a
binomial GAM of the outcome on $\operatorname{logit}\hat p$, with a pointwise
95% band) and observed rates in 20 equal-count bins of the predictions, each
holding about 830 plays.

Stage 1 is omitted: its predictions span
`r sprintf("%.1f", 100 * diff(s1_range))` percentage points, so its slope is not
meaningful, and its CITL is zero by construction.

nflverse `cp` is calibrated alongside stage 2, as a fixed external set of
predictions. It is not refit or recalibrated (§8).

```{r}
#| label: add-cp

preds_all <- bind_rows(
  preds,
  mf_complete |>
    transmute(
      game_id,
      play_id,
      week,
      complete,
      .pred = cp,
      n_train = NA_integer_,
      model = "cp"
    )
)

calib_models <- c("2. play-by-play", "cp")
```

## 7.2 Overall

```{r}
#| label: calibration-summary

calib <- map(calib_models, \(m) {
  d <- filter(preds_all, model == m)
  calibration_summary(d$complete, d$.pred, cluster = d$game_id) |>
    mutate(model = m)
}) |>
  list_rbind()

# Adding 0 turns a rounded -0 into 0, so it prints without a sign.
cs <- function(m, stat, col = "estimate", digits = 3) {
  x <- calib[[col]][calib$model == m & calib$stat == stat]
  sprintf(paste0("%.", digits, "f"), round(x, digits) + 0)
}

calib |>
  select(model, stat, estimate, conf_low, conf_high) |>
  gt(groupname_col = "model") |>
  tab_header(
    title = "Calibration, out-of-fold",
    subtitle = "Game-cluster bootstrap, 2,000 draws, 95% percentile interval"
  ) |>
  fmt_number(columns = c(estimate, conf_low, conf_high), decimals = 3) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: calibration-curves
#| fig-height: 5

curves <- map(calib_models, \(m) {
  d <- filter(preds_all, model == m)
  cc <- calibration_curve(d$complete, d$.pred)
  list(
    bins = mutate(cc$bins, model = m),
    smooth = mutate(cc$smooth, model = m)
  )
})
cal_bins <- list_rbind(map(curves, "bins"))
cal_smooth <- list_rbind(map(curves, "smooth"))

p_cal <- ggplot() +
  geom_abline(linetype = "dashed", linewidth = 0.4, colour = "grey50") +
  geom_ribbon(
    data = cal_smooth,
    aes(pred, ymin = low, ymax = high),
    fill = "grey70",
    alpha = 0.5
  ) +
  geom_line(data = cal_smooth, aes(pred, obs), linewidth = 0.6) +
  geom_pointrange(
    data = cal_bins,
    aes(
      mean_pred,
      obs_rate,
      ymin = obs_rate - 1.96 * se,
      ymax = obs_rate + 1.96 * se
    ),
    size = 0.15,
    colour = "#1f4fd8"
  ) +
  facet_wrap(~model) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Calibration of out-of-fold predictions",
    subtitle = "Smoothed curve with 95% band; points are 20 equal-count bins",
    x = "predicted P(complete)",
    y = "observed completion rate"
  ) +
  theme_minimal()

ggsave(
  here("figs", "week06", "calibration.png"),
  p_cal,
  width = 8,
  height = 4.5,
  dpi = 200
)

p_cal
```

**Stage 2 is well calibrated overall.** CITL is
`r cs("2. play-by-play", "citl")`
[`r cs("2. play-by-play", "citl", "conf_low")`,
`r cs("2. play-by-play", "citl", "conf_high")`] and the slope
`r cs("2. play-by-play", "slope")`
[`r cs("2. play-by-play", "slope", "conf_low")`,
`r cs("2. play-by-play", "slope", "conf_high")`]. A slope this close to 1 out of
fold says the spread of stage 2's predictions is earned rather than overfit.

**`cp` is close to calibrated but not quite, on this population**: CITL
`r cs("cp", "citl")` [`r cs("cp", "citl", "conf_low")`,
`r cs("cp", "citl", "conf_high")`] and slope `r cs("cp", "slope")`
[`r cs("cp", "slope", "conf_low")`, `r cs("cp", "slope", "conf_high")`]. On
throws beyond the line of scrimmage it slightly underpredicts completion, and
its predictions are slightly too timid. `cp` was fit to all passes, including
the screens and other short throws removed here, so some miscalibration on a
subpopulation is expected.

## 7.3 Within depth buckets

Overall calibration can hide offsetting errors in subgroups. Depth is the
natural first split, since it is the strongest predictor and since tracking
features are reported against it. Bucket edges on
$\text{depth}_{\text{arr}}$ are fixed in advance (`depth_bucket()`): short
(0–10 yd), intermediate (10–20), and deep (20+).

```{r}
#| label: calibration-buckets

with_bucket <- preds_all |>
  filter(model %in% calib_models) |>
  left_join(
    select(mf_complete, game_id, play_id, depth_arr),
    by = c("game_id", "play_id")
  ) |>
  mutate(bucket = droplevels(depth_bucket(depth_arr)))

stopifnot(!anyNA(with_bucket$bucket))

by_bucket <- with_bucket |>
  group_by(model, bucket) |>
  group_modify(\(d, k) {
    calibration_summary(d$complete, d$.pred, cluster = d$game_id, B = 1000) |>
      mutate(observed = mean(d$complete), predicted = mean(d$.pred))
  }) |>
  ungroup()

ci_txt <- function(est, lo, hi) sprintf("%.3f [%.3f, %.3f]", est, lo, hi)

by_bucket |>
  filter(stat != "intercept") |>
  mutate(value = ci_txt(estimate, conf_low, conf_high)) |>
  select(model, bucket, n, observed, predicted, stat, value) |>
  pivot_wider(names_from = stat, values_from = value) |>
  gt(groupname_col = "model") |>
  tab_header(
    title = "Calibration by depth bucket",
    subtitle = "Game-cluster bootstrap, 1,000 draws, 95% percentile interval"
  ) |>
  fmt_number(columns = n, decimals = 0) |>
  fmt_percent(columns = c(observed, predicted), decimals = 1) |>
  gt_theme_538(quiet = TRUE)

bk <- function(m, b, stat, col = "estimate") {
  by_bucket[[col]][
    by_bucket$model == m & by_bucket$bucket == b & by_bucket$stat == stat
  ]
}
```

```{r}
#| label: calibration-bucket-curves
#| fig-height: 5.5

bucket_bins <- with_bucket |>
  group_by(model, bucket) |>
  mutate(bin = ntile(.pred, 10)) |>
  group_by(model, bucket, bin) |>
  summarize(
    n = n(),
    mean_pred = mean(.pred),
    obs_rate = mean(complete),
    .groups = "drop"
  ) |>
  mutate(se = sqrt(obs_rate * (1 - obs_rate) / n))

p_cal_bucket <- ggplot(bucket_bins, aes(mean_pred, obs_rate)) +
  geom_abline(linetype = "dashed", linewidth = 0.4, colour = "grey50") +
  geom_pointrange(
    aes(ymin = obs_rate - 1.96 * se, ymax = obs_rate + 1.96 * se),
    size = 0.15,
    colour = "#1f4fd8"
  ) +
  facet_grid(model ~ bucket) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Calibration within depth buckets",
    subtitle = "10 equal-count bins per bucket",
    x = "predicted P(complete)",
    y = "observed completion rate"
  ) +
  theme_minimal()

ggsave(
  here("figs", "week06", "calibration_by_depth.png"),
  p_cal_bucket,
  width = 9,
  height = 5.5,
  dpi = 200
)

p_cal_bucket
```

**Stage 2's calibration-in-the-large holds in every bucket**: each CITL
interval contains zero. `cp` underpredicts on short throws
(`r sprintf("%.3f", bk("cp", "0-10", "citl"))`
[`r sprintf("%.3f", bk("cp", "0-10", "citl", "conf_low"))`,
`r sprintf("%.3f", bk("cp", "0-10", "citl", "conf_high"))`]).

**Stage 2's slopes are within sampling error of 1 except at intermediate
depth**, where its predictions are slightly too timid
(`r sprintf("%.3f", bk("2. play-by-play", "10-20", "slope"))`
[`r sprintf("%.3f", bk("2. play-by-play", "10-20", "slope", "conf_low"))`,
`r sprintf("%.3f", bk("2. play-by-play", "10-20", "slope", "conf_high"))`]). On
deep throws the slope is
`r sprintf("%.2f", bk("2. play-by-play", "20+", "slope"))`
[`r sprintf("%.2f", bk("2. play-by-play", "20+", "slope", "conf_low"))`,
`r sprintf("%.2f", bk("2. play-by-play", "20+", "slope", "conf_high"))`]; on
the earlier specification, with charted air yards, a constant QB-hit effect, and
throws behind the line included, it was $0.62$ [$0.41$, $0.84$]
(`notes/decisions.md` §9.6). `cp` is too timid on deep throws
(`r sprintf("%.2f", bk("cp", "20+", "slope"))`
[`r sprintf("%.2f", bk("cp", "20+", "slope", "conf_low"))`,
`r sprintf("%.2f", bk("cp", "20+", "slope", "conf_high"))`]).

The stage 3–4 comparison is reported by depth as well as overall.

# 8. The `cp` benchmark

nflverse `cp` is scored as a fixed external set of predictions on the same rows
as the stages. It is not a clean out-of-sample rival: nflfastR's model was
trained on seasons including 2018, uses charted air yards, and is a gradient
boosted model trained on many seasons (`notes/decisions.md` §8.6). The project's
claim is that tracking geometry adds information beyond play-by-play — measured
by stages 3 and 4 against stage 2 — not that the project beats `cp`. The
benchmark places stage 2 relative to the field standard.

```{r}
#| label: cp-benchmark

ll_all <- preds_all |>
  group_by(model) |>
  summarize(n = n(), log_loss = log_loss(complete, .pred), .groups = "drop")

cp_deltas <- bind_rows(
  paired_delta(preds_all, "cp", "2. play-by-play", cluster = "game_id"),
  paired_delta(preds_all, "cp", "2. play-by-play", cluster = "week"),
  paired_delta(preds_all, "1. intercept", "cp", cluster = "game_id")
)

ll_all |>
  gt() |>
  tab_header(title = "Out-of-fold log loss, stages and benchmark") |>
  fmt_number(columns = log_loss, decimals = 4) |>
  fmt_number(columns = n, decimals = 0) |>
  gt_theme_538(quiet = TRUE)

cp_deltas |>
  select(reference, model, cluster, estimate, conf_low, conf_high, se_boot) |>
  gt() |>
  tab_header(
    title = "Paired differences in log loss (model minus reference)",
    subtitle = "Cluster bootstrap, 10,000 draws, 95% percentile interval"
  ) |>
  fmt_number(columns = c(estimate, conf_low, conf_high), decimals = 4) |>
  fmt_number(columns = se_boot, decimals = 5) |>
  gt_theme_538(quiet = TRUE)

d2cp <- cp_deltas[
  cp_deltas$model == "2. play-by-play" &
    cp_deltas$cluster == "game_id",
]
dcp1 <- cp_deltas[cp_deltas$model == "cp", ]
```

**`cp` beats stage 2 by `r sprintf("%.4f", d2cp$estimate)` nats**, 95% interval
[`r sprintf("%.4f", d2cp$conf_low)`, `r sprintf("%.4f", d2cp$conf_high)`] by
game. Against the intercept, `cp` gains `r sprintf("%.4f", -dcp1$estimate)`
nats and stage 2 `r sprintf("%.4f", -dg$estimate)`, so stage 2 captures
`r percent(dg$estimate / dcp1$estimate, accuracy = 1)` of `cp`'s improvement
over the base rate.

`cp` leads despite being slightly miscalibrated on this population (§7.2), so
the gap is in discrimination: `cp` ranks plays better. Several plausible
sources, none separable here: its model class can represent interactions that
stage 2's additive smooths cannot, it is trained on far more plays, and 2018 is
partly in its training data.

The gap is the reference point for stages 3 and 4. If coverage geometry closes
it, tracking matches the field standard with a simpler model class; if
geometry's improvement over stage 2 exceeds it, tracking adds more than the
field-standard play-by-play model does.

# 9. Stage 3: separation at the throw

## 9.1 The features

Both are measured at the throw frame, from `scripts/07_build_features.R`. With
$\mathbf{x}_r$ the targeted receiver's position and $D$ the tracked defenders,
the nearest defender is
$j^* = \arg\min_{j \in D} \lVert \mathbf{x}_j - \mathbf{x}_r \rVert$ and

$$
\text{sep} = \lVert \mathbf{x}_{j^*} - \mathbf{x}_r \rVert,
\qquad
\dot{\text{sep}} =
\frac{(\mathbf{v}_{j^*} - \mathbf{v}_r) \cdot (\mathbf{x}_{j^*} - \mathbf{x}_r)}{\text{sep}},
$$

where velocities come from speed and direction (`velocity_xy()`). Closing speed
is negative when the defender is gaining on the receiver. It is computed from
the instantaneous velocities rather than by differencing positions, so it is
defined at a single frame.

Design choices:

- **At the throw**, because the question is the completion probability at the
  moment of release. Separation at arrival is partly the catch itself.
- **The nearest defender**, not an assigned one: assignment needs coverage
  labels, which exist for week 1 only (`notes/decisions.md` §3).
- **Smooths** in both, since separation's effect should flatten once a receiver
  is clearly open.
- **A separation-by-depth interaction**, `ti(sep_throw, depth_arr)`. On a deep
  throw the ball is in the air long enough for separation at the release to
  change before it arrives, so separation should matter less. Without the
  interaction, stage 3 applied separation's full-sample effect to deep throws
  and was too extreme there (calibration slope 0.73 [0.59, 0.86] on 20+ yd;
  `notes/decisions.md` §9.8).

Stage 3 is stage 2 plus `s(sep_throw)`, `s(closing_throw)`, and
`ti(sep_throw, depth_arr)`, so the comparison is nested: any improvement is
credited to separation alone.

```{r}
#| label: sep-descriptive

mf_complete |>
  mutate(
    separation = cut(
      sep_throw,
      c(0, 1, 2, 3, 5, 8, Inf),
      labels = c("0-1", "1-2", "2-3", "3-5", "5-8", "8+")
    )
  ) |>
  group_by(separation) |>
  summarize(plays = n(), completion = mean(complete), .groups = "drop") |>
  gt() |>
  tab_header(title = "Completion rate by separation at the throw (yd)") |>
  fmt_number(columns = plays, decimals = 0) |>
  fmt_percent(columns = completion, decimals = 1) |>
  gt_theme_538(quiet = TRUE)
```

## 9.2 Stage 3, interpreted

```{r}
#| label: stage3-fit

f3 <- fit_full(mf_complete, MODEL_SPECS[["3. + separation"]])
st3 <- summary(f3)$s.table
sep_rows <- c("s(sep_throw)", "s(closing_throw)", "ti(sep_throw,depth_arr)")

st3[sep_rows, , drop = FALSE] |>
  as.data.frame() |>
  rownames_to_column("smooth") |>
  rename(p = `p-value`) |>
  gt() |>
  fmt_number(columns = c(edf, Ref.df, Chi.sq), decimals = 2) |>
  fmt_scientific(columns = p, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: stage3-smooths
#| fig-height: 4

par(mfrow = c(1, 3))
for (i in match(sep_rows, rownames(st3))) {
  plot(f3, select = i, shade = TRUE, seWithMean = TRUE, scale = 0)
}
par(mfrow = c(1, 1))
```

Deviance explained in-sample rises from
`r percent(summary(f2)$dev.expl, accuracy = 0.01)` (stage 2) to
`r percent(summary(f3)$dev.expl, accuracy = 0.01)` (stage 3).

## 9.3 Out of fold

```{r}
#| label: stage3-cv

preds3 <- cv_predict(
  mf_complete,
  folds,
  MODEL_SPECS[["3. + separation"]],
  "3. + separation",
  verbose = FALSE
)
preds_s3 <- bind_rows(preds_all, preds3)

ll3 <- preds_s3 |>
  group_by(model) |>
  summarize(n = n(), log_loss = log_loss(complete, .pred), .groups = "drop") |>
  arrange(desc(log_loss))

ll3 |>
  gt() |>
  tab_header(title = "Out-of-fold log loss") |>
  fmt_number(columns = log_loss, decimals = 4) |>
  fmt_number(columns = n, decimals = 0) |>
  gt_theme_538(quiet = TRUE)

d3 <- bind_rows(
  paired_delta(preds_s3, "2. play-by-play", "3. + separation", "game_id"),
  paired_delta(preds_s3, "2. play-by-play", "3. + separation", "week"),
  paired_delta(preds_s3, "cp", "3. + separation", "game_id")
)

d3 |>
  mutate(design_effect = (se_cluster / se_iid)^2) |>
  select(reference, model, cluster, estimate, conf_low, conf_high, se_boot, design_effect) |>
  gt() |>
  tab_header(
    title = "Paired differences in log loss (model minus reference)",
    subtitle = "Cluster bootstrap, 10,000 draws, 95% percentile interval"
  ) |>
  fmt_number(columns = c(estimate, conf_low, conf_high), decimals = 4) |>
  fmt_number(columns = se_boot, decimals = 5) |>
  fmt_number(columns = design_effect, decimals = 2) |>
  gt_theme_538(quiet = TRUE)

d32 <- d3[d3$reference == "2. play-by-play" & d3$cluster == "game_id", ]
d3cp <- d3[d3$reference == "cp", ]
```

**Separation improves on stage 2 by `r sprintf("%.4f", -d32$estimate)` nats**,
95% interval [`r sprintf("%.4f", d32$conf_low)`,
`r sprintf("%.4f", d32$conf_high)`] by game. For scale, stage 2's gain over the
base rate is `r sprintf("%.4f", -dg$estimate)` nats and `cp`'s lead over stage 2
was `r sprintf("%.4f", d2cp$estimate)` (§8). Against `cp`, stage 3 differs by
`r sprintf("%+.4f", d3cp$estimate)` [`r sprintf("%+.4f", d3cp$conf_low)`,
`r sprintf("%+.4f", d3cp$conf_high)`].

## 9.4 By depth

```{r}
#| label: stage3-by-depth

depth_of <- select(mf_complete, game_id, play_id, depth_arr)

d3_depth <- preds_s3 |>
  left_join(depth_of, by = c("game_id", "play_id")) |>
  mutate(bucket = droplevels(depth_bucket(depth_arr))) |>
  group_split(bucket) |>
  map(\(d) {
    paired_delta(d, "2. play-by-play", "3. + separation", "game_id") |>
      mutate(
        bucket = as.character(d$bucket[1]),
        stage2_log_loss = log_loss(
          d$complete[d$model == "2. play-by-play"],
          d$.pred[d$model == "2. play-by-play"]
        )
      )
  }) |>
  list_rbind()

d3_depth |>
  mutate(share_of_stage2 = -estimate / stage2_log_loss) |>
  select(bucket, n, stage2_log_loss, estimate, conf_low, conf_high, share_of_stage2) |>
  gt() |>
  tab_header(
    title = "Stage 3 minus stage 2, by depth (yd)",
    subtitle = "Game-cluster bootstrap, 10,000 draws"
  ) |>
  fmt_number(columns = n, decimals = 0) |>
  fmt_number(columns = c(stage2_log_loss, estimate, conf_low, conf_high), decimals = 4) |>
  fmt_percent(columns = share_of_stage2, decimals = 1) |>
  gt_theme_538(quiet = TRUE)

db <- function(b, col = "estimate") d3_depth[[col]][d3_depth$bucket == b]
```

**Separation improves on stage 2 at every depth, least on deep throws.** The
gain is `r sprintf("%.4f", -db("0-10"))` nats on short throws and
`r sprintf("%.4f", -db("10-20"))` on intermediate ones, about
`r percent(-db("0-10") / db("0-10", "stage2_log_loss"), accuracy = 1)` and
`r percent(-db("10-20") / db("10-20", "stage2_log_loss"), accuracy = 1)` of
stage 2's loss. On throws of 20+ yards it is
`r sprintf("%.4f", -db("20+"))` [`r sprintf("%.4f", -db("20+", "conf_high"))`,
`r sprintf("%.4f", -db("20+", "conf_low"))`], about
`r percent(-db("20+") / db("20+", "stage2_log_loss"), accuracy = 1)`. On a
deep throw the ball is in the air long enough for separation at the release to
change substantially before it arrives, so the separation at the throw says less
about the separation at the catch. That is the gap stage 4's time-based
features (time to arrival, the passing window) are designed to address.

## 9.5 Calibration

```{r}
#| label: stage3-calibration

d_s3 <- filter(preds_s3, model == "3. + separation") |>
  left_join(depth_of, by = c("game_id", "play_id")) |>
  mutate(bucket = droplevels(depth_bucket(depth_arr)))

cal3 <- bind_rows(
  calibration_summary(d_s3$complete, d_s3$.pred, d_s3$game_id) |>
    mutate(bucket = "all"),
  d_s3 |>
    group_split(bucket) |>
    map(\(d) {
      calibration_summary(d$complete, d$.pred, d$game_id, B = 1000) |>
        mutate(bucket = as.character(d$bucket[1]))
    }) |>
    list_rbind()
)

cal3 |>
  filter(stat != "intercept") |>
  mutate(value = sprintf("%.3f [%.3f, %.3f]", estimate, conf_low, conf_high)) |>
  select(bucket, n, stat, value) |>
  pivot_wider(names_from = stat, values_from = value) |>
  gt() |>
  tab_header(
    title = "Stage 3 calibration",
    subtitle = "Game-cluster bootstrap; 2,000 draws overall, 1,000 per bucket"
  ) |>
  fmt_number(columns = n, decimals = 0) |>
  gt_theme_538(quiet = TRUE)

c3 <- function(b, stat, col = "estimate") {
  cal3[[col]][cal3$bucket == b & cal3$stat == stat]
}
```

**Stage 3 is calibrated overall** — CITL `r sprintf("%.3f", round(c3("all", "citl"), 3) + 0)`
[`r sprintf("%.3f", c3("all", "citl", "conf_low"))`,
`r sprintf("%.3f", c3("all", "citl", "conf_high"))`], slope
`r sprintf("%.3f", c3("all", "slope"))`
[`r sprintf("%.3f", c3("all", "slope", "conf_low"))`,
`r sprintf("%.3f", c3("all", "slope", "conf_high"))`] — so adding two flexible
smooths did not cost calibration out of fold, and the improvement in §9.3 is
earned rather than overfit.

**On deep throws** the slope is `r sprintf("%.2f", c3("20+", "slope"))`
[`r sprintf("%.2f", c3("20+", "slope", "conf_low"))`,
`r sprintf("%.2f", c3("20+", "slope", "conf_high"))`]. Without the
separation-by-depth interaction it was 0.73 [0.59, 0.86]: stage 3 applied
separation's full-sample effect to deep throws, where §9.4 shows it is weaker.
This is the same pattern as the QB-hit effect before its depth interaction
(§3.2), and the same remedy.

## 9.6 Summary

```{r}
#| label: stage3-summary

ll3 |>
  mutate(
    vs_intercept = log_loss - log_loss[model == "1. intercept"],
    vs_stage2 = log_loss - log_loss[model == "2. play-by-play"]
  ) |>
  gt() |>
  tab_header(
    title = "Out-of-fold log loss, stages 1-3 and cp",
    subtitle = sprintf(
      "%s throws beyond the line to a targeted receiver",
      comma(nrow(mf_complete))
    )
  ) |>
  fmt_number(columns = c(log_loss, vs_intercept, vs_stage2), decimals = 4) |>
  fmt_number(columns = n, decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

Separation at the throw — two features and their interaction with depth, added
to a calibrated play-by-play baseline — lowers out-of-fold log loss by
`r sprintf("%.4f", -d32$estimate)` nats
[`r sprintf("%.4f", -d32$conf_high)`, `r sprintf("%.4f", -d32$conf_low)`] and
moves the model from `r sprintf("%.4f", d2cp$estimate)` nats behind `cp` to
`r sprintf("%.4f", -d3cp$estimate)` ahead of it. These are provisional numbers:
the model frame is assembled in this notebook rather than by `08`, and stage 4
is still to come.
