---
title: "Completion probability — baseline models"
subtitle: "STAT 1961 Capstone — preliminary results"
author: "David Neumann"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: [brite, ../theme/capstone.scss]
    code-summary: "Show code"
    fig-width: 8
    fig-height: 5
    embed-resources: true
execute:
  warning: false
  message: false
---

# 0. Setup

```{r}
#| label: setup
#| include: false
if (Sys.getenv("RENV_PROJECT") == "") {
  root <- normalizePath("..", mustWork = TRUE)
  Sys.setenv(RENV_PROJECT = root)
  setwd(root)
  source("renv/activate.R")
}

library(tidyverse)
library(arrow)
library(here)
library(gt)
library(gtExtras)
library(scales)

# Attached after tidyverse deliberately: nlme (pulled in by mgcv) masks
# dplyr::collapse(), which is unused here. mgcv must be ATTACHED rather
# than merely installed, because the MODEL_SPECS formulas are written
# with bare s() and the formula environment has to resolve it.
library(mgcv)

source(here("R", "constants.R"))
source(here("R", "utils.R"))
source(here("R", "evaluate.R"))

fig_dir <- here("figs", "week06")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
```

## 0.1 What this notebook is, and is not

This notebook reports. It reads persisted artifacts and writes nothing to
`data/`, the same contract as `analysis/01_eda.qmd` and
`analysis/02_arrival_anchor.qmd`. Figures are written to `figs/week06/` and
referenced statically from the slide deck.

Two things about it are **provisional**, and both belong on the slide rather
than in a footnote.

::: {.callout-important appearance="simple"}

**The model frame is assembled here, in §1.1.** That assembly is
`scripts/08_build_model_frame.R`'s job and that script does not exist yet.
Nothing is written to disk, so the architectural rule holds, but these results
are not yet reproducible from a build artifact.

**The population filters for this question are not applied.** Beyond the line
of scrimmage, a tracked targeted receiver, not a spike, not a scramble — none
of them. These are results on the **full throw population**, every play in
`analytic_sample.parquet` with `pass_result` in {C, I, IN}.

The beyond-the-LOS filter alone will remove roughly a fifth of these plays
(`notes/decisions.md` §7), which changes every number below. Stage 3 and 4
features do not exist yet either, so what follows is the **floor and the
baseline**, plus the external benchmark. The thesis is untested.

:::

```{r}
#| label: load_data

analytic_sample <- read_parquet(here(
  "data",
  "processed",
  "analytic_sample.parquet"
))
plays <- read_parquet(here("data", "processed", "plays.parquet"))
pbp <- read_parquet(here("data", "processed", "pbp.parquet"))
```

# 1. The model frame

## 1.1 Assembly

Three sources, joined on `(game_id, play_id)`. The join takes from each table
only the columns the other two lack, because `analytic_sample.parquet` already
carries `down`, `yards_to_go`, `los_x`, and `week` — pulling them again would
produce suffixed `.x` / `.y` pairs rather than an error.

- `analytic_sample.parquet` — the spine, and the response via `pass_result`
- `plays.parquet` — `number_of_pass_rushers`, which `04_build_sample.R` did not
  carry forward because the sample rule did not need it
- `pbp.parquet` — `air_yards`, `pass_location`, `qb_hit`, `shotgun`,
  `posteam_type`, and the benchmark columns `cp` / `has_cp`

`plays.parquet` is not widened by re-running `04`: the analytical sample is a
locked artifact of the sample layer, and adding model covariates to it would
conflate the sample layer with the feature layer.

```{r}
#| label: assemble

model_input <- analytic_sample |>
  left_join(
    plays |> select(game_id, play_id, number_of_pass_rushers),
    by = c("game_id", "play_id")
  ) |>
  left_join(pbp, by = c("game_id", "play_id")) |>
  filter(pass_result %in% c("C", "I", "IN"))

tibble(
  quantity = c(
    "plays in analytic_sample",
    "thrown passes (C, I, IN)",
    "removed: sacks and scrambles",
    "matched to play-by-play",
    "carrying the cp benchmark"
  ),
  n = c(
    nrow(analytic_sample),
    nrow(model_input),
    nrow(analytic_sample) - nrow(model_input),
    sum(model_input$pbp_matched),
    sum(model_input$has_cp)
  )
) |>
  mutate(pct_of_throws = n / nrow(model_input)) |>
  gt() |>
  fmt_percent(columns = pct_of_throws, decimals = 1) |>
  fmt_number(columns = n, decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

Sacks are removed because they are not throws. `keep_any_throw` in the sample
rule defines the population as plays where a forward pass was *released*
(`notes/decisions.md` §4.6), and a sack carries a throw event on a small number
of plays; this filter is on the outcome code rather than the event, so it
catches those.

## 1.2 Derived predictors

`prepare_model_frame()` in `R/evaluate.R` owns the response and three derived
columns, so that every stage and the benchmark see identical inputs. The
reasoning for each is in that file's header; the observable effects:

```{r}
#| label: prepare

mf <- prepare_model_frame(model_input)

tibble(
  transformation = c(
    "response: complete = 1 for C, 0 for I and IN",
    "number_of_pass_rushers clamped to [2, 7]",
    "dist_to_sticks = air_yards - yards_to_go",
    "air_yards_zero: plays at exactly 0 air yards",
    "home = posteam_type == 'home'"
  ),
  affected = c(
    sum(mf$complete),
    sum(
      model_input$number_of_pass_rushers != mf$number_of_pass_rushers,
      na.rm = TRUE
    ),
    sum(!is.na(mf$dist_to_sticks)),
    sum(mf$air_yards_zero, na.rm = TRUE),
    sum(mf$home, na.rm = TRUE)
  )
) |>
  gt() |>
  fmt_number(columns = affected, decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

An interception counts as an incompletion. This is how nflverse `cp` is defined,
and matching it is what makes the benchmark comparison meaningful rather than a
comparison of two different response variables.

## 1.3 The scoring population

Every stage's log loss has to be computed on the same rows, or the *differences*
between stages are not differences. `glm()` and `mgcv::gam()` default to
`na.action = na.omit`, so a stage referencing a column with missingness would
quietly fit and score fewer plays than the stage below it — no warning, no
error, and a delta computed across two different datasets.

`assert_complete()` therefore **asserts rather than filters**. Dropping rows is
an exclusion, exclusions are checked for differential completion rates and
logged with counts (`notes/decisions.md` §4.7), and the modeling layer is not
permitted to change the population silently.

```{r}
#| label: required-vars

tibble(variable = model_vars()) |>
  gt() |>
  tab_header(title = "Columns required by MODEL_SPECS") |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: incomplete-columns

rep <- incomplete_report(mf)

rep$by_column |>
  gt() |>
  fmt_percent(columns = pct_missing, decimals = 4) |>
  gt_theme_538(quiet = TRUE)
```

Three plays, and the seven rows above are not seven independent gaps: the four
`air_yards`-derived columns fail together on the same 3 plays, and the single
missing `shotgun` / `home` / `qb_hit` is one of those 3 — the one play that
found no play-by-play partner (`notes/decisions.md` §8.2).

```{r}
#| label: incomplete-composition

rep$composition |>
  gt() |>
  fmt_percent(columns = share, decimals = 1) |>
  gt_theme_538(quiet = TRUE)
```

::: {.callout-note appearance="simple"}

**Checked and immaterial.** The dropped block is 1 completion and 2
incompletions. Three plays cannot be selected on anything, and the base rate
moves by two parts in a million. This is the routine version of the check that
rejected the `pass_arrived` anchor in `02_arrival_anchor.qmd` §1; it is run on
every exclusion regardless of expected size, because the expensive failures are
the ones that look routine.

:::

```{r}
#| label: scoring-population

vars <- model_vars()
mf_complete <- filter(mf, complete.cases(mf[vars]))

assert_complete(mf_complete)

tibble(
  quantity = c(
    "thrown passes",
    "complete cases across all specs",
    "dropped",
    "completions",
    "base rate"
  ),
  value = c(
    comma(nrow(mf)),
    comma(nrow(mf_complete)),
    comma(nrow(mf) - nrow(mf_complete)),
    comma(sum(mf_complete$complete)),
    sprintf("%.4f", mean(mf_complete$complete))
  )
) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

The base rate of $0.6529$ is the number stage 1 exists to reproduce, and the
number every later stage has to beat.

# 2. Resampling

## 2.1 Leave-one-week-out, not random $k$-fold

Plays are not independent. Two plays from the same game share a quarterback, a
defense, a game plan, the weather, and the officiating crew. Under a random
split, one of them can sit in the training set while the other is scored, and
the model predicts the second partly by having memorized conditions from the
first. The resulting estimate is optimistic and **carries no signature** — there
is nothing in the number that reveals the problem.

Grouping the folds by week makes every fold a clean block: fit on 16 weeks,
predict the 17th, rotate. It also mirrors how such a model would actually be
used, which is a better justification than a technical one. The general argument
is in Roberts et al. (2017), *Ecography*.

No seed is needed. With one group per fold there is no random assignment of
groups to folds, so the only thing that could vary is the returned order, and
`make_folds()` sorts.

```{r}
#| label: folds

folds <- make_folds(mf_complete)

tibble(
  held_out_week = folds$held_out,
  train_n = map_int(folds$splits, \(s) nrow(rsample::analysis(s))),
  test_n = map_int(folds$splits, \(s) nrow(rsample::assessment(s)))
) |>
  gt() |>
  fmt_number(columns = c(train_n, test_n), decimals = 0) |>
  grand_summary_rows(
    columns = test_n,
    fns = list(total = ~ sum(.)),
    fmt = ~ fmt_number(., decimals = 0)
  ) |>
  gt_theme_538(quiet = TRUE)
```

The `test_n` total equals the scoring population exactly, which is the property
that makes the pooled out-of-fold predictions in §4 a complete prediction for
every play: each one is predicted exactly once, by a fit that never saw its
week.

# 3. Stage 2, interpreted

Two different jobs, kept separate. The 17 cross-validation fits answer *how well
does this predict* and are discarded — their only product is the prediction
table in §4. A single fit on all the data answers *what did it learn*, and that
is the object below.

::: {.callout-warning appearance="simple"}

**Nothing in this section is out-of-sample.** No metric reported here is a
performance estimate. Section 5 onward is where performance is measured.

:::

```{r}
#| label: full-fit

f2 <- fit_full(mf_complete, MODEL_SPECS[["2. play-by-play"]])

summary(f2)
```

## 3.1 Reading the parametric terms

```{r}
#| label: parametric-terms

summary(f2)$p.table |>
  as.data.frame() |>
  rownames_to_column("term") |>
  as_tibble() |>
  rename(estimate = Estimate, se = `Std. Error`, p = `Pr(>|z|)`) |>
  select(term, estimate, se, p) |>
  arrange(p) |>
  gt() |>
  fmt_number(columns = c(estimate, se), decimals = 3) |>
  fmt_scientific(columns = p, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

Three of these are worth reporting as findings rather than as controls.

**`air_yards_zero` at $-0.87$ on the logit scale**, against an intercept of
$1.00$: a throw at exactly zero air yards has roughly half the odds of
completion of one nearby on either side. nflverse describes this indicator as a
relic of its pre-tree models, and for a tree it is — a tree can split at exactly
$0$. A penalized spline cannot; it smooths straight over a point mass of 1,093
plays. Including it raised the effective degrees of freedom on `s(air_yards)`
from $4.37$ to $7.67$ and improved deviance explained on a strictly smaller
model.

**`shotgun` at $-0.171$**, which is the opposite sign from the conventional
wisdom that passing statistics are inflated out of shotgun. Both can be true:
shotgun throws are systematically shorter, so the raw advantage is confounded by
depth, and conditioning on depth inverts it. This coefficient was *not*
estimable in the first fit, where `offense_formation` was also in the model — an
EMPTY-backfield snap is essentially always from shotgun, so the two were aliased
and the standard error on `shotgun` was $0.349$ rather than $0.047$.

**`qb_hit` at $-0.88$** is the largest effect among the parametric terms. This
vindicates treating it as a covariate rather than as an exclusion
(`notes/decisions.md` §7): it is a play-level flag that does not distinguish
"hit during the throw" from "hit afterwards," so filtering on it would remove
real passes and may itself select on outcome, while including it is informative
and cheap to defend.

## 3.2 Reading the smooths

`edf` is the effective degrees of freedom the penalty actually chose. An `edf`
near $1$ means the penalty shrank the smooth to a straight line — the term is in
the model but the relationship is linear.

```{r}
#| label: smooth-terms

summary(f2)$s.table |>
  as.data.frame() |>
  rownames_to_column("smooth") |>
  as_tibble() |>
  rename(p = `p-value`) |>
  gt() |>
  fmt_number(columns = c(edf, Ref.df, `Chi.sq`), decimals = 2) |>
  fmt_scientific(columns = p, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

`s(dist_to_sticks)` comes back essentially null: `edf` $1.01$, $p = 0.10$. The
motivation was sound — additive smooths in `air_yards` and `yards_to_go` cannot
represent their difference, which is an interaction, and a 3-yard throw on
3rd-and-2 is a different play from a 3-yard throw on 3rd-and-15 — but the data
does not support it. Conditional on depth, whether the throw reaches the line to
gain barely affects whether it is caught. In hindsight this is sensible: the
sticks govern *play calling*, which is upstream of the throw rather than a
property of it.

**It stays in the model anyway.** Stage 2 exists to be a strong baseline, and
pruning a baseline on in-sample significance is what makes a nested comparison
look tuned. It costs one effective degree of freedom out of 17,074 observations.
It also spans the `yards_to_go` direction, so dropping it would remove that
variable from the model entirely.

```{r}
#| label: gam-check

gam.check(f2)
```

The basis dimension check is the part to read. `k-index` below $1$ with a small
$p$-value suggests the basis is too small — *especially* when `edf` sits close to
`k'`. Here the largest `edf` is $7.67$ against `k'` of $9$, and the penalty is
choosing well short of the available flexibility everywhere else, so the default
basis is adequate.

Do not over-read the residual plots. Deviance residuals from a binomial fit on a
binary response always separate into two arcs; that is a property of binary data
rather than a diagnosis.

```{r}
#| label: smooth-plots
#| fig-height: 6

par(mfrow = c(2, 2))
plot(f2, shade = TRUE, seWithMean = TRUE, scale = 0)
par(mfrow = c(1, 1))
```

`seWithMean = TRUE` includes the uncertainty in the intercept, which is the
interval you want when reading a smooth as a partial effect. `scale = 0` lets
each panel set its own $y$ range, so a flat term looks flat rather than being
stretched to fill the plot.

## 3.3 Model fit, in context

```{r}
#| label: fit-summary

sm <- summary(f2)

tibble(
  quantity = c(
    "deviance explained",
    "adjusted R-squared",
    "-REML",
    "n",
    "total effective degrees of freedom"
  ),
  value = c(
    percent(sm$dev.expl, accuracy = 0.01),
    sprintf("%.3f", sm$r.sq),
    sprintf("%.1f", f2$gcv.ubre),
    comma(sm$n),
    sprintf("%.1f", sum(f2$edf))
  )
) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

::: {.callout-note appearance="simple"}

**Deviance explained of 9.85% is the expected magnitude, not a problem.**
Completion is a genuinely noisy outcome and play-by-play information explains
little of it. Two consequences worth stating up front.

First, nflverse `cp` sees roughly this same information, so it should not be
dramatically better than stage 2 — and if it is, that difference is informative
about model class rather than about features.

Second, the unexplained variance is the room stages 3 and 4 have to work in. A
baseline that explained most of the outcome would leave nothing for coverage
geometry to claim. This number is the reason the project is worth doing.

:::

# 4. Out-of-fold predictions

Each stage is fit 17 times and every play is predicted once, by a fit that never
saw its week. The result is one row per (play, model).

Predictions are **pooled** rather than averaged across folds. With folds of
unequal size, the mean of per-fold log losses and the log loss of the pooled
predictions are different numbers, and only the pooled version can be subset
afterwards — to the `has_cp` rows, or to separation buckets — without refitting
anything.

```{r}
#| label: cv-predict

preds <- cv_predict_all(mf_complete, folds, verbose = FALSE)

preds |>
  group_by(model) |>
  summarize(
    n = n(),
    min = min(.pred),
    p25 = quantile(.pred, 0.25),
    median = median(.pred),
    mean = mean(.pred),
    p75 = quantile(.pred, 0.75),
    max = max(.pred),
    .groups = "drop"
  ) |>
  gt() |>
  fmt_number(columns = -c(model, n), decimals = 3) |>
  fmt_number(columns = n, decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

Two things in that table are checks rather than description.

**The mean prediction should sit on the base rate.** It does, for both stages, to
four decimal places against $0.6529$. A visible gap here would be the first sign
of a calibration problem, and it shows up before any metric is computed.

**Stage 1's spread is the fold-to-fold variation in the base rate itself**, not a
model learning anything: each of its 17 distinct predictions is the completion
rate of the *other* 16 weeks. The range $0.650$ to $0.655$ is therefore a useful
incidental measurement — week-to-week completion rate in 2018 varied by about
half a percentage point around its mean.

```{r}
#| label: prediction-distributions
#| fig-height: 4

p_dist <- preds |>
  ggplot(aes(.pred, fill = model)) +
  geom_histogram(binwidth = 0.02, alpha = 0.75, position = "identity") +
  geom_vline(
    xintercept = mean(mf_complete$complete),
    linetype = "dashed",
    linewidth = 0.5
  ) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(
    title = "Out-of-fold predicted completion probability, by stage",
    subtitle = paste0(
      "Dashed line is the sample base rate (",
      percent(mean(mf_complete$complete), accuracy = 0.1),
      "). Stage 1 predicts it by construction."
    ),
    x = "predicted P(complete)",
    y = "plays",
    fill = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

ggsave(
  here("figs", "week06", "prediction_distributions.png"),
  p_dist,
  width = 8,
  height = 4,
  dpi = 200
)

p_dist
```

A model that discriminates has to *spread out*: pushing predictions away from the
base rate is how it earns a lower log loss, and being wrong while spread out is
how it earns a worse one. Stage 2 spans roughly $0.07$ to $0.92$. Whether that
spread is justified is the question §5 and §6 answer.

```{r}
#| label: observed-vs-predicted-band

# A coarse look ahead at calibration: within deciles of prediction, does
# the observed rate track? Section 6 does this properly, with intervals.
preds |>
  filter(model == "2. play-by-play") |>
  mutate(bin = ntile(.pred, 10)) |>
  group_by(bin) |>
  summarize(
    n = n(),
    mean_pred = mean(.pred),
    obs_rate = mean(complete),
    .groups = "drop"
  ) |>
  mutate(gap = obs_rate - mean_pred) |>
  gt() |>
  fmt_number(columns = c(mean_pred, obs_rate, gap), decimals = 3) |>
  gt_theme_538(quiet = TRUE)
```

Sections 5 through 8 — scoring, calibration, the `cp` benchmark, and $\Delta$ log
loss with a clustered interval — follow.
