---
title: "The arrival anchor at season scale"
subtitle: "STAT 1961 Capstone — validation of scripts/06_build_throw_frame.R"
author: "David Neumann"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: [brite, ../theme/capstone.scss]
    code-summary: "Show code"
    fig-width: 8
    fig-height: 5
    embed-resources: true
execute:
  warning: false
  message: false
---

# 0. Setup

```{r}
#| label: setup
#| include: false
if (Sys.getenv("RENV_PROJECT") == "") {
  root <- normalizePath("..", mustWork = TRUE)
  Sys.setenv(RENV_PROJECT = root)
  setwd(root)
  source("renv/activate.R")
}

library(tidyverse)
library(arrow)
library(here)
library(gt)
library(gtExtras)
library(scales)
library(patchwork)

source(here("R", "constants.R"))
source(here("R", "utils.R"))
source(here("R", "arrival.R"))
source(here("R", "viz.R"))

fig_dir <- here("figs", "week06")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
```

## 0.1 What this notebook is

`analysis/02_arrival_anchor.qmd` defined the arrival anchor on week 1 and
listed the checks to repeat on the full season (`notes/decisions.md` §6.8).
`scripts/06_build_throw_frame.R` now applies `detect_arrival()` to every play
with a tracked target. This notebook runs those checks on its output:

1. class similarity of $d_{\text{arr}}$;
2. speed equalization by lane length;
3. the bounce check;
4. three plays drawn frame by frame around arrival;
5. external validation of the tracking-derived depth against nflverse
   `air_yards`, which also informs the beyond-the-line-of-scrimmage threshold.

It also reruns the $\delta$ and gap-tolerance sensitivity checks at season
scale. It reads persisted artifacts and writes nothing to `data/`.

```{r}
#| label: load_data

arrival <- read_parquet(here("data", "processed", "arrival.parquet"))
approach <- read_parquet(here("data", "processed", "approach.parquet"))
analytic_sample <- read_parquet(here(
  "data",
  "processed",
  "analytic_sample.parquet"
))
pbp <- read_parquet(here("data", "processed", "pbp.parquet"))
trk <- open_dataset(here("data", "processed", "tracking"))

# arrival.parquet carries no outcome columns; pass_result is joined here.
arr <- arrival |>
  left_join(
    select(analytic_sample, game_id, play_id, pass_result, throw_anchor),
    by = c("game_id", "play_id")
  )
```

# 1. Coverage

```{r}
#| label: coverage

targets <- filter(analytic_sample, target_tracked)
no_arrival <- anti_join(targets, arrival, by = c("game_id", "play_id"))

tibble(
  quantity = c(
    "plays in analytic_sample",
    "with a tracked target",
    "with an arrival",
    "without an approach series",
    "closest-approach fallback (used_fallback)",
    "closest approach at the last searched frame (at_window_edge)",
    "arrival at the window cap (5.0 s)"
  ),
  n = c(
    nrow(analytic_sample),
    nrow(targets),
    nrow(arrival),
    nrow(no_arrival),
    sum(arrival$used_fallback),
    sum(arrival$at_window_edge),
    sum(arrival$f_arr == arrival$f_throw + MAX_FLIGHT_FRAMES)
  )
) |>
  mutate(
    pct_of_tracked_target = if_else(
      row_number() == 1,
      NA_real_,
      n / nrow(targets)
    )
  ) |>
  gt() |>
  fmt_number(columns = n, decimals = 0) |>
  fmt_percent(columns = pct_of_tracked_target, decimals = 2) |>
  sub_missing(missing_text = "") |>
  gt_theme_538(quiet = TRUE)

no_arrival |>
  select(
    game_id,
    play_id,
    week,
    pass_result,
    throw_anchor,
    f_throw,
    last_frame
  ) |>
  gt() |>
  tab_header(title = "Plays with a tracked target but no approach series") |>
  gt_theme_538(quiet = TRUE)
```

The plays without an approach series have no frame after the throw on which
both the ball and the targeted receiver are tracked: in one, tracking ends at
the throw frame; in the other two, the targeted receiver appears only on the
first two frames of the play. `target_tracked` records that the target appears
in the tracking data at all, not that they are tracked at the throw.

`at_window_edge` is common on completions and is not a failure: after a catch
the ball stays in the receiver's hands, so the distance can keep shrinking to
the last frame searched. It marks where the *closest approach* fell, not the
arrival frame, which is found before it.

# 2. Week 1 reproduces the prototype

The week-1 subset of `arrival.parquet` should match `02_arrival_anchor.qmd`
§7.1 exactly: 890 plays, and the same medians.

```{r}
#| label: week1-reproduction

arr |>
  filter(week == 1) |>
  group_by(pass_result) |>
  summarize(
    plays = n(),
    p50_d_arr = median(d_arr),
    p50_d_min = median(d_min),
    p50_to_min = median(frames_to_min),
    .groups = "drop"
  ) |>
  gt() |>
  tab_header(title = "Week 1, from scripts/06_build_throw_frame.R") |>
  fmt_number(columns = starts_with("p50_d"), decimals = 3) |>
  fmt_number(columns = p50_to_min, decimals = 1) |>
  gt_theme_538(quiet = TRUE)
```

`r sum(arr$week == 1)` plays, with medians matching the prototype's rendered
table (completion $d_{\text{arr}}$ 0.70, incompletion 1.16, completion
$d_{\min}$ 0.12).

# 3. Class similarity

```{r}
#| label: class-similarity

class_sim <- arr |>
  group_by(pass_result) |>
  summarize(
    plays = n(),
    p50_d_arr = median(d_arr),
    p90_d_arr = quantile(d_arr, 0.90),
    p50_d_min = median(d_min),
    p50_to_min = median(frames_to_min),
    fallback = mean(used_fallback),
    .groups = "drop"
  )

class_sim |>
  gt() |>
  tab_header(title = "All 17 weeks") |>
  fmt_number(columns = c(p50_d_arr, p90_d_arr, p50_d_min), decimals = 2) |>
  fmt_number(columns = p50_to_min, decimals = 1) |>
  fmt_percent(columns = fallback, decimals = 1) |>
  gt_theme_538(quiet = TRUE)

cls <- function(r, col) class_sim[[col]][class_sim$pass_result == r]
```

The season-scale medians stay close to week 1: $d_{\text{arr}}$ of
`r sprintf("%.2f", cls("C", "p50_d_arr"))` yd on completions and
`r sprintf("%.2f", cls("I", "p50_d_arr"))` on incompletions, against 0.70 and
1.16. As in the prototype, the residual difference is a football fact — a
caught ball ends up closer to the receiver than a dropped one — and the target
is similarity, not zero. Median $d_{\min}$ on completions,
`r sprintf("%.2f", cls("C", "p50_d_min"))` yd, confirms that the distance
series is right. The fallback rate is similar across classes.

# 4. Flight time

```{r}
#| label: flight-time

arr |>
  summarize(
    p01 = quantile(t_flight, 0.01),
    p50 = median(t_flight),
    p99 = quantile(t_flight, 0.99),
    max = max(t_flight),
    `under 0.4 s` = sum(t_flight < 0.4)
  ) |>
  pivot_longer(everything(), names_to = "statistic") |>
  gt() |>
  fmt_number(columns = value, decimals = 2) |>
  gt_theme_538(quiet = TRUE)

short <- filter(arr, t_flight < 0.4)
```

p99 of `r sprintf("%.1f", quantile(arr$t_flight, 0.99))` s supports the 5.0 s
window. Flights under 0.4 s number `r comma(nrow(short))`; their median depth
is `r sprintf("%.1f", median(short$depth_arr))` yd, so they are mostly throws
at or behind the line — shovels, screens, batted balls — that the
beyond-the-line filter removes.

# 5. Speed equalization

The decisive check. An artifact in arrival *timing* would show up as an
outcome-dependent implied ball speed, lane length divided by flight time,
within bins of lane length.

```{r}
#| label: speed-equalization

speed <- arr |>
  filter(pass_result %in% c("C", "I"), between(lane_len, 8, 30)) |>
  mutate(
    v_ball = lane_len / t_flight,
    bin = cut(lane_len, seq(8, 30, by = 2))
  ) |>
  group_by(bin, pass_result) |>
  summarize(
    n = n(),
    p50_lane = median(lane_len),
    p50_v = median(v_ball),
    .groups = "drop"
  )

speed_wide <- speed |>
  pivot_wider(names_from = pass_result, values_from = c(n, p50_lane, p50_v)) |>
  mutate(diff_v = p50_v_C - p50_v_I)

speed_wide |>
  gt() |>
  tab_header(title = "Median implied ball speed (yd/s) by lane length (yd)") |>
  fmt_number(columns = c(starts_with("p50"), diff_v), decimals = 2) |>
  fmt_number(columns = starts_with("n_"), decimals = 0) |>
  gt_theme_538(quiet = TRUE)

p_speed <- speed |>
  ggplot(aes(p50_lane, p50_v, colour = pass_result)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(limits = c(0, NA)) +
  labs(
    title = "Implied ball speed by lane length, all 17 weeks",
    subtitle = "Median within 2-yard bins of lane length",
    x = "lane length (yd)",
    y = "implied ball speed (yd/s)",
    colour = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

ggsave(
  here("figs", "week06", "arrival_speed_equalization.png"),
  p_speed,
  width = 7,
  height = 4,
  dpi = 200
)

p_speed
```

**Completions and incompletions agree within
`r sprintf("%.2f", max(abs(speed_wide$diff_v)))` yd/s in every bin**, against
about 1 yd/s on week 1. This is the result the definition rests on: the anchor
does not time completions and incompletions differently.

The season-scale curve is not flat across depth, as week 1 suggested. Implied
speed rises from about `r sprintf("%.0f", speed_wide$p50_v_C[1])` yd/s on
8–10 yd lanes to about `r sprintf("%.0f", max(speed_wide$p50_v_C))` yd/s
beyond 18 yd, then levels off. Both classes trace the same curve, so it is not
an outcome artifact. Two explanations fit, and the data here cannot separate
them: short passes are thrown with more touch, and any fixed offset in the
`pass_forward` frame inflates flight time proportionally more on short lanes.

# 6. Bounce check

```{r}
#| label: bounce-check

bounce <- approach |>
  inner_join(
    select(arrival, game_id, play_id, f_arr, f_min),
    by = c("game_id", "play_id")
  ) |>
  filter(frame_id >= f_arr, frame_id <= f_min) |>
  group_by(game_id, play_id) |>
  summarize(max_rise = max(d) - first(d), .groups = "drop")

bounce |>
  summarize(
    plays = n(),
    `rise > 1 yd` = sum(max_rise > 1),
    `rise > 3 yd` = sum(max_rise > 3),
    p99_rise = quantile(max_rise, 0.99)
  ) |>
  pivot_longer(everything(), names_to = "statistic") |>
  gt() |>
  fmt_number(columns = value, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

`r sum(bounce$max_rise > 1)` plays of `r comma(nrow(bounce))`
(`r percent(mean(bounce$max_rise > 1), accuracy = 0.01)`) show more than a
yard of rise between the arrival frame and closest approach, against 6 of 890
on week 1, and p99 is
`r sprintf("%.2f", quantile(bounce$max_rise, 0.99))` yd. No rarer artifact
appears at 17 times the sample.

# 7. Sensitivity at season scale

`approach.parquet` holds the full search window, so the anchor can be
recomputed under other parameter values without rescanning tracking.

```{r}
#| label: sensitivity

base <- select(arrival, game_id, play_id, f_base = f_arr)

agree <- function(...) {
  detect_arrival(approach, ...) |>
    select(game_id, play_id, f_arr) |>
    inner_join(base, by = c("game_id", "play_id")) |>
    summarize(
      agreement = mean(f_arr == f_base),
      median_shift = median(f_arr - f_base)
    )
}

stopifnot(identical(
  detect_arrival(approach) |>
    arrange(game_id, play_id) |>
    pull(f_arr),
  arrange(arrival, game_id, play_id)$f_arr
))

sens <- bind_rows(
  agree(dis_flight = 1.0) |> mutate(parameter = "delta = 1.0"),
  agree(dis_flight = 1.6) |> mutate(parameter = "delta = 1.6"),
  agree(gap_tol = 0L) |> mutate(parameter = "gap_tol = 0"),
  agree(gap_tol = 1L) |> mutate(parameter = "gap_tol = 1"),
  agree(gap_tol = 3L) |> mutate(parameter = "gap_tol = 3"),
  agree(gap_tol = 5L) |> mutate(parameter = "gap_tol = 5")
) |>
  relocate(parameter)

sens |>
  gt() |>
  tab_header(
    title = "Agreement with the adopted anchor (delta = 1.3, gap_tol = 2)"
  ) |>
  fmt_number(columns = agreement, decimals = 3) |>
  gt_theme_538(quiet = TRUE)
```

The adopted parameters reproduce `arrival.parquet` exactly (asserted). Frame
agreement stays high across a near two-fold range of $\delta$ with a median
shift of zero, as on week 1, and the gap tolerance only matters between 0 and
1.

# 8. Visual check

One completion, one incompletion, and one interception, drawn at random among
plays 10–25 yd deep that did not use the fallback. Each row shows four frames
around arrival. The dashed line is the lane from the ball at the throw to the
arrival point (×); the targeted receiver is ringed in gold; the ball is the
small brown point. At $f_{\text{arr}}$ the ball should be at the ×, close to
the target, and moving toward the target in the frames before.

```{r}
#| label: visual-check
#| fig-height: 9

set.seed(1961)
picks <- arr |>
  filter(
    pass_result %in% c("C", "I", "IN"),
    !used_fallback,
    between(depth_arr, 10, 25)
  ) |>
  group_by(pass_result) |>
  slice_sample(n = 1) |>
  ungroup()

rel_frames <- c(-4L, -2L, 0L, 2L)

frame_strip <- function(a1) {
  frames <- a1$f_arr + rel_frames
  rows <- trk |>
    filter(
      game_id == a1$game_id,
      play_id == a1$play_id,
      frame_id %in% frames
    ) |>
    select(frame_id, nfl_id, side, is_ball, x, y) |>
    collect() |>
    cast_keys() |>
    mutate(
      rel = factor(
        frame_id - a1$f_arr,
        levels = rel_frames,
        labels = c("f_arr - 4", "f_arr - 2", "f_arr", "f_arr + 2")
      ),
      is_target = !is_ball & nfl_id == a1$target_nfl_id
    )

  xr <- range(c(rows$x, a1$x_throw, a1$x_arr, a1$los_x), na.rm = TRUE)
  xlim <- c(max(0, xr[1] - 4), min(FIELD_LENGTH, xr[2] + 4))

  gg_field(los_x = a1$los_x, xlim = xlim) +
    geom_segment(
      data = a1,
      aes(x = x_throw, y = y_throw, xend = x_arr, yend = y_arr),
      linetype = "dashed",
      colour = "white"
    ) +
    geom_point(
      data = filter(rows, !is_ball),
      aes(x, y, fill = side),
      shape = 21,
      size = 2.4,
      colour = "grey15"
    ) +
    geom_point(
      data = filter(rows, is_target),
      aes(x, y),
      shape = 21,
      size = 4.8,
      colour = "#f5c518",
      stroke = 1.2
    ) +
    geom_point(
      data = a1,
      aes(x_arr, y_arr),
      shape = 4,
      size = 3,
      colour = "white",
      stroke = 1.1
    ) +
    geom_point(
      data = filter(rows, is_ball),
      aes(x, y),
      shape = 21,
      size = 1.8,
      fill = "#7a4a1f",
      colour = "white"
    ) +
    scale_fill_manual(values = c(offense = "#1f4fd8", defense = "#d62728")) +
    facet_wrap(~rel, nrow = 1) +
    labs(
      title = sprintf(
        "%s — game %s, play %s: d_arr %.2f yd, depth %.1f yd, flight %.1f s",
        a1$pass_result,
        a1$game_id,
        a1$play_id,
        a1$d_arr,
        a1$depth_arr,
        a1$t_flight
      ),
      fill = NULL
    ) +
    theme(
      legend.position = "none",
      plot.title = element_text(size = 9),
      strip.text = element_text(size = 8)
    )
}

p_visual <- wrap_plots(
  map(split(picks, seq_len(nrow(picks))), frame_strip),
  ncol = 1
)

ggsave(
  here("figs", "week06", "arrival_frames.png"),
  p_visual,
  width = 10,
  height = 9,
  dpi = 200
)

p_visual
```

## 8.1 Animations

The same three plays animated with `render_play()` from the snap to 10 frames
after arrival, one animation frame per tracking frame. Offense is blue, defense
red. GIFs are written to `figs/week06/` and are not committed (`*.gif` is
ignored); they are embedded in the rendered notebook.

```{r}
#| label: animations
#| results: asis

plays_meta <- read_parquet(here("data", "processed", "plays.parquet")) |>
  select(game_id, play_id, yards_to_go, play_description)

gif_paths <- map_chr(seq_len(nrow(picks)), \(i) {
  a1 <- picks[i, ]
  f_snap <- analytic_sample$f_ball_snap[
    analytic_sample$game_id == a1$game_id &
      analytic_sample$play_id == a1$play_id
  ]

  play_trk <- trk |>
    filter(
      game_id == a1$game_id,
      play_id == a1$play_id,
      frame_id >= f_snap,
      frame_id <= a1$f_arr + 10L
    ) |>
    collect() |>
    cast_keys() |>
    arrange(frame_id)

  teams <- play_trk |>
    filter(!is_ball) |>
    distinct(team_abbr, side)
  pal <- setNames(
    if_else(teams$side == "offense", "#1f4fd8", "#d62728"),
    teams$team_abbr
  )

  this_play <- a1 |>
    select(game_id, play_id, los_x) |>
    left_join(plays_meta, by = c("game_id", "play_id"))

  path <- here(
    "figs",
    "week06",
    sprintf("arrival_%s_%s_%s.gif", a1$pass_result, a1$game_id, a1$play_id)
  )
  render_play(play_trk, this_play, path, pal = pal)
  path
})

# Relative to this notebook, so the HTML can embed them.
for (g in gif_paths) {
  cat(sprintf("![](%s)\n\n", file.path("..", "figs", "week06", basename(g))))
}
```

# 9. External validation against `air_yards`

nflverse `air_yards` is charted by hand: the distance downfield from the line
of scrimmage to where the pass is caught or targeted. The tracking-derived
depth is
$\text{depth}_{\text{arr}} = x_{\text{arr}} - x_{\text{LOS}}$ in standardized
coordinates. The two are independent measurements of the same quantity, so
their agreement checks the anchor, the sign convention, and the line of
scrimmage at once.

```{r}
#| label: air-yards-join

av <- arr |>
  filter(pass_result %in% c("C", "I", "IN")) |>
  inner_join(
    select(pbp, game_id, play_id, air_yards) |> filter(!is.na(air_yards)),
    by = c("game_id", "play_id")
  ) |>
  mutate(diff = depth_arr - air_yards)

tibble(
  quantity = c(
    "plays compared",
    "correlation",
    "median depth_arr - air_yards (yd)",
    "IQR of the difference (yd)",
    "mean absolute difference (yd)"
  ),
  value = c(
    comma(nrow(av)),
    sprintf("%.3f", cor(av$depth_arr, av$air_yards)),
    sprintf("%.2f", median(av$diff)),
    sprintf(
      "[%.2f, %.2f]",
      quantile(av$diff, 0.25),
      quantile(av$diff, 0.75)
    ),
    sprintf("%.2f", mean(abs(av$diff)))
  )
) |>
  gt() |>
  gt_theme_538(quiet = TRUE)

av |>
  group_by(pass_result) |>
  summarize(
    plays = n(),
    correlation = cor(depth_arr, air_yards),
    p10_diff = quantile(diff, 0.10),
    p50_diff = median(diff),
    p90_diff = quantile(diff, 0.90),
    .groups = "drop"
  ) |>
  gt() |>
  tab_header(title = "depth_arr - air_yards by outcome") |>
  fmt_number(columns = -c(pass_result, plays), decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: air-yards-plot
#| fig-height: 5.5

p_air <- av |>
  ggplot(aes(air_yards, depth_arr)) +
  geom_point(alpha = 0.06, size = 0.6) +
  geom_abline(colour = "#1f4fd8", linewidth = 0.5) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey40") +
  geom_vline(xintercept = 0, linewidth = 0.3, colour = "grey40") +
  coord_equal() +
  labs(
    title = "Tracking-derived depth against charted air yards",
    subtitle = "Blue line: equality",
    x = "nflverse air_yards",
    y = "depth_arr = x_arr - LOS (yd)"
  ) +
  theme_minimal()

ggsave(
  here("figs", "week06", "depth_vs_air_yards.png"),
  p_air,
  width = 6,
  height = 6,
  dpi = 200
)

p_air
```

```{r}
#| label: air-yards-by-band

av |>
  mutate(
    air_yards_band = cut(
      air_yards,
      c(-Inf, -5, -1, 0, 4, 9, 19, Inf),
      labels = c("<= -5", "-4 to -1", "0", "1-4", "5-9", "10-19", "20+")
    )
  ) |>
  group_by(air_yards_band) |>
  summarize(
    plays = n(),
    p10_depth = quantile(depth_arr, 0.10),
    p50_depth = median(depth_arr),
    p90_depth = quantile(depth_arr, 0.90),
    mean_diff = mean(diff),
    .groups = "drop"
  ) |>
  gt() |>
  tab_header(title = "depth_arr within bands of charted air yards") |>
  fmt_number(columns = -c(air_yards_band, plays), decimals = 2) |>
  gt_theme_538(quiet = TRUE)

zero <- filter(av, air_yards == 0)
by_class <- av |>
  group_by(pass_result) |>
  summarize(m = median(diff))
```

**The two measurements agree closely and the sign convention is right**: the
correlation is `r sprintf("%.3f", cor(av$depth_arr, av$air_yards))`, and the
median difference is `r sprintf("%.2f", median(av$diff))` yd. The tracking
depth runs slightly short of the charted value, as it should: the anchor is the
last frame the ball is still in flight, about a frame before the ball reaches
the receiver, while charting records the catch point. The shortfall grows on
deep throws, where the ball covers about 2 yd per frame.

**The offset does not depend on the outcome**: the median difference is
`r sprintf("%.2f", by_class$m[by_class$pass_result == "C"])` yd on completions
and `r sprintf("%.2f", by_class$m[by_class$pass_result == "I"])` on
incompletions. The spread is wider on incompletions, where the charter records
an intended spot rather than a catch.

**Charted zero is mostly just behind the line.** The
`r comma(nrow(zero))` throws charted at exactly zero air yards have a median
tracking depth of `r sprintf("%.2f", median(zero$depth_arr))` yd, and
`r percent(mean(zero$depth_arr <= 0), accuracy = 1)` of them arrive at or
behind the line by tracking. Zero is a charting convention for throws at the
line, which is why the threshold is set on the tracking depth.

## 9.1 The beyond-the-line-of-scrimmage threshold

```{r}
#| label: threshold-candidates

thresholds <- c(0, 0.5, 1, 2)

map(thresholds, \(th) {
  keep <- av$depth_arr > th
  tibble(
    rule = sprintf("depth_arr > %.1f", th),
    kept = sum(keep),
    pct_kept = mean(keep),
    `kept, charted <= 0` = sum(keep & av$air_yards <= 0),
    `dropped, charted > 0` = sum(!keep & av$air_yards > 0),
    complete_kept = mean(av$pass_result[keep] == "C"),
    complete_dropped = mean(av$pass_result[!keep] == "C")
  )
}) |>
  list_rbind() |>
  bind_rows(tibble(
    rule = "air_yards > 0 (charted)",
    kept = sum(av$air_yards > 0),
    pct_kept = mean(av$air_yards > 0),
    complete_kept = mean(av$pass_result[av$air_yards > 0] == "C"),
    complete_dropped = mean(av$pass_result[av$air_yards <= 0] == "C")
  )) |>
  gt() |>
  tab_header(title = "Candidate beyond-the-line rules") |>
  fmt_number(
    columns = c(kept, starts_with("kept,"), starts_with("dropped,")),
    decimals = 0
  ) |>
  fmt_percent(
    columns = c(pct_kept, complete_kept, complete_dropped),
    decimals = 1
  ) |>
  sub_missing(missing_text = "") |>
  gt_theme_538(quiet = TRUE)

at0 <- av$depth_arr > 0
```

The adopted rule is **$\text{depth}_{\text{arr}} > 0$**: the ball arrives
beyond the line of scrimmage. It keeps `r comma(sum(at0))` throws
(`r percent(mean(at0), accuracy = 0.1)`), within
`r abs(sum(at0) - sum(av$air_yards > 0))` of the charted rule, and the two
disagree on
`r comma(sum(at0 != (av$air_yards > 0)))` throws
(`r percent(mean(at0 != (av$air_yards > 0)), accuracy = 0.1)`):
`r sum(at0 & av$air_yards <= 0)` kept although charted at or behind the line,
and `r sum(!at0 & av$air_yards > 0)` dropped although charted beyond it. Raising the threshold mostly drops throws that were
charted beyond the line. The excluded throws complete at
`r percent(mean(av$pass_result[!at0] == "C"), accuracy = 0.1)`, against
`r percent(mean(av$pass_result[at0] == "C"), accuracy = 0.1)` for those kept:
screens and other throws at the line are easy completions, so this is a
population definition, stated with its composition, not ignorable attrition.
The threshold is recorded in `notes/decisions.md` §7.
