---
title: "Robustness of the stage comparisons"
subtitle: "STAT 1961 Capstone — sensitivity of the tracking gain to measurement and sample choices"
author: "David Neumann"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: [brite, ../theme/capstone.scss]
    code-summary: "Show code"
    fig-width: 8
    fig-height: 5
    embed-resources: true
execute:
  warning: false
  message: false
---

# 0. Setup

```{r}
#| label: setup
#| include: false
if (Sys.getenv("RENV_PROJECT") == "") {
  root <- normalizePath("..", mustWork = TRUE)
  Sys.setenv(RENV_PROJECT = root)
  setwd(root)
  source("renv/activate.R")
}

library(tidyverse)
library(arrow)
library(here)
library(gt)
library(gtExtras)
library(scales)

# mgcv must be attached: the spec formulas use bare s() and ti().
library(mgcv)

source(here("R", "constants.R"))
source(here("R", "utils.R"))
source(here("R", "evaluate.R"))
```

## 0.1 What this notebook is

`analysis/03_model_baseline.qmd` scores four nested stages on
`model_frame.parquet` and finds that coverage geometry at the throw (stage 4)
improves on the play-by-play baseline (stage 2) by nearly as much as the
play-by-play features improve on the base rate. This notebook checks whether that comparison
depends on choices made along the way (`notes/decisions.md` §6.7, §7):

| Check | Change |
|---|---|
| R1 | Charted `air_yards` (with its zero indicator) instead of the tracking depth $\text{depth}_{\text{arr}}$, everywhere depth enters |
| R2 | Stage 4's timing features with the p90 motion constants (9.98 yd/s, 6.50 yd/s²) instead of the median (9.22, 5.90) |
| R3 | Without plays whose arrival used the closest-approach fallback |
| R4 | Without plays with a kinematic defect outside the snap-to-throw window |
| R5 | Without plays with a QB hit (the QB-hit term leaves the specs) |
| R6 | Without plays whose closest approach is at the last searched frame |
| R7 | Charted three-level `pass_location` instead of `pass_middle` and the receiver's sideline distance |

Each check refits the stages with leave-one-week-out cross-validation and reports
the paired differences in out-of-fold log loss with game-cluster bootstrap
intervals, as in notebook 03. R6 selects on the outcome — those plays are mostly
completions where the receiver carries the ball — so it is reported with that
caveat rather than as a clean check.

It reads `model_frame.parquet` and, for the baseline, the out-of-fold
predictions persisted by `scripts/09_fit_models.R`; it writes nothing to
`data/`.

```{r}
#| label: load_data

mf <- read_parquet(here("data", "processed", "model_frame.parquet"))
stopifnot(identical(mf, assert_complete(mf)))
```

# 1. Method

```{r}
#| label: helpers

stage_specs <- MODEL_SPECS[c("2. play-by-play", "3. + separation", "4. + geometry")]

# Replace variable names inside each spec's formula, keeping its environment.
swap_vars <- function(specs, from, to) {
  map(specs, \(spec) {
    txt <- paste(deparse(spec$formula, width.cutoff = 500L), collapse = " ")
    for (k in seq_along(from)) {
      txt <- gsub(from[k], to[k], txt, fixed = TRUE)
    }
    spec$formula <- stats::as.formula(txt, env = environment(spec$formula))
    spec
  })
}

edit_terms <- function(specs, change) {
  map(specs, \(spec) {
    spec$formula <- stats::update(spec$formula, change)
    spec
  })
}

pairs <- tribble(
  ~reference, ~model, ~comparison,
  "2. play-by-play", "3. + separation", "stage 3 - stage 2",
  "2. play-by-play", "4. + geometry", "stage 4 - stage 2",
  "3. + separation", "4. + geometry", "stage 4 - stage 3"
)

score <- function(preds, df, label) {
  ll <- preds |>
    group_by(model) |>
    summarize(log_loss = log_loss(complete, .pred), .groups = "drop")

  deltas <- pmap(pairs, \(reference, model, comparison) {
    paired_delta(preds, reference, model, "game_id") |>
      mutate(comparison = comparison)
  }) |>
    list_rbind()

  deltas |>
    select(comparison, estimate, conf_low, conf_high) |>
    mutate(
      scenario = label,
      n = nrow(df),
      base_rate = mean(df$complete),
      ll_stage2 = ll$log_loss[ll$model == "2. play-by-play"],
      ll_stage4 = ll$log_loss[ll$model == "4. + geometry"]
    )
}

run <- function(df, specs, label) {
  assert_complete(df, vars = model_vars(specs))
  folds <- make_folds(df)
  preds <- cv_predict_all(df, folds, specs = specs, verbose = FALSE)
  list(preds = preds, result = score(preds, df, label))
}
```

The baseline is not refit: it is scored from the out-of-fold predictions that
`scripts/09_fit_models.R` persisted, the same ones notebook 03 reports, so the
two notebooks agree by construction. `run()` calls the same harness functions
as `09` (`make_folds()`, `cv_predict_all()`), so each check below differs from
the baseline only in what it changes.

```{r}
#| label: baseline

# read_oof_preds() fails if MODEL_SPECS or the model frame has changed
# since 09 ran.
base_preds <- read_oof_preds(mf) |>
  filter(model %in% names(stage_specs))
base <- list(preds = base_preds, result = score(base_preds, mf, "baseline"))

base$result |>
  select(comparison, estimate, conf_low, conf_high) |>
  gt() |>
  tab_header(title = "Baseline, as in notebook 03") |>
  fmt_number(columns = c(estimate, conf_low, conf_high), decimals = 4) |>
  gt_theme_538(quiet = TRUE)
```

# 2. Measurement choices

## 2.1 R1: charted air yards

```{r}
#| label: r1-air-yards

mf_air <- mf |>
  mutate(
    dist_to_sticks_air = air_yards - yards_to_go,
    air_yards_zero = as.integer(air_yards == 0)
  )

specs_air <- stage_specs |>
  swap_vars(
    c("depth_arr", "dist_to_sticks"),
    c("air_yards", "dist_to_sticks_air")
  ) |>
  edit_terms(. ~ . + air_yards_zero)

r1 <- run(mf_air, specs_air, "R1 charted air yards")
```

## 2.2 R2: motion constants

Only stage 4 uses the motion model, so stage 4 is refit with the p90 timing
features and compared against the baseline's stages 2 and 3.

```{r}
#| label: r2-motion

mf_p90 <- mf |>
  mutate(tta_nearest = tta_nearest_p90, window_margin = window_margin_p90)

# Folds from mf_p90 itself: an rsample split carries the data it was built
# from, so folds built from mf would refit on the median-constant features.
folds_p90 <- make_folds(mf_p90)
preds_r2 <- bind_rows(
  filter(base$preds, model != "4. + geometry"),
  cv_predict(
    mf_p90,
    folds_p90,
    stage_specs[["4. + geometry"]],
    "4. + geometry",
    verbose = FALSE
  )
)

# The refit must actually differ from the baseline's stage 4.
stopifnot(!isTRUE(all.equal(
  preds_r2$.pred[preds_r2$model == "4. + geometry"],
  base$preds$.pred[base$preds$model == "4. + geometry"]
)))
r2 <- list(result = score(preds_r2, mf, "R2 p90 motion constants"))

# Stage 4 with p90 constants against stage 4 with median constants.
d_p90 <- paired_delta(
  bind_rows(
    filter(base$preds, model == "4. + geometry"),
    filter(preds_r2, model == "4. + geometry") |> mutate(model = "4 (p90)")
  ),
  "4. + geometry",
  "4 (p90)",
  "game_id"
)
```

With the p90 constants, stage 4's out-of-fold log loss changes by
`r sprintf("%+.4f", d_p90$estimate)` [`r sprintf("%+.4f", d_p90$conf_low)`,
`r sprintf("%+.4f", d_p90$conf_high)`] nats against the median constants.

## 2.3 R7: charted pass location

Stage 2 codes location as middle or not plus a smooth in the targeted
receiver's distance to the sideline (`notes/decisions.md` §9.13). R7 restores
the charted three-level `pass_location` in their place, in every stage.

```{r}
#| label: r7-location

specs_loc <- edit_terms(
  stage_specs,
  . ~ . - pass_middle - s(sideline_rec) + pass_location
)
# update() leaves a misspelled term in place silently.
stopifnot(!any(map_lgl(specs_loc, \(s) {
  any(c("pass_middle", "sideline_rec") %in% all.vars(s$formula))
})))

r7 <- run(mf, specs_loc, "R7 charted pass_location")
```

# 3. Sample choices

```{r}
#| label: r3-r6

r3 <- run(filter(mf, !used_fallback), stage_specs, "R3 without arrival fallbacks")
r4 <- run(filter(mf, !has_kinematic_defect), stage_specs, "R4 without kinematic defects")
r5 <- run(
  filter(mf, qb_hit == 0),
  edit_terms(stage_specs, . ~ . - s(depth_arr, by = qb_hit)),
  "R5 without QB-hit plays"
)
r6 <- run(filter(mf, !at_window_edge), stage_specs, "R6 without at_window_edge (outcome-selected)")
```

# 4. Results

```{r}
#| label: results

results <- bind_rows(
  base$result,
  r1$result,
  r2$result,
  r3$result,
  r4$result,
  r5$result,
  r6$result,
  r7$result
) |>
  mutate(scenario = fct_inorder(scenario))

results |>
  mutate(value = sprintf("%.4f [%.4f, %.4f]", estimate, conf_low, conf_high)) |>
  select(scenario, n, base_rate, ll_stage2, ll_stage4, comparison, value) |>
  pivot_wider(names_from = comparison, values_from = value) |>
  gt() |>
  tab_header(
    title = "Out-of-fold log loss differences under each check",
    subtitle = "Game-cluster bootstrap, 10,000 draws, 95% percentile interval"
  ) |>
  fmt_number(columns = n, decimals = 0) |>
  fmt_number(columns = c(base_rate, ll_stage2, ll_stage4), decimals = 4) |>
  gt_theme_538(quiet = TRUE)

rs <- function(scenario_prefix, comparison, col = "estimate") {
  results[[col]][
    startsWith(as.character(results$scenario), scenario_prefix) &
      results$comparison == comparison
  ]
}
```

```{r}
#| label: results-plot
#| fig-height: 5

p_rob <- results |>
  filter(comparison != "stage 4 - stage 3") |>
  ggplot(aes(estimate, fct_rev(scenario), colour = comparison)) +
  geom_vline(xintercept = 0, linewidth = 0.3, colour = "grey40") +
  geom_pointrange(
    aes(xmin = conf_low, xmax = conf_high),
    position = position_dodge(width = 0.5),
    size = 0.3
  ) +
  labs(
    title = "Tracking gain over the play-by-play baseline, by check",
    subtitle = "Out-of-fold log loss difference, 95% game-cluster interval",
    x = "difference in log loss (nats; negative favours the later stage)",
    y = NULL,
    colour = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

ggsave(
  here("figs", "week06", "robustness.png"),
  p_rob,
  width = 8,
  height = 5,
  dpi = 200
)

p_rob
```

**The tracking gain does not depend on any of these choices.** Across every
check but R6,
stage 4 improves on stage 2 by between
`r sprintf("%.4f", -max(results$estimate[results$comparison == "stage 4 - stage 2" & !startsWith(as.character(results$scenario), "R6")]))`
and
`r sprintf("%.4f", -min(results$estimate[results$comparison == "stage 4 - stage 2" & !startsWith(as.character(results$scenario), "R6")]))`
nats, against `r sprintf("%.4f", -rs("baseline", "stage 4 - stage 2"))` in the
baseline, and every interval is well away from zero and overlaps the
baseline's. Stage 3's gain over stage 2 and stage 4's over stage 3 are equally
stable.

- **R1, charted air yards.** Stage 2 is slightly *better* with charted air yards
  than with the tracking depth (log loss
  `r sprintf("%.4f", rs("R1", "stage 4 - stage 2", "ll_stage2"))` against
  `r sprintf("%.4f", rs("baseline", "stage 4 - stage 2", "ll_stage2"))`): the
  charter records the catch or target point, which the last in-flight frame
  approximates. The geometry's gain over that stronger baseline is
  `r sprintf("%.4f", -rs("R1", "stage 4 - stage 2"))`
  [`r sprintf("%.4f", -rs("R1", "stage 4 - stage 2", "conf_high"))`,
  `r sprintf("%.4f", -rs("R1", "stage 4 - stage 2", "conf_low"))`], so the
  result does not rest on the depth measure.
- **R2, motion constants.** The p90 constants fit slightly worse than the
  median ones (above), and the gain is essentially unchanged: the choice of
  motion-model parameters does not drive the result.
- **R3 and R4, arrival fallbacks and kinematic defects.** Dropping either set
  leaves every comparison where it was.
- **R5, QB-hit plays.** Dropping them, and the QB-hit term with them, leaves
  every comparison where it was.
- **R6, `at_window_edge`.** The gain is larger without these plays, but they are
  mostly completions where the receiver carries the ball, so dropping them
  lowers the base rate from
  `r sprintf("%.3f", rs("baseline", "stage 4 - stage 2", "base_rate"))` to
  `r sprintf("%.3f", rs("R6", "stage 4 - stage 2", "base_rate"))` and changes the
  population on the outcome. It is not evidence either way; it shows only that
  these plays are not where the gain comes from.
- **R7, charted pass location.** With the three charted levels in place of
  `pass_middle` and the sideline distance, stage 2's log loss is
  `r sprintf("%.4f", rs("R7", "stage 4 - stage 2", "ll_stage2"))` against
  `r sprintf("%.4f", rs("baseline", "stage 4 - stage 2", "ll_stage2"))`, and the
  geometry's gain over it is `r sprintf("%.4f", -rs("R7", "stage 4 - stage 2"))`
  [`r sprintf("%.4f", -rs("R7", "stage 4 - stage 2", "conf_high"))`,
  `r sprintf("%.4f", -rs("R7", "stage 4 - stage 2", "conf_low"))`]: the
  stronger location baseline costs the tracking features little.
