---
title: "Term contributions"
subtitle: "STAT 1961 Capstone — what each tracking term adds to stage 4, out of fold"
author: "David Neumann"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: [brite, ../theme/capstone.scss]
    code-summary: "Show code"
    fig-width: 8
    fig-height: 5
    embed-resources: true
execute:
  warning: false
  message: false
---

# 0. Setup

```{r}
#| label: setup
#| include: false
if (Sys.getenv("RENV_PROJECT") == "") {
  root <- normalizePath("..", mustWork = TRUE)
  Sys.setenv(RENV_PROJECT = root)
  setwd(root)
  source("renv/activate.R")
}

library(tidyverse)
library(arrow)
library(here)
library(gt)
library(gtExtras)
library(scales)

# mgcv must be attached: the spec formulas use bare s() and ti().
library(mgcv)

source(here("R", "constants.R"))
source(here("R", "utils.R"))
source(here("R", "evaluate.R"))
```

## 0.1 What this notebook is

Stage 4 adds six tracking terms to the play-by-play baseline (stage 2):
separation at the throw, its interaction with depth, closing speed, the nearest
defender's leverage angle, time-to-arrival margin, and the passing-window
margin. `analysis/03_model_baseline.qmd` measures what they add as two blocks.
This notebook asks what each one adds **given the others**: stage 4 is refit
without a term, or a block of terms, and the rise in out-of-fold log loss is
that term's unique contribution.

It is diagnostic. No spec changes here; a term that adds nothing becomes a
candidate for removal, to be argued in `notes/decisions.md`.

It reads `model_frame.parquet` and the predictions and fits persisted by
`scripts/09_fit_models.R`, and writes nothing to `data/`.

```{r}
#| label: load_data

mf <- read_parquet(here("data", "processed", "model_frame.parquet"))
stopifnot(identical(mf, assert_complete(mf)))

oof <- read_oof_preds(mf)
fits <- read_full_fits(mf)
```

# 1. Method

For a term or block $j$, with $\hat p_i^{(4)}$ stage 4's out-of-fold prediction
for play $i$ and $\hat p_i^{(4 \setminus j)}$ that of stage 4 refit without
$j$,

$$
\bar d_j = \frac{1}{n} \sum_{i=1}^{n}
\left[ \ell\big(y_i, \hat p_i^{(4 \setminus j)}\big)
     - \ell\big(y_i, \hat p_i^{(4)}\big) \right],
$$

where $\ell$ is the per-play log loss. A positive $\bar d_j$ means stage 4 is
worse without $j$. The interval is the game-cluster bootstrap of notebook 03
(`paired_delta()`), and the folds are the same leave-one-week-out folds.

**Drop-one measures unique contribution.** Two terms that carry the same
information can each have a small $\bar d_j$ while the pair matters a great
deal: dropping either lets the other absorb its effect. The blocks in the table
measure that shared part, and the concurvity in §3 shows which terms overlap.
The individual rows therefore do not sum to the block rows.

**Separation and its depth interaction are dropped together** in the
separation row. `ti()` is built to sit alongside its main effects, so
`ti(sep_throw, depth_arr)` without `s(sep_throw)` would be a different model of
separation rather than stage 4 without it. The interaction alone gets its own
row.

The geometry block, all three stage 4 terms, is stage 4 against stage 3 and
needs no refit.

**Stage 2's features are measured the same way**, to show which features the
model leans on overall. These rows are descriptive: baseline terms are chosen
by argument, never cut on numbers like these (`notes/decisions.md` §9.3). Throw
depth drops `s(depth_arr)`, `s(dist_to_sticks)` (depth minus yards to go), and
the separation-by-depth interaction together, since the interaction would
otherwise carry depth without its main effect; depth still enters through the
QB-hit smooth. Throw location drops `pass_middle` and `s(sideline_rec)`
together.

```{r}
#| label: units

full <- MODEL_SPECS[["4. + geometry"]]

# Term labels exactly as terms() writes them.
units <- list(
  `separation (s + ti)` = c("s(sep_throw)", "ti(sep_throw, depth_arr)"),
  `separation x depth (ti)` = "ti(sep_throw, depth_arr)",
  `closing speed` = "s(closing_throw)",
  `leverage angle` = "s(lev_angle, bs = \"cc\", k = 8)",
  `time to arrival` = "s(tta_nearest)",
  `passing window` = "s(window_margin)",
  `separation block (stage 3 terms)` = c(
    "s(sep_throw)",
    "s(closing_throw)",
    "ti(sep_throw, depth_arr)"
  ),
  `throw depth` = c(
    "s(depth_arr)",
    "s(dist_to_sticks)",
    "ti(sep_throw, depth_arr)"
  ),
  `QB hit` = "s(depth_arr, by = qb_hit)",
  `pass rushers` = "s(number_of_pass_rushers, k = 5)",
  `field position` = "s(los_x)",
  down = "down",
  `throw location` = c("pass_middle", "s(sideline_rec)"),
  shotgun = "shotgun",
  home = "home"
)

stage_of <- c(
  `separation (s + ti)` = "stage 3",
  `separation x depth (ti)` = "stage 3",
  `closing speed` = "stage 3",
  `leverage angle` = "stage 4",
  `time to arrival` = "stage 4",
  `passing window` = "stage 4",
  `throw depth` = "stage 2",
  `QB hit` = "stage 2",
  `pass rushers` = "stage 2",
  `field position` = "stage 2",
  down = "stage 2",
  `throw location` = "stage 2",
  shotgun = "stage 2",
  home = "stage 2"
)

# Remove terms by label, keeping the formula's environment so that bare s()
# resolves. Fails on a label that is not in the formula, rather than
# leaving the formula unchanged as update() would.
drop_terms <- function(spec, drop) {
  labels <- attr(stats::terms(spec$formula), "term.labels")
  stopifnot(all(drop %in% labels))
  spec$formula <- stats::reformulate(
    setdiff(labels, drop),
    response = "complete",
    env = environment(spec$formula)
  )
  spec
}

reduced <- map(units, \(drop) drop_terms(full, drop))

n_full <- length(attr(terms(full$formula), "term.labels"))
stopifnot(
  map_int(reduced, \(s) length(attr(terms(s$formula), "term.labels"))) ==
    n_full - lengths(units)
)
```

# 2. Contributions

Each row refits stage 4 seventeen times; this chunk is the notebook's cost,
about 20 minutes.

```{r}
#| label: refit

folds <- make_folds(mf)

preds_drop <- imap(
  reduced,
  \(spec, name) cv_predict(mf, folds, spec, name, verbose = FALSE)
) |>
  list_rbind()

preds <- bind_rows(
  filter(oof, model %in% c("3. + separation", "4. + geometry")),
  preds_drop
)

# Every refit must actually differ from stage 4.
p4 <- oof$.pred[oof$model == "4. + geometry"]
stopifnot(map_lgl(
  names(units),
  \(m) !isTRUE(all.equal(preds$.pred[preds$model == m], p4))
))
```

```{r}
#| label: contributions

contrib <- c(names(units), "3. + separation") |>
  map(\(m) paired_delta(preds, "4. + geometry", m, "game_id")) |>
  list_rbind() |>
  mutate(
    unit = if_else(
      model == "3. + separation",
      "geometry block (stage 4 terms)",
      model
    ),
    kind = if_else(str_detect(unit, "block"), "block", "term")
  )

contrib |>
  select(kind, unit, estimate, conf_low, conf_high) |>
  arrange(kind == "block", desc(estimate)) |>
  gt(groupname_col = "kind") |>
  tab_header(
    title = "Rise in out-of-fold log loss when a term is dropped from stage 4",
    subtitle = "Positive: stage 4 is worse without it. Game-cluster bootstrap, 95% interval"
  ) |>
  fmt_number(columns = c(estimate, conf_low, conf_high), decimals = 4) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: contributions-plot
#| fig-height: 4

contrib |>
  mutate(unit = fct_reorder(unit, estimate)) |>
  ggplot(aes(estimate, unit, colour = kind)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = conf_low, xmax = conf_high)) +
  scale_colour_manual(values = c(block = "grey40", term = "#1f5f8b")) +
  labs(
    x = "Rise in out-of-fold log loss without it (nats)",
    y = NULL,
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

For the slides: one bar per feature, coloured by the stage that adds it. The
separation-by-depth interaction is left out because the separation bar
already includes it.

```{r}
#| label: feature-contributions-figure
#| fig-height: 5

feature_labels <- c(
  `separation (s + ti)` = "Separation",
  `closing speed` = "Closing speed",
  `time to arrival` = "Time to arrival",
  `passing window` = "Passing window",
  `leverage angle` = "Leverage",
  `throw depth` = "Throw depth",
  `QB hit` = "QB hit",
  `pass rushers` = "Pass rushers",
  `field position` = "Field position",
  down = "Down",
  `throw location` = "Throw location",
  shotgun = "Shotgun",
  home = "Home"
)

feat <- contrib |>
  filter(unit %in% names(feature_labels)) |>
  mutate(
    feature = fct_reorder(feature_labels[unit], estimate),
    stage = stage_of[unit]
  )

p_feat <- ggplot(feat, aes(estimate, feature, fill = stage)) +
  geom_col(width = 0.6) +
  geom_errorbar(
    aes(xmin = conf_low, xmax = conf_high),
    width = 0.25,
    colour = "grey35",
    linewidth = 0.4
  ) +
  geom_text(
    aes(x = pmax(conf_high, estimate), label = sprintf("%.4f", estimate)),
    hjust = -0.2,
    size = 3.2,
    colour = "grey15"
  ) +
  scale_fill_manual(
    values = c(
      "stage 2" = "#2a78d6",
      "stage 3" = "#eb6834",
      "stage 4" = "#1baf7a"
    ),
    labels = c(
      "stage 2" = "stage 2 (baseline)",
      "stage 3" = "stage 3 (separation)",
      "stage 4" = "stage 4 (geometry)"
    ),
    name = NULL
  ) +
  scale_x_continuous(
    limits = c(min(0, min(feat$conf_low)), max(feat$conf_high) * 1.2),
    expand = c(0, 0)
  ) +
  labs(
    title = "How much worse the model gets without each feature",
    subtitle = "Rise in out-of-fold log loss when the feature is removed from stage 4; bars show 95% intervals",
    x = "rise in log loss",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "top",
    legend.justification = "left"
  )

dir.create(here("figs", "week07"), showWarnings = FALSE)
ggsave(
  here("figs", "week07", "feature_contributions.png"),
  p_feat,
  width = 8,
  height = 5,
  dpi = 200
)

p_feat
```

# 3. Overlap between terms

`concurvity()` is the smooth-term analogue of collinearity: for each smooth,
the share of it that the other terms can reproduce (0 none, 1 all). The
`worst` measure is the pessimistic one. It is computed on the stage 4 fit on
all rows and is descriptive.

```{r}
#| label: concurvity-full

f4 <- fits[["4. + geometry"]]
tracking <- c(
  "s(sep_throw)",
  "ti(sep_throw,depth_arr)",
  "s(closing_throw)",
  "s(lev_angle)",
  "s(tta_nearest)",
  "s(window_margin)"
)

cc_full <- concurvity(f4, full = TRUE)

tibble(
  smooth = tracking,
  worst = cc_full["worst", tracking],
  observed = cc_full["observed", tracking]
) |>
  gt() |>
  tab_header(
    title = "Concurvity of each tracking smooth with all other terms"
  ) |>
  fmt_number(columns = c(worst, observed), decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

Pairwise, among the tracking smooths and depth (`worst`; row term as
reproduced by the column term):

```{r}
#| label: concurvity-pairwise

cc_pair <- concurvity(f4, full = FALSE)$worst
keep <- c("s(depth_arr)", tracking)
cc_keep <- cc_pair[keep, keep]
# A term trivially reproduces itself.
diag(cc_keep) <- NA

cc_keep |>
  as.data.frame() |>
  rownames_to_column("smooth") |>
  gt() |>
  fmt_number(columns = -smooth, decimals = 2) |>
  sub_missing(missing_text = "") |>
  data_color(
    columns = -smooth,
    palette = c("white", "#1f5f8b"),
    domain = c(0, 1),
    na_color = "white"
  ) |>
  gt_theme_538(quiet = TRUE)
```
