---
title: "Lateral location"
subtitle: "STAT 1961 Capstone — tracking-derived throw location as a replacement for charted pass_location"
author: "David Neumann"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: [brite, ../theme/capstone.scss]
    code-summary: "Show code"
    fig-width: 8
    fig-height: 5
    embed-resources: true
execute:
  warning: false
  message: false
---

# 0. Setup

```{r}
#| label: setup
#| include: false
if (Sys.getenv("RENV_PROJECT") == "") {
  root <- normalizePath("..", mustWork = TRUE)
  Sys.setenv(RENV_PROJECT = root)
  setwd(root)
  source("renv/activate.R")
}

library(tidyverse)
library(arrow)
library(here)
library(gt)
library(gtExtras)
library(scales)

# mgcv must be attached: the spec formulas use bare s() and ti().
library(mgcv)

source(here("R", "constants.R"))
source(here("R", "utils.R"))
source(here("R", "evaluate.R"))
```

## 0.1 What this notebook is

Stage 2 codes the throw's direction with charted `pass_location` (left,
middle, right). `notes/decisions.md` §9.13 derives two tracking measures from
the arrival point: `lat_arr`, the lateral offset from the release point, of
which the charted variable is a three-level version, and `sideline_arr`, the
distance to the nearer sideline, which the charted variable does not measure.
This notebook compares candidate stage 2 specs that replace `pass_location`
with them, and refits stage 4 on the candidates to see what happens to the
tracking gain. No spec changes here.

Expectations, recorded before fitting (§9.13):

- **A** (`s(lat_arr)` for `pass_location`): about 0 to 0.001 nats, since it is
  the same information made continuous.
- **B** (A + `s(sideline_arr)`): 0.002–0.006 nats, from the boundary.
- **C** (B with `sideline_arr` clamped at 0): below B. Out-of-bounds arrivals
  are largely missed throws, so the gap from B to C is how much of B's gain is
  throw accuracy, which is decided after the release, rather than location.
- **Stage 4 on B or C**: the gain over its own stage 2 shrinks slightly, mostly
  at the expense of `s(window_margin)`, which overlaps with depth (§9.12).

It reads `model_frame.parquet` and the predictions persisted by
`scripts/09_fit_models.R`, and writes nothing to `data/`.

Candidate F (§5) was adopted. "Current" stage 2 and 4 in this notebook are
therefore the specs as they were before, labelled `2 (charted)` and
`4 (charted)`: rebuilt from `MODEL_SPECS` by putting charted `pass_location`
back in place of the adopted terms, and refit here. Candidate F is checked to
reproduce the persisted stage 2 and 4 predictions.

```{r}
#| label: load_data

mf <- read_parquet(here("data", "processed", "model_frame.parquet")) |>
  mutate(
    sideline_in = pmax(sideline_arr, 0),
    sideline_rec_in = pmax(sideline_rec, 0)
  )
stopifnot(identical(mf, assert_complete(mf)))

oof <- read_oof_preds(mf)
```

# 1. Candidates

```{r}
#| label: candidates

# Replace terms by label, keeping the formula's environment so that bare s()
# resolves. Fails on a label to drop that is not in the formula.
swap_terms <- function(spec, drop = character(0), add = character(0)) {
  labels <- attr(stats::terms(spec$formula), "term.labels")
  stopifnot(all(drop %in% labels), !any(add %in% labels))
  spec$formula <- stats::reformulate(
    c(setdiff(labels, drop), add),
    response = "complete",
    env = environment(spec$formula)
  )
  spec
}

# The specs before §9.13's revision: charted three-level pass_location in
# place of the adopted pass_middle and s(sideline_rec).
adopted <- c("pass_middle", "s(sideline_rec)")
s2 <- swap_terms(MODEL_SPECS[["2. play-by-play"]], adopted, "pass_location")
s4 <- swap_terms(MODEL_SPECS[["4. + geometry"]], adopted, "pass_location")

lat <- "s(lat_arr)"
side <- c(lat, "s(sideline_arr)")
side_in <- c(lat, "s(sideline_in)")
window <- "s(window_margin)"

candidates <- list(
  `2A` = swap_terms(s2, "pass_location", lat),
  `2B` = swap_terms(s2, "pass_location", side),
  `2C` = swap_terms(s2, "pass_location", side_in),
  `4B` = swap_terms(s4, "pass_location", side),
  `4C` = swap_terms(s4, "pass_location", side_in),
  `4B without window` = swap_terms(s4, c("pass_location", window), side),
  `4 without window` = swap_terms(s4, window)
)

tibble(
  model = names(candidates),
  formula = map_chr(candidates, \(s) {
    paste(deparse(s$formula, width.cutoff = 500L), collapse = " ")
  })
) |>
  gt() |>
  gt_theme_538(quiet = TRUE)
```

# 2. Out-of-fold comparison

```{r}
#| label: refit

# Folds from mf itself, which carries sideline_in.
folds <- make_folds(mf)

preds <- imap(
  c(list(`2 (charted)` = s2, `4 (charted)` = s4), candidates),
  \(spec, name) cv_predict(mf, folds, spec, name, verbose = FALSE)
) |>
  list_rbind()
```

```{r}
#| label: log-loss

preds |>
  group_by(model) |>
  summarize(n = n(), log_loss = log_loss(complete, .pred), .groups = "drop") |>
  arrange(desc(log_loss)) |>
  gt() |>
  tab_header(title = "Out-of-fold log loss") |>
  fmt_number(columns = log_loss, decimals = 4) |>
  fmt_number(columns = n, decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

Paired differences, model minus reference (negative favours the model):

```{r}
#| label: deltas

comparisons <- tribble(
  ~question, ~reference, ~model,
  "A: lateral offset for pass_location", "2 (charted)", "2A",
  "B: + sideline distance", "2 (charted)", "2B",
  "C: + sideline distance, clamped at 0", "2 (charted)", "2C",
  "B vs C: out-of-bounds distance", "2C", "2B",
  "sideline beyond lateral offset", "2A", "2B",
  "stage 4 with B vs stage 4", "4 (charted)", "4B",
  "tracking gain, current", "2 (charted)", "4 (charted)",
  "tracking gain on B", "2B", "4B",
  "tracking gain on C", "2C", "4C",
  "window margin, current stage 4", "4 (charted)", "4 without window",
  "window margin, stage 4 on B", "4B", "4B without window"
)

deltas <- comparisons |>
  pmap(\(question, reference, model) {
    paired_delta(preds, reference, model, "game_id") |>
      mutate(question = question)
  }) |>
  list_rbind()

deltas |>
  select(question, reference, model, estimate, conf_low, conf_high) |>
  gt() |>
  tab_header(
    title = "Paired differences in out-of-fold log loss",
    subtitle = "Model minus reference. Game-cluster bootstrap, 95% interval"
  ) |>
  fmt_number(columns = c(estimate, conf_low, conf_high), decimals = 4) |>
  gt_theme_538(quiet = TRUE)
```

# 3. Calibration

Point estimates only (`calibration_stats()`), overall and by depth; the
bootstrap intervals are in notebook 03 for the adopted stages.

```{r}
#| label: calibration

stage2_models <- c("2 (charted)", "2A", "2B", "2C")

by_depth <- preds |>
  filter(model %in% stage2_models) |>
  left_join(
    select(mf, game_id, play_id, depth_arr),
    by = c("game_id", "play_id")
  ) |>
  mutate(depth = as.character(depth_bucket(depth_arr)))

bind_rows(mutate(by_depth, depth = "all"), by_depth) |>
  group_by(model, depth) |>
  summarize(
    n = n(),
    stats = list(calibration_stats(complete, .pred)),
    .groups = "drop"
  ) |>
  unnest_wider(stats) |>
  select(depth, model, n, intercept, slope) |>
  arrange(depth, model) |>
  gt(groupname_col = "depth") |>
  fmt_number(columns = c(intercept, slope), decimals = 3) |>
  fmt_number(columns = n, decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

# 4. Candidate B, interpreted

A single fit on all rows; in-sample, for the shapes.

```{r}
#| label: fit-2b

f2b <- fit_full(mf, candidates[["2B"]])
st <- summary(f2b)$s.table

st[c("s(lat_arr)", "s(sideline_arr)"), , drop = FALSE] |>
  as.data.frame() |>
  rownames_to_column("smooth") |>
  rename(p = `p-value`) |>
  gt() |>
  fmt_number(columns = c(edf, Ref.df, Chi.sq), decimals = 2) |>
  fmt_scientific(columns = p, decimals = 2) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: smooths-2b
#| fig-height: 4

par(mfrow = c(1, 2))
for (sm in c("s(lat_arr)", "s(sideline_arr)")) {
  plot(
    f2b,
    select = match(sm, rownames(st)),
    shade = TRUE,
    seWithMean = TRUE,
    scale = 0,
    rug = TRUE
  )
  abline(h = 0, lty = 2, col = "grey50")
}
par(mfrow = c(1, 1))
```

Concurvity of the new smooths with the other stage 2 terms (`worst`):

```{r}
#| label: concurvity-2b

cc <- concurvity(f2b, full = FALSE)$worst
keep <- c("s(depth_arr)", "s(lat_arr)", "s(sideline_arr)")
cc_keep <- cc[keep, keep]
diag(cc_keep) <- NA

cc_keep |>
  as.data.frame() |>
  rownames_to_column("smooth") |>
  gt() |>
  fmt_number(columns = -smooth, decimals = 2) |>
  sub_missing(missing_text = "") |>
  gt_theme_538(quiet = TRUE)
```

# 5. Second round: binary middle and the receiver's sideline distance

Round one (§2–4, `notes/decisions.md` §9.13) found that the lateral offset adds
nothing and that the location effect is the sideline. Two revisions follow from
it:

- **`pass_location` becomes binary**, middle or not (`pass_middle`), the coding
  nflverse `cp` uses. There is no football argument for left and right to
  differ, the two levels are not distinguishable (§9.3), and the signed lateral
  offset fit as a flat line in candidate B.
- **Sideline distance is measured at the targeted receiver** at the arrival
  frame (`sideline_rec`) rather than at the ball. A ball sailing out of bounds
  is largely a missed throw, decided after the release; the receiver's
  position records where the catch would be made. The two measures correlate
  at 0.99 and differ almost only at the boundary. The receiver's tracked
  position is out of bounds on 109 plays (against 203 balls), partly because
  the sensor is on the shoulder pads, so the measure is also fit clamped at 0.

Expectations, recorded before fitting:

- **E** (`pass_middle` for `pass_location`): $|\bar d| < 0.0005$.
- **F** (E + `s(sideline_rec)`): close to B, a gain of 0.004 to 0.006.
- **G** (F clamped at 0): slightly below F, by less than B–C's 0.0009, since
  fewer receivers than balls are out of bounds.
- **Stage 4 on F**: a tracking gain near the 0.0552 found on B.

```{r}
#| label: candidates-2

side_rec <- c("pass_middle", "s(sideline_rec)")
side_rec_in <- c("pass_middle", "s(sideline_rec_in)")

candidates_2 <- list(
  `2E` = swap_terms(s2, "pass_location", "pass_middle"),
  `2F` = swap_terms(s2, "pass_location", side_rec),
  `2G` = swap_terms(s2, "pass_location", side_rec_in),
  `4F` = swap_terms(s4, "pass_location", side_rec)
)

preds_2 <- bind_rows(
  preds,
  imap(
    candidates_2,
    \(spec, name) cv_predict(mf, folds, spec, name, verbose = FALSE)
  ) |>
    list_rbind()
)

# Candidate F is the adopted spec, so it must reproduce the persisted
# predictions; the term order differs, hence the tolerance.
stopifnot(
  isTRUE(all.equal(
    preds_2$.pred[preds_2$model == "2F"],
    oof$.pred[oof$model == "2. play-by-play"],
    tolerance = 1e-6
  )),
  isTRUE(all.equal(
    preds_2$.pred[preds_2$model == "4F"],
    oof$.pred[oof$model == "4. + geometry"],
    tolerance = 1e-6
  ))
)
```

```{r}
#| label: deltas-2

comparisons_2 <- tribble(
  ~question, ~reference, ~model,
  "E: pass_middle for pass_location", "2 (charted)", "2E",
  "F: + receiver sideline distance", "2 (charted)", "2F",
  "G: + receiver sideline distance, clamped at 0", "2 (charted)", "2G",
  "F vs G: receiver out of bounds", "2G", "2F",
  "F vs C: receiver vs ball (clamped)", "2C", "2F",
  "stage 4 with F vs stage 4", "4 (charted)", "4F",
  "tracking gain on F", "2F", "4F"
)

comparisons_2 |>
  pmap(\(question, reference, model) {
    paired_delta(preds_2, reference, model, "game_id") |>
      mutate(question = question)
  }) |>
  list_rbind() |>
  select(question, reference, model, estimate, conf_low, conf_high) |>
  gt() |>
  tab_header(
    title = "Paired differences in out-of-fold log loss, round two",
    subtitle = "Model minus reference. Game-cluster bootstrap, 95% interval"
  ) |>
  fmt_number(columns = c(estimate, conf_low, conf_high), decimals = 4) |>
  gt_theme_538(quiet = TRUE)
```

```{r}
#| label: calibration-2

by_depth_2 <- preds_2 |>
  filter(model %in% c("2 (charted)", "2C", "2E", "2F", "2G")) |>
  left_join(
    select(mf, game_id, play_id, depth_arr),
    by = c("game_id", "play_id")
  ) |>
  mutate(depth = as.character(depth_bucket(depth_arr)))

bind_rows(mutate(by_depth_2, depth = "all"), by_depth_2) |>
  group_by(model, depth) |>
  summarize(
    n = n(),
    log_loss = log_loss(complete, .pred),
    stats = list(calibration_stats(complete, .pred)),
    .groups = "drop"
  ) |>
  unnest_wider(stats) |>
  select(depth, model, n, log_loss, intercept, slope) |>
  arrange(depth, model) |>
  gt(groupname_col = "depth") |>
  fmt_number(columns = c(log_loss, intercept, slope), decimals = 4) |>
  fmt_number(columns = n, decimals = 0) |>
  gt_theme_538(quiet = TRUE)
```

Candidate F on all rows, in-sample, for the shapes:

```{r}
#| label: fit-2f
#| fig-height: 4

f2f <- fit_full(mf, candidates_2[["2F"]])

summary(f2f)$p.table["pass_middle", , drop = FALSE] |>
  as.data.frame() |>
  rownames_to_column("term") |>
  gt() |>
  fmt_number(columns = -term, decimals = 3) |>
  gt_theme_538(quiet = TRUE)

st_f <- summary(f2f)$s.table
plot(
  f2f,
  select = match("s(sideline_rec)", rownames(st_f)),
  shade = TRUE,
  seWithMean = TRUE,
  scale = 0,
  rug = TRUE
)
abline(h = 0, lty = 2, col = "grey50")
```
